---
title: "Explore And Visualize"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

The worked example today showed how to:

- load a course dataset
- inspect what one observation represents
- create a small table before plotting
- filter to the level that matches the question
- build a plot step by step with `ggplot2`

# Practice Task

Pick one course dataset and create:

1. a structure table that describes the dataset
2. one count or summary table
3. one visualization
4. a two- or three-sentence interpretation

Helpful module reference: `modules/01-workflow-and-basics/02_first-visualization.qmd`.

Course data summary and codebooks: [rendered Course Data page](../../data/data.html).

# Load And Inspect

```{r}
# dataset_path <- "data/real/prams_2011_selected.csv"
# dataset_path <- "data/real/census_state_race_ethnicity_long.csv"
# dataset_path <- "data/real/flu_vaccination_age_time_clean.csv"
# dataset_path <- "data/real/nhanes_2021_2023_analysis.csv"
# dataset_path <- "data/real/places_county_core_measures_long.csv"

dataset_path <- "data/real/prams_2011_selected.csv"

dat <- read_csv(dataset_path, show_col_types = FALSE)

dat |>
  glimpse()
```

# Prepare Data

Create the structure table, count table, summary table, or filtered dataset needed for the plot.

```{r}

```

# Visualization

Build one visualization from the prepared data.

```{r}

```

# Interpretation

Write briefly:

- what one observation represents
- what table or plot was created
- what the plot shows
- what caution belongs with the interpretation
