---
title: "Explore And Visualize"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# Goal

Choose one course dataset, inspect it, tabulate a few variables, and create one visualization.

This practice is intentionally open-ended. The goal is to get comfortable opening a dataset, asking what is in it, and making a first plot.

# Choose A Dataset

Uncomment one dataset path below, or replace it with another path listed in the Course Data Summary or `data/manifest.csv`.

```{r}
# dataset_path <- "data/real/prams_2011_selected.csv"
# dataset_path <- "data/real/census_state_race_ethnicity_long.csv"
# dataset_path <- "data/real/flu_vaccination_age_time_clean.csv"
# dataset_path <- "data/real/nhanes_2021_2023_analysis.csv"
# dataset_path <- "data/real/places_county_core_measures_long.csv"

dataset_path <- "data/real/prams_2011_selected.csv"
```

# Load And Inspect

```{r}
dat <- read_csv(dataset_path, show_col_types = FALSE)

dat |>
  glimpse()
```

# Tabulate A Few Things

Pick one or two categorical variables and count the rows in each category.

Use the `glimpse()` output above to choose variables that look categorical.

```{r}
# Count one categorical variable here.
```

```{r}
# Count another categorical variable here.
```

Write one sentence about what you notice in the counts.

# Make One Visualization

Start with one simple question.

Examples:

- How are rows distributed across one categorical variable?
- What is the shape of one numeric variable?
- How are two numeric variables related?

Prepare the data you want to plot.

```{r}
# Prepare your plot data here, if needed.
```

Start a plot object.

```{r}
# p1 <- ggplot(...)

# p1
```

Add a geom.

```{r}
# p2 <- p1 +
#   geom_...()

# p2
```

Add labels and a theme.

```{r}
# p3 <- p2 +
#   labs(
#     title = "...",
#     x = "...",
#     y = "..."
#   ) +
#   theme_minimal()

# p3
```

Write two or three sentences interpreting your plot.

# Your Turn

If you have time, revise your plot or try a second plot.

Try one of these:

- choose a different dataset path
- count a different categorical variable
- make a histogram of a numeric variable
- make a scatterplot with two numeric variables
- improve the title and axis labels

```{r}
# Your code here
```

# Render Check

Render this notebook. Confirm that the rendered HTML includes your table output and your visualization.
