---
title: "Practice: Open And Render Check"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
```

# Goal

Use this notebook to confirm that your course workflow is working.

You should be able to run the code chunks, render the notebook, and find the rendered HTML file.

# Step 1: Check The Project

Run this chunk.

```{r}
getwd()
```

Question: does the output look like the course project folder?

Write a short note here:

> TODO: Replace this with your answer.

# Step 2: Check R Code

Run this simple calculation.

```{r}
2 + 2
```

Now change the calculation below to something else.

```{r}
10 * 5
```

# Step 3: Check Data

We will use `mtcars`, a small built-in dataset.

```{r}
mtcars |>
  head()
```

Run a simple summary.

```{r}
mtcars |>
  summarise(
    rows = n(),
    average_mpg = mean(mpg),
    minimum_mpg = min(mpg),
    maximum_mpg = max(mpg)
  )
```

Question: how many rows are in the dataset?

> TODO: Replace this with your answer.

# Step 4: Check Plotting

Run this plot.

```{r}
ggplot(mtcars, aes(x = mpg)) +
  geom_histogram(binwidth = 5) +
  labs(
    title = "Distribution of Fuel Efficiency",
    x = "Miles per gallon",
    y = "Number of cars"
  )
```

Question: what does the plot show?

> TODO: Replace this with one or two sentences.

# Step 5: Render

Render this notebook.

After rendering, answer these questions:

- Did the notebook render successfully?
- Where did the HTML file appear?
- Did the plot appear in the rendered HTML file?

Notes:

> TODO: Replace this with your notes.
