---
title: "Practice: First Visualization"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# Goal

Create simple visualizations by building `ggplot2` objects piece by piece.

For each plot:

1. create `p1`
2. render `p1`
3. create `p2` by adding one thing
4. render `p2`
5. keep going

# Load Data

```{r}
prams_path <- "data/real/prams_2011_selected.csv"

prams <- read_csv(prams_path, show_col_types = FALSE)

prams |>
  glimpse()
```

What does one row represent?

> TODO: Replace this with your answer.

# Plot 1: Distribution

Question: what is the distribution of reported binge drinking before pregnancy?

```{r}
prams_binge <- prams |>
  filter(!is.na(binge_drinking_within_3_months_birth))
```

Start with data and mapping.

```{r}
p1 <- ggplot(
  prams_binge,
  aes(x = binge_drinking_within_3_months_birth)
)

p1
```

Add the histogram geom.

```{r}
p2 <- p1 +
  geom_histogram(binwidth = 5)

p2
```

Add labels.

```{r}
p3 <- p2 +
  labs(
    title = "Distribution of Reported Binge Drinking Before Pregnancy",
    x = "Percent reporting binge drinking",
    y = "Number of state/subgroup rows"
  )

p3
```

Your turn: create `p4` by adding `theme_minimal()`.

```{r}
p4 <- p3 +
  theme_minimal()

p4
```

Interpretation:

> TODO: Write one or two sentences describing the center, spread, shape, or unusual values.

# Plot 2: Counts By Category

Question: which subgroup categories appear most often?

```{r}
subgroup_counts <- prams |>
  count(subgroup_cat, sort = TRUE) |>
  mutate(subgroup_cat = reorder(subgroup_cat, n))

subgroup_counts
```

Start with data and mappings.

```{r}
p1 <- ggplot(subgroup_counts, aes(x = subgroup_cat, y = n))

p1
```

Add columns.

```{r}
p2 <- p1 +
  geom_col()

p2
```

Add coordinate flip.

```{r}
p3 <- p2 +
  coord_flip()

p3
```

Your turn: create `p4` by adding labels and a theme.

```{r}
p4 <- p3 +
  labs(
    title = "PRAMS Rows By Subgroup Category",
    x = NULL,
    y = "Number of rows"
  ) +
  theme_minimal()

p4
```

Interpretation:

> TODO: Write one or two sentences about the most common and least common subgroup categories.

# Plot 3: Association

Question: how are reported alcohol use and binge drinking related?

```{r}
prams_alcohol <- prams |>
  filter(
    !is.na(alcohol_use_within_3_months_birth),
    !is.na(binge_drinking_within_3_months_birth)
  )
```

Start with data and mappings.

```{r}
p1 <- ggplot(
  prams_alcohol,
  aes(
    x = alcohol_use_within_3_months_birth,
    y = binge_drinking_within_3_months_birth
  )
)

p1
```

Add points.

```{r}
p2 <- p1 +
  geom_point(alpha = 0.6)

p2
```

Add a trend line.

```{r}
p3 <- p2 +
  geom_smooth(method = "lm", se = FALSE)

p3
```

Your turn: create `p4` by adding labels and a theme.

```{r}
p4 <- p3 +
  labs(
    title = "Alcohol Use And Binge Drinking Before Pregnancy",
    x = "Percent reporting any alcohol use",
    y = "Percent reporting binge drinking"
  ) +
  theme_minimal()

p4
```

Interpretation:

> TODO: Write one or two sentences describing the association. Do not overclaim causality.

# Reflection

What did the object-building workflow help you see?

> TODO: Replace this with your answer.

# Render

Render this notebook and make sure each intermediate plot appears in the HTML file.
