---
title: "Practice: Grammar Of Graphics With ggplot2"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# Goal

Practice the grammar of graphics by building `ggplot2` objects step by step.

For each plot object, ask what changed from the previous object.

# Load Data

```{r}
prams_path <- "data/real/prams_2011_selected.csv"

prams <- read_csv(prams_path, show_col_types = FALSE)

prams |>
  glimpse()
```

What does one row represent?

> TODO: Replace this with your answer.

# Part 1: Data

```{r}
prams_binge <- prams |>
  filter(!is.na(binge_drinking_within_3_months_birth))
```

Start with data only.

```{r}
p1 <- ggplot(data = prams_binge)

p1
```

What does this show, and what is missing?

> TODO: Replace this with your answer.

# Part 2: Aesthetic Mapping

Add an aesthetic mapping.

```{r}
p2 <- ggplot(
  data = prams_binge,
  mapping = aes(x = binge_drinking_within_3_months_birth)
)

p2
```

What variable is mapped to the x-axis?

> TODO: Replace this with your answer.

# Part 3: Geom

Add a geom.

```{r}
p3 <- p2 +
  geom_histogram(binwidth = 5)

p3
```

What geom is used?

> TODO: Replace this with your answer.

# Part 4: Labels And Theme

Add labels.

```{r}
p4 <- p3 +
  labs(
    title = "Binge Drinking Before Pregnancy",
    x = "Percent reporting binge drinking",
    y = "Number of state/subgroup rows"
  )

p4
```

Add a theme.

```{r}
p5 <- p4 +
  theme_minimal()

p5
```

# Part 5: Two-Variable Mapping

```{r}
prams_alcohol <- prams |>
  filter(
    !is.na(alcohol_use_within_3_months_birth),
    !is.na(binge_drinking_within_3_months_birth)
  )
```

Start a scatterplot.

```{r}
p1 <- ggplot(
  prams_alcohol,
  aes(
    x = alcohol_use_within_3_months_birth,
    y = binge_drinking_within_3_months_birth
  )
)

p1
```

Add points.

```{r}
p2 <- p1 +
  geom_point(alpha = 0.6)

p2
```

Add a smoother.

```{r}
p3 <- p2 +
  geom_smooth(method = "lm", se = FALSE)

p3
```

What did the smoother add?

> TODO: Replace this with your answer.

# Part 6: Statistical Transformation Inside ggplot2

`geom_bar()` counts rows for you.

```{r}
p1 <- ggplot(prams, aes(x = subgroup_cat))

p1
```

```{r}
p2 <- p1 +
  geom_bar()

p2
```

What is `geom_bar()` counting?

> TODO: Replace this with your answer.

# Part 7: Aggregate First, Then Plot

Calculate counts yourself.

```{r}
subgroup_counts <- prams |>
  count(subgroup_cat, sort = TRUE) |>
  mutate(subgroup_cat = reorder(subgroup_cat, n))

subgroup_counts
```

Build the plot piece by piece.

```{r}
p1 <- ggplot(subgroup_counts, aes(x = subgroup_cat, y = n))

p1
```

```{r}
p2 <- p1 +
  geom_col()

p2
```

```{r}
p3 <- p2 +
  coord_flip()

p3
```

Why do we use `geom_col()` here instead of `geom_bar()`?

> TODO: Replace this with your answer.

# Part 8: Transform First, Then Plot

Create a new variable.

```{r}
prams_transformed <- prams_alcohol |>
  mutate(
    alcohol_minus_binge = alcohol_use_within_3_months_birth -
      binge_drinking_within_3_months_birth
  )

prams_transformed |>
  select(alcohol_use_within_3_months_birth, binge_drinking_within_3_months_birth, alcohol_minus_binge) |>
  head()
```

Plot the transformed variable.

```{r}
p1 <- ggplot(prams_transformed, aes(x = alcohol_minus_binge))

p1
```

```{r}
p2 <- p1 +
  geom_histogram(binwidth = 5)

p2
```

Why might we create a new variable before plotting?

> TODO: Replace this with your answer.

# Part 9: Facets

Create a base plot for selected subgroup categories.

```{r}
p1 <- prams_alcohol |>
  filter(subgroup_cat %in% c("None", "Maternal Age (3 Levels)", "Maternal Education")) |>
  ggplot(
    aes(
      x = alcohol_use_within_3_months_birth,
      y = binge_drinking_within_3_months_birth
    )
  ) +
  geom_point(alpha = 0.7)

p1
```

Add facets.

```{r}
p2 <- p1 +
  facet_wrap(~ subgroup_cat)

p2
```

What comparison does faceting make easier?

> TODO: Replace this with your answer.

# Grammar Checklist

Choose one plot from this notebook and fill in the grammar.

- data:
- row meaning:
- x mapping:
- y mapping:
- other aesthetic mapping:
- geom:
- statistic or aggregation:
- scale or coordinate choice:
- facet:
- labels/theme:

# Final Reflection

What is one thing the step-by-step plot object workflow helped you understand?

> TODO: Replace this with your reflection.
