---
title: "Practice: Submission Check"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
```

# Goal

Use this notebook as a checklist before submitting or sharing a rendered Quarto file.

# Render Check

Before submitting, confirm:

- The notebook renders without errors.
- The rendered HTML file opens.
- All plots appear in the rendered HTML file.
- Your written interpretations appear in the rendered HTML file.
- The file name is clear enough for someone else to understand.

# File Check

Write the name of the notebook you rendered:

> TODO: Replace this with the `.qmd` file name.

Write the name of the rendered HTML file:

> TODO: Replace this with the `.html` file name.

Where did the HTML file appear on your computer?

> TODO: Replace this with the folder path or description.

# Content Check

How many plots are in your rendered notebook?

> TODO: Replace this with your answer.

Which plot do you think is most successful, and why?

> TODO: Replace this with two or three sentences.

What is one thing you would improve if you had more time?

> TODO: Replace this with one or two sentences.

# Tiny Code Check

This small chunk exists so we can confirm this checklist notebook also renders correctly.

```{r}
today <- Sys.Date()
today
```

# Final Note

For this course, the rendered HTML notebook is the main artifact. It should show both your code and your thinking.
