---
title: "Practice: Counts And Ordering"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# What We Practiced

The module showed how to:

- count observations by category with `count()`
- aggregate numeric values by category with `group_by()` and `summarize()`
- order categories before plotting
- use `geom_col()` when the bar height is already in the data

# Task

Pick one dataset and create:

1. a categorical count or grouped summary table
2. an ordered categorical plot
3. a short interpretation explaining what was counted or summarized

Suggested datasets:

- `data/real/census_state_race_ethnicity_long.csv`
- `data/real/state_population_ranks.csv`
- `data/simulated/state_school_enrollment_groups_long.csv`

Helpful module reference: `modules/02_categorical-data/01_counts-and-ordering.qmd`.

# Work

```{r}
# Load data, create the count or summary table, and build the ordered plot.
```

# Interpretation

Write briefly: what was counted or summarized, what one plotted observation represents, and why the ordering was chosen.
