---
title: "Practice: Counts And Ordering"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# Goal

Practice making a categorical count or summary plot and ordering the categories deliberately.

# Suggested Data

Choose one:

- `data/real/census_state_race_ethnicity_long.csv`
- `data/real/state_population_ranks.csv`
- `data/simulated/state_school_enrollment_groups_long.csv`

Helpful module reference: `modules/02_categorical-data/01_counts-and-ordering.qmd`.

# Load And Inspect

Load one dataset and inspect it with `glimpse()`.

```{r}
# Load your dataset here.
```

```{r}
# Use glimpse() here.
```

# Count Or Summarize

Pick one categorical variable. Decide whether you are counting rows or summarizing a numeric variable by category.

```{r}
# Create a table with one row per category.
```

# Plot

Build the plot step by step. Try `geom_col()` after you have summarized the data yourself.

```{r}
# Start p1 with ggplot().
```

```{r}
# Add a geom.
```

```{r}
# Reorder categories if useful, then add labels and a theme.
```

# Interpret

Write two or three sentences.

> TODO: What did you count or summarize? Which categories are largest or smallest? Why did you choose this ordering?
