---
title: "Practice: Counts And Ordering"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# Goal

Practice making categorical count plots and ordering categories deliberately.

# Load Data

```{r}
census_path <- "data/real/census_state_race_ethnicity_long.csv"

census <- read_csv(census_path, show_col_types = FALSE)

census |>
  glimpse()
```

What does one row represent?

> TODO: Replace this with your answer.

# Part 1: Count Rows

```{r}
category_counts <- census |>
  count(race_ethnicity, sort = TRUE)

category_counts
```

```{r}
p1 <- ggplot(category_counts, aes(x = race_ethnicity, y = n))

p1
```

```{r}
p2 <- p1 +
  geom_col()

p2
```

What is this plot counting?

> TODO: Replace this with your answer.

# Part 2: Summarize Population

Calculate total included population by race/ethnicity category.

```{r}
category_population <- census |>
  group_by(race_ethnicity) |>
  summarise(
    total_population = sum(race_ethnicity_population),
    .groups = "drop"
  ) |>
  arrange(desc(total_population))

category_population
```

Build a plot.

```{r}
p1 <- ggplot(category_population, aes(x = race_ethnicity, y = total_population))

p1
```

```{r}
p2 <- p1 +
  geom_col()

p2
```

# Part 3: Reorder And Flip

```{r}
category_population_ordered <- category_population |>
  mutate(race_ethnicity = reorder(race_ethnicity, total_population))

p3 <- ggplot(
  category_population_ordered,
  aes(x = race_ethnicity, y = total_population)
) +
  geom_col()

p3
```

```{r}
p4 <- p3 +
  coord_flip()

p4
```

# Part 4: Labels And Theme

```{r}
p5 <- p4 +
  labs(
    title = "Included Population By Race/Ethnicity Category",
    x = NULL,
    y = "Population count"
  ) +
  theme_minimal()

p5
```

Write a two-sentence interpretation.

> TODO: Replace this with your interpretation.

# Part 5: Try A Dot Plot

```{r}
p1 <- ggplot(
  category_population_ordered,
  aes(x = total_population, y = race_ethnicity)
)

p1
```

```{r}
p2 <- p1 +
  geom_point(size = 3)

p2
```

Which version do you prefer for this question: columns or dots? Why?

> TODO: Replace this with your answer.

# Checklist

- What is the unit of observation?
- What did you count or summarize?
- Are categories ordered for the question?
- Does the plot show row counts, population counts, or proportions?
