---
title: "Practice: Composition And Proportions"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# Goal

Practice visualizing composition with proportions.

# Load Data

```{r}
census_path <- "data/real/census_state_race_ethnicity_long.csv"

census <- read_csv(census_path, show_col_types = FALSE)

census |>
  glimpse()
```

# Part 1: Select States

Start with these states. You may change them later.

```{r}
selected_states <- census |>
  filter(state %in% c("CA", "FL", "MD", "MA", "TX", "WA"))

selected_states |>
  select(state_name, race_ethnicity, race_ethnicity_population, share) |>
  head()
```

Why is it useful to keep only a few states for this plot?

> TODO: Replace this with your answer.

# Part 2: Raw Counts

```{r}
p1 <- ggplot(
  selected_states,
  aes(
    x = state_name,
    y = race_ethnicity_population,
    fill = race_ethnicity
  )
)

p1
```

```{r}
p2 <- p1 +
  geom_col()

p2
```

What question does this count plot answer?

> TODO: Replace this with your answer.

# Part 3: Proportions

```{r}
p1 <- ggplot(
  selected_states,
  aes(
    x = state_name,
    y = share,
    fill = race_ethnicity
  )
)

p1
```

```{r}
p2 <- p1 +
  geom_col()

p2
```

```{r}
p3 <- p2 +
  scale_y_continuous(
    labels = function(x) paste0(round(x * 100), "%")
  ) +
  labs(
    title = "Race/Ethnicity Composition In Selected States",
    x = NULL,
    y = "Share of included race/ethnicity population",
    fill = "Category"
  ) +
  theme_minimal()

p3
```

What question does this proportion plot answer?

> TODO: Replace this with your answer.

# Part 4: One Category Across States

Choose one race/ethnicity category and compare its share across states.

```{r}
one_category <- census |>
  filter(race_ethnicity == "Hispanic or Latino") |>
  mutate(state_name = reorder(state_name, share))

one_category |>
  select(state_name, share) |>
  head()
```

```{r}
p1 <- ggplot(one_category, aes(x = state_name, y = share))

p1
```

```{r}
p2 <- p1 +
  geom_col(fill = "steelblue")

p2
```

```{r}
p3 <- p2 +
  coord_flip() +
  scale_y_continuous(
    labels = function(x) paste0(round(x * 100), "%")
  ) +
  labs(
    title = "Hispanic Or Latino Share By State",
    x = NULL,
    y = "Share"
  ) +
  theme_minimal()

p3
```

Write a short interpretation. Be clear about the denominator.

> TODO: Replace this with your interpretation.

# Extension

Change `"Hispanic or Latino"` to another category and rerun the notebook.

Which category did you choose, and what changed?

> TODO: Replace this with your answer.
