---
title: "Practice: Composition And Proportions"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# Goal

Practice showing how parts relate to a whole.

# Suggested Data

Choose one:

- `data/real/census_state_race_ethnicity_long.csv`
- `data/real/state_race_population_summary.csv`
- `data/simulated/state_school_enrollment_groups_long.csv`

Helpful module reference: `modules/02_categorical-data/02_composition-and-proportions.qmd`.

# Load And Inspect

```{r}
# Load your dataset here.
```

```{r}
# Inspect the rows and variables.
```

# Choose A Whole

Decide what the whole is. Examples: one state, one year, one county, or one subgroup.

```{r}
# Filter or summarize the data so the parts add up within a clear whole.
```

# Calculate Proportions

Create a proportion or percentage variable if the dataset does not already have one.

```{r}
# Calculate the part-to-whole measure here.
```

# Plot

Build a composition plot. Consider columns, stacked columns, or a dot plot of proportions.

```{r}
# Build the plot here.
```

# Interpret

> TODO: What is the whole? What are the parts? What should readers be careful not to over-interpret?
