---
title: "Practice: Composition And Proportions"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# What We Practiced

The module showed how to:

- define the whole before plotting parts
- use counts and proportions for different questions
- calculate or use a share variable
- simplify a stacked comparison by filtering to one category

# Task

Pick one dataset and create:

1. a table that defines the whole and its parts
2. one composition or proportion plot
3. a short interpretation naming the denominator

Suggested datasets:

- `data/real/census_state_race_ethnicity_long.csv`
- `data/real/state_race_population_summary.csv`
- `data/simulated/state_school_enrollment_groups_long.csv`

Helpful module reference: `modules/02_categorical-data/02_composition-and-proportions.qmd`.

# Work

```{r}
# Load data, define the whole, calculate or use proportions, and build the plot.
```

# Interpretation

Write briefly: what the whole is, what the parts are, and what readers should not over-interpret.
