---
title: "Practice: Ranked Comparisons"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# Goal

Practice making ranked categorical comparisons.

# Load Data

```{r}
population_path <- "data/real/state_population_ranks.csv"

state_population <- read_csv(population_path, show_col_types = FALSE)

state_population |>
  glimpse()
```

What does one row represent?

> TODO: Replace this with your answer.

# Part 1: First Plot

```{r}
p1 <- ggplot(state_population, aes(x = state_name, y = population))

p1
```

```{r}
p2 <- p1 +
  geom_col()

p2
```

What makes this first plot hard to read?

> TODO: Replace this with your answer.

# Part 2: Top 15 States

```{r}
top_states <- state_population |>
  arrange(desc(population)) |>
  slice_head(n = 15) |>
  mutate(
    state_name = reorder(state_name, population),
    population_millions = population / 1000000
  )

top_states
```

What did `slice_head(n = 15)` do here?

> TODO: Replace this with your answer.

# Part 3: Ranked Bar Chart

```{r}
p1 <- ggplot(top_states, aes(x = state_name, y = population_millions))

p1
```

```{r}
p2 <- p1 +
  geom_col(fill = "gray60")

p2
```

```{r}
p3 <- p2 +
  coord_flip()

p3
```

```{r}
p4 <- p3 +
  labs(
    title = "Largest States By 2020 Population",
    subtitle = "Top 15 states and DC in the course population file",
    x = NULL,
    y = "Population, in millions"
  ) +
  theme_minimal()

p4
```

Write a short interpretation.

> TODO: Replace this with your interpretation.

# Part 4: Dot Plot

```{r}
p1 <- ggplot(top_states, aes(x = population_millions, y = state_name))

p1
```

```{r}
p2 <- p1 +
  geom_point(size = 3)

p2
```

```{r}
p3 <- p2 +
  labs(
    title = "Largest States By 2020 Population",
    x = "Population, in millions",
    y = NULL
  ) +
  theme_minimal()

p3
```

Which version do you prefer: bars or dots? Why?

> TODO: Replace this with your answer.

# Part 5: Highlight One State

Choose a state to highlight. The example highlights California.

```{r}
top_states <- top_states |>
  mutate(is_selected_state = state == "CA")

p1 <- ggplot(
  top_states,
  aes(
    x = population_millions,
    y = state_name,
    color = is_selected_state
  )
) +
  geom_point(size = 3)

p1
```

```{r}
p2 <- p1 +
  scale_color_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "gray65"),
    labels = c("FALSE" = "Other top-15 state", "TRUE" = "Selected state"),
    name = NULL
  ) +
  labs(
    title = "Selected State Compared With Other Large States",
    x = "Population, in millions",
    y = NULL
  ) +
  theme_minimal()

p2
```

What does highlighting add? What could it overemphasize?

> TODO: Replace this with your answer.

# Checklist

- Did you sort by the variable that matches the question?
- Did you say if the plot shows only a subset?
- Are labels readable?
- Does highlighting help the reader, or distract them?
