---
title: "Practice: Ranked Comparisons"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# Goal

Practice using rank to compare places or groups.

# Suggested Data

Choose one:

- `data/real/state_population_ranks.csv`
- `data/real/places_county_core_measures_long.csv`
- `data/simulated/state_enrollment_ranks.csv`

Helpful module reference: `modules/02_categorical-data/03_ranked-comparisons.qmd`.

# Load And Inspect

```{r}
# Load your dataset here.
```

```{r}
# Inspect the dataset here.
```

# Choose A Ranking Variable

Pick the numeric variable you want to rank by. If the data already include a rank, decide whether to use it or calculate your own.

```{r}
# Create a ranked table here. Consider showing only the top or bottom 10.
```

# Plot

Build a ranked plot. A horizontal dot plot or column plot often works well.

```{r}
# Build your ranked comparison plot here.
```

# Interpret

> TODO: What is being ranked? What is the comparison group? Does rank hide any important magnitude differences?
