---
title: "Practice: Ranked Comparisons"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# What We Practiced

The module showed how to:

- sort observations by a numeric variable
- limit a plot to a readable top or bottom group
- show rank without hiding magnitude
- highlight one category while preserving context

# Task

Pick one dataset and create:

1. a ranked table
2. one ranked comparison plot
3. a short interpretation that names what was included and excluded

Suggested datasets:

- `data/real/state_population_ranks.csv`
- `data/real/places_county_core_measures_long.csv`
- `data/simulated/state_enrollment_ranks.csv`

Helpful module reference: `modules/02_categorical-data/03_ranked-comparisons.qmd`.

# Work

```{r}
# Load data, create a ranked table, and build a ranked comparison plot.
```

# Interpretation

Write briefly: what is ranked, what comparison group is shown, and whether rank hides any important magnitude differences.
