---
title: "Practice: Histograms And Density Plots"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# Goal

Practice describing the distribution of one numeric variable.

# Suggested Data

Choose one:

- `data/real/nhanes_2021_2023_analysis.csv`
- `data/real/davis_height_weight.csv`
- `data/simulated/neighborhood_health_survey_analysis.csv`

Helpful module reference: `modules/03_continuous-data/01_histograms-and-density.qmd`.

# Load And Inspect

```{r}
# Load your dataset here.
```

```{r}
# Use glimpse() to find numeric variables.
```

# Choose A Numeric Variable

Pick one numeric variable. Filter out missing values if needed.

```{r}
# Prepare the variable you want to plot.
```

# Plot

Make a histogram. Try at least two bin widths.

```{r}
# Histogram with your first bin width.
```

```{r}
# Histogram with another bin width or a density plot.
```

# Interpret

> TODO: Describe center, spread, shape, and unusual values. How did bin width affect what you saw?
