---
title: "Practice: Histograms And Density Plots"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# What We Practiced

The module showed how to:

- filter to the observations needed for a numeric question
- summarize center and spread with `mean()`, `median()`, `IQR()`, and `sd()`
- use histograms and density plots to show shape
- compare bin-width choices

# Task

Pick one dataset and create:

1. a summary table for one numeric variable
2. two distribution plots using different bin widths or plot types
3. a short interpretation comparing the table and plots

Suggested datasets:

- `data/real/nhanes_2021_2023_analysis.csv`
- `data/real/davis_height_weight.csv`
- `data/simulated/neighborhood_health_survey_analysis.csv`

Helpful module reference: `modules/03_continuous-data/01_histograms-and-density.qmd`.

# Work

```{r}
# Load data, summarize one numeric variable, and build two distribution plots.
```

# Interpretation

Write briefly: what the summary table showed, what the plots added, and how bin width affected the pattern.
