---
title: "Practice: Histograms And Density Plots"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# Goal

Practice describing one continuous distribution.

# Load Data

```{r}
nhanes_path <- "data/real/nhanes_2021_2023_analysis.csv"

nhanes <- read_csv(nhanes_path, show_col_types = FALSE)

nhanes |>
  glimpse()
```

What does one row represent?

> TODO: Replace this with your answer.

# Part 1: Prepare Data

```{r}
adult_bp <- nhanes |>
  filter(
    age_years >= 18,
    !is.na(systolic_mean)
  )

adult_bp |>
  select(respondent_id, age_years, gender, age_group, systolic_mean) |>
  head()
```

# Part 2: Histogram

```{r}
p1 <- ggplot(adult_bp, aes(x = systolic_mean))

p1
```

```{r}
p2 <- p1 +
  geom_histogram(binwidth = 5)

p2
```

```{r}
p3 <- p2 +
  labs(
    title = "Distribution Of Adult Mean Systolic Blood Pressure",
    x = "Mean systolic blood pressure",
    y = "Number of participants"
  ) +
  theme_minimal()

p3
```

Describe the center, spread, shape, and any unusual values.

> TODO: Replace this with your answer.

# Part 3: Change Bin Width

```{r}
p1 <- ggplot(adult_bp, aes(x = systolic_mean)) +
  geom_histogram(binwidth = 2)

p1
```

```{r}
p2 <- ggplot(adult_bp, aes(x = systolic_mean)) +
  geom_histogram(binwidth = 10)

p2
```

What changes when the bin width changes?

> TODO: Replace this with your answer.

# Part 4: Density Plot

```{r}
p1 <- ggplot(adult_bp, aes(x = systolic_mean))

p1
```

```{r}
p2 <- p1 +
  geom_density(fill = "steelblue", alpha = 0.3)

p2
```

What does the density plot make easier or harder to see than the histogram?

> TODO: Replace this with your answer.

# Part 5: Summaries

```{r}
adult_bp |>
  summarise(
    n = n(),
    mean_systolic = mean(systolic_mean),
    median_systolic = median(systolic_mean),
    min_systolic = min(systolic_mean),
    max_systolic = max(systolic_mean),
    iqr_systolic = IQR(systolic_mean)
  )
```

How do the summaries support or complicate what you saw in the plot?

> TODO: Replace this with your answer.
