---
title: "Practice: Grouped Distributions"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# What We Practiced

The module showed how to:

- check group sizes before comparing distributions
- compare center, spread, shape, and unusual values across groups
- use box plots, violin plots, or faceted distribution plots
- pair plots with grouped summary tables

# Task

Pick one dataset and create:

1. a group-size table
2. a grouped summary table
3. one grouped distribution plot
4. a short interpretation comparing summaries and the plot

Suggested datasets:

- `data/real/nhanes_2021_2023_analysis.csv`
- `data/real/exercise_eating_disorders.csv`
- `data/simulated/physical_therapy_activity.csv`

Helpful module reference: `modules/03_continuous-data/02_grouped-distributions.qmd`.

# Work

```{r}
# Load data, create group-size and summary tables, and build one grouped distribution plot.
```

# Interpretation

Write briefly: which groups differ in center, spread, shape, or unusual values, and how the summary table compares with the plot.
