---
title: "Practice: Grouped Distributions"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# Goal

Practice comparing continuous distributions across groups.

# Load Data

```{r}
nhanes_path <- "data/real/nhanes_2021_2023_analysis.csv"

nhanes <- read_csv(nhanes_path, show_col_types = FALSE)

nhanes |>
  glimpse()
```

# Part 1: Prepare Data

```{r}
adult_bp <- nhanes |>
  filter(
    age_years >= 18,
    !is.na(systolic_mean),
    !is.na(age_group)
  )

adult_bp |>
  count(age_group)
```

Why is it useful to count rows by group before comparing distributions?

> TODO: Replace this with your answer.

# Part 2: Box Plot

```{r}
p1 <- ggplot(adult_bp, aes(x = age_group, y = systolic_mean))

p1
```

```{r}
p2 <- p1 +
  geom_boxplot()

p2
```

```{r}
p3 <- p2 +
  labs(
    title = "Adult Systolic Blood Pressure By Age Group",
    x = "Age group",
    y = "Mean systolic blood pressure"
  ) +
  theme_minimal()

p3
```

What differs across age groups: center, spread, shape, outliers, or overlap?

> TODO: Replace this with your answer.

# Part 3: Add Points

```{r}
p1 <- ggplot(adult_bp, aes(x = age_group, y = systolic_mean)) +
  geom_boxplot(outlier.shape = NA)

p1
```

```{r}
p2 <- p1 +
  geom_jitter(width = 0.15, alpha = 0.15, size = 0.8)

p2
```

What did the points add? Did they also add clutter?

> TODO: Replace this with your answer.

# Part 4: Violin Plot

```{r}
p1 <- ggplot(adult_bp, aes(x = age_group, y = systolic_mean))

p1
```

```{r}
p2 <- p1 +
  geom_violin(fill = "gray85")

p2
```

```{r}
p3 <- p2 +
  geom_boxplot(width = 0.12, outlier.shape = NA)

p3
```

Which plot works better for this question: box plot, points, or violin? Why?

> TODO: Replace this with your answer.

# Part 5: Group Summaries

```{r}
age_group_summary <- adult_bp |>
  group_by(age_group) |>
  summarise(
    n = n(),
    mean_systolic = mean(systolic_mean),
    median_systolic = median(systolic_mean),
    iqr_systolic = IQR(systolic_mean),
    .groups = "drop"
  )

age_group_summary
```

Write a short interpretation using both the plot and the summary table.

> TODO: Replace this with your interpretation.
