---
title: "Practice: Grouped Distributions"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# Goal

Practice comparing the distribution of one numeric variable across groups.

# Suggested Data

Choose one:

- `data/real/nhanes_2021_2023_analysis.csv`
- `data/real/exercise_eating_disorders.csv`
- `data/simulated/physical_therapy_activity.csv`

Helpful module reference: `modules/03_continuous-data/02_grouped-distributions.qmd`.

# Load And Inspect

```{r}
# Load your dataset here.
```

```{r}
# Inspect numeric and grouping variables.
```

# Prepare Groups

Choose one numeric variable and one grouping variable. Keep the number of groups manageable.

```{r}
# Filter and prepare your grouped data here.
```

# Plot

Try a boxplot, violin plot, faceted histogram, or density plot.

```{r}
# Build your grouped distribution plot here.
```

# Interpret

> TODO: Which groups differ in center, spread, shape, or unusual values? What limitations should readers keep in mind?
