---
title: "Practice: Outliers And Summaries"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# Goal

Practice identifying unusual values and comparing plots with summary statistics.

# Suggested Data

Choose one:

- `data/real/nhanes_2021_2023_analysis.csv`
- `data/real/davis_height_weight.csv`
- `data/simulated/measured_vs_self_reported_exercise_minutes.csv`

Helpful module reference: `modules/03_continuous-data/03_outliers-and-summaries.qmd`.

# Load And Inspect

```{r}
# Load your dataset here.
```

```{r}
# Inspect the dataset here.
```

# Choose A Variable

Pick one numeric variable where unusual values might matter.

```{r}
# Prepare that variable here.
```

# Plot And Summarize

Create a plot that can reveal unusual values, then calculate a few summaries.

```{r}
# Plot the variable here.
```

```{r}
# Calculate summaries such as n, mean, median, min, max, and IQR.
```

# Interpret

> TODO: Are there unusual values? Do they look like errors, rare real observations, or values that need more context?
