---
title: "Practice: Outliers And Summaries"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# What We Practiced

The module showed how to:

- use summaries and plots together
- flag unusual values with a simple rule
- inspect the records behind surprising values
- avoid treating every outlier as an error

# Task

Pick one dataset and create:

1. a summary table for one numeric variable
2. one plot that shows unusual values
3. one table of flagged or extreme observations
4. a short interpretation of whether the unusual values need more context

Suggested datasets:

- `data/real/nhanes_2021_2023_analysis.csv`
- `data/real/davis_height_weight.csv`
- `data/simulated/measured_vs_self_reported_exercise_minutes.csv`

Helpful module reference: `modules/03_continuous-data/03_outliers-and-summaries.qmd`.

# Work

```{r}
# Load data, summarize a numeric variable, flag unusual values, and build one plot.
```

# Interpretation

Write briefly: what looks unusual, whether it may be an error or a real value, and what context is needed before deciding what to do.
