---
title: "Practice: Outliers And Summaries"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# Goal

Practice identifying and interpreting unusual continuous values.

# Load Data

```{r}
nhanes_path <- "data/real/nhanes_2021_2023_analysis.csv"

nhanes <- read_csv(nhanes_path, show_col_types = FALSE)

nhanes |>
  glimpse()
```

# Part 1: Prepare Data

```{r}
diet <- nhanes |>
  filter(!is.na(mean_daily_energy_kcal))

diet |>
  select(respondent_id, age_years, gender, age_group, mean_daily_energy_kcal) |>
  head()
```

What does `mean_daily_energy_kcal` represent in this teaching file?

> TODO: Replace this with your answer.

# Part 2: Plot The Distribution

```{r}
p1 <- ggplot(diet, aes(x = mean_daily_energy_kcal))

p1
```

```{r}
p2 <- p1 +
  geom_histogram(binwidth = 250)

p2
```

```{r}
p3 <- p2 +
  labs(
    title = "Distribution Of Mean Daily Energy Intake",
    x = "Mean daily energy intake, kcal",
    y = "Number of participants"
  ) +
  theme_minimal()

p3
```

Describe the shape and tail of the distribution.

> TODO: Replace this with your answer.

# Part 3: IQR Outlier Rule

```{r}
energy_cutoffs <- diet |>
  summarise(
    q1 = quantile(mean_daily_energy_kcal, 0.25),
    q3 = quantile(mean_daily_energy_kcal, 0.75),
    iqr = IQR(mean_daily_energy_kcal),
    lower_cutoff = q1 - 1.5 * iqr,
    upper_cutoff = q3 + 1.5 * iqr
  )

energy_cutoffs
```

What does the upper cutoff mean?

> TODO: Replace this with your answer.

# Part 4: Flag Unusual Values

```{r}
diet_flagged <- diet |>
  mutate(
    high_energy_outlier = mean_daily_energy_kcal > energy_cutoffs$upper_cutoff
  )

diet_flagged |>
  count(high_energy_outlier)
```

```{r}
p1 <- ggplot(
  diet_flagged,
  aes(
    x = mean_daily_energy_kcal,
    fill = high_energy_outlier
  )
) +
  geom_histogram(binwidth = 250)

p1
```

```{r}
p2 <- p1 +
  scale_fill_manual(
    values = c("TRUE" = "firebrick", "FALSE" = "gray70"),
    labels = c("FALSE" = "Not flagged", "TRUE" = "Flagged high"),
    name = NULL
  ) +
  labs(
    title = "High Energy Intake Values Flagged By IQR Rule",
    x = "Mean daily energy intake, kcal",
    y = "Number of participants"
  ) +
  theme_minimal()

p2
```

Does the flagged group look like isolated errors, a long tail, or something else?

> TODO: Replace this with your answer.

# Part 5: Inspect Rows

```{r}
diet_flagged |>
  filter(high_energy_outlier) |>
  select(respondent_id, age_years, gender, age_group, dietary_days, mean_daily_energy_kcal) |>
  arrange(desc(mean_daily_energy_kcal)) |>
  head(10)
```

What would you check before removing or changing these rows?

> TODO: Replace this with your answer.

# Checklist

- Did you describe shape, center, spread, and tail?
- Did you inspect the rows behind unusual values?
- Did you avoid calling outliers errors without evidence?
- Did you mention the teaching-file caution for dietary summaries?
