---
title: "Problem Set 2 Example Answers"
subtitle: "Instructor data suitability check"
format:
  html:
    toc: true
    embed-resources: true
---

```{r}
#| message: false
#| warning: false
library(dplyr)
library(ggplot2)


flu <- readr::read_csv("data/real/flu_vaccination_age_time_clean.csv")
```

# Task 1: Dataset Orientation

One row represents one flu vaccination estimate for a specific season, month, geography, and age group. The main time variables are `season`, `season_start_year`, and `month`. Useful grouping variables include `age_group`, `geography_type`, and `geography`.

A key caution is that estimates have confidence intervals and sample sizes. Comparisons should account for uncertainty and should not treat all estimates as equally precise.

```{r}
flu |>
  select(season, season_start_year, month, geography, age_group, estimate, ci_lower, ci_upper, sample_size) |>
  head(10)
```

# Task 2: Change Over Time

This plot uses national estimates for month 5 and compares three age groups over flu seasons.

```{r}
us_month_5_selected <- flu |>
  filter(
    geography == "United States",
    month == 5,
    age_group %in% c("6 Months - 4 Years", "18-49 Years", ">=65 Years")
  )

p1 <- ggplot(
  us_month_5_selected,
  aes(x = season_start_year, y = estimate, color = age_group)
) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(
    title = "Flu Vaccination Estimates Over Time",
    subtitle = "United States, month 5, selected age groups",
    x = "Season start year",
    y = "Vaccination estimate (%)",
    color = "Age group"
  ) +
  theme_minimal()

p1
```

A line plot is appropriate because the main task is change over ordered seasons. The plot also shows that age-group differences are large relative to most year-to-year changes.

# Task 3: Grouping Or Small Multiples

This plot compares the regional time pattern for adults ages 50-64 years.

```{r}
regions_50_64 <- flu |>
  filter(
    geography != "United States",
    month == 5,
    age_group == "50-64 Years"
  )

p2 <- ggplot(regions_50_64, aes(x = season_start_year, y = estimate)) +
  geom_line(color = "steelblue", linewidth = 0.8) +
  geom_point(color = "steelblue", size = 1.4) +
  facet_wrap(vars(geography)) +
  labs(
    title = "Regional Time Patterns For Ages 50-64",
    subtitle = "Month 5 estimates by HHS region",
    x = "Season start year",
    y = "Vaccination estimate (%)"
  ) +
  theme_minimal()

p2
```

Faceting helps compare regional patterns without putting ten colored lines in one panel. The same axes and repeated layout make the regional differences easier to scan.

# Task 4: Uncertainty Or Data Quality

This plot uses the confidence interval columns for the national estimate among adults ages 65 and older.

```{r}
us_65_uncertainty <- flu |>
  filter(
    geography == "United States",
    month == 5,
    age_group == ">=65 Years"
  )

p3 <- ggplot(us_65_uncertainty, aes(x = season_start_year, y = estimate)) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper), fill = "gray70", alpha = 0.4) +
  geom_line(color = "gray20", linewidth = 1) +
  geom_point(color = "gray20", size = 2) +
  labs(
    title = "National Estimates Include Sampling Uncertainty",
    subtitle = "United States, month 5, ages 65 and older",
    x = "Season start year",
    y = "Vaccination estimate (%)"
  ) +
  theme_minimal()

p3
```

The intervals are narrow enough that broad changes over time remain visible, but they still show that each estimate has uncertainty.

# Task 5: Rough Plot To Final Plot

Here is a rough first version of the change-over-time plot.

```{r}
rough_plot <- ggplot(us_month_5_selected, aes(x = season_start_year, y = estimate, color = age_group)) +
  geom_line() +
  geom_point()

rough_plot
```

Here is the revised version.

```{r}
p1
```

The revised version adds a title, subtitle, axis labels, legend title, stronger line weight, and a cleaner theme. Those changes make the time variable, geography, month, and compared age groups clearer.

# Data Suitability Notes

- `flu_vaccination_age_time_clean.csv` supports all required tasks.
- `season_start_year` works well for change over time.
- `age_group` and `geography` work well for grouping and faceting.
- `ci_lower`, `ci_upper`, and `sample_size` support uncertainty and data-quality checks.
- The dataset has enough seasons and groups to make students choose rather than plot everything at once.
