---
title: "Grouped Summaries"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
```

# Goal

This notebook compares groups by choosing one statistic to represent each group, then plotting those summaries.

# Analytic Questions

- How does mean systolic blood pressure differ by adult age group?
- How do we separate the statistical summary from the visual display?
- How does a dot plot compare with a bar chart for group means?

# Load Data

```{r}
nhanes <- readr::read_csv("data/real/nhanes_2021_2023_analysis.csv")
```

# Prepare Data

<!-- data-prep-anatomy-start -->
::: {.data-prep-anatomy}
| Data preparation step | Functions in this chunk |
|---|---|
| Keep or remove rows | `filter()` |
| Create or change columns | `mutate()` |
| Count or summarize | `group_by()`; `summarize()`; `n()` |
:::
<!-- data-prep-anatomy-end -->

```{r}
adult_bp <- nhanes |>
  filter(
    age_years >= 18,
    !is.na(age_group),
    !is.na(systolic_mean)
  )

bp_by_age <- adult_bp |>
  group_by(age_group) |>
  summarize(
    n = n(),
    mean_systolic = mean(systolic_mean),
    median_systolic = median(systolic_mean),
    sd_systolic = sd(systolic_mean),
    se_systolic = sd_systolic / sqrt(n),
    lower = mean_systolic - 1.96 * se_systolic,
    upper = mean_systolic + 1.96 * se_systolic,
    .groups = "drop"
  ) |>
  mutate(age_group = reorder(age_group, mean_systolic))

bp_by_age
```

These are simple unweighted mean +/- 1.96 SE intervals for teaching, not NHANES survey-adjusted confidence intervals.

# First Plot: Bars

Bars emphasize distance from zero. For group means, the main comparison is often the difference between groups.

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Aesthetic mappings | `x`; `y` |
| Geoms / stats | `geom_bar()` |
| `geom_bar()` properties | `fill` |
:::
<!-- plot-anatomy-end -->

```{r}
p1 <- ggplot(bp_by_age, aes(x = age_group, y = mean_systolic)) +
  geom_bar(stat = "identity", fill = "gray75")

p1
```

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Scales / coordinates / facets / labels | `coord_flip()`; `labs()`; `theme_minimal()` |
:::
<!-- plot-anatomy-end -->

```{r}
p2 <- p1 +
  coord_flip() +
  labs(
    title = "Mean Systolic Blood Pressure By Age Group",
    x = "Age group",
    y = "Mean systolic blood pressure"
  ) +
  theme_minimal()

p2
```

# Dot Plot With Intervals

For group means, points and intervals often focus attention on comparisons between groups.

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Aesthetic mappings | `x`; `y`; `ymin`; `ymax` |
| Geoms / stats | `geom_errorbar()`; `geom_point()` |
| `geom_errorbar()` properties | `width`; `color` |
| `geom_point()` properties | `size`; `color` |
:::
<!-- plot-anatomy-end -->

```{r}
p1 <- ggplot(bp_by_age, aes(x = age_group, y = mean_systolic)) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.15, color = "gray50") +
  geom_point(size = 3, color = "steelblue")

p1
```

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Scales / coordinates / facets / labels | `coord_flip()`; `labs()`; `theme_minimal()` |
:::
<!-- plot-anatomy-end -->

```{r}
p2 <- p1 +
  coord_flip() +
  labs(
    title = "Mean Systolic Blood Pressure Increases Across Adult Age Groups",
    subtitle = "Points show means; intervals are simple unweighted mean +/- 1.96 SE",
    x = "Age group",
    y = "Mean systolic blood pressure"
  ) +
  theme_minimal()

p2
```

Older age groups have higher mean systolic blood pressure in this sample. The intervals summarize uncertainty around each unweighted group mean, not the full spread of individual blood pressure values.

You'll notice here we added error bars. You'll want to make sure this makes sense for your case. Error bars helpfully convey uncertainty, but they also trigger a focus on statistical inference and hypothesis testing in many readers, which may be unwanted.

You'll also want to investigate what process was used to calculate the error bars. Here, `ggplot2` is not calculating the intervals. We calculated `lower` and `upper` earlier as simple unweighted mean +/- 1.96 SE intervals, then `geom_errorbar()` draws those precomputed values.  

# Check The Distribution Behind The Summary

Group means should not replace a distribution check.

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Aesthetic mappings | `x`; `y` |
| Geoms / stats | `geom_boxplot()` |
| `geom_boxplot()` properties | `fill`; `outlier.alpha` |
:::
<!-- plot-anatomy-end -->

```{r}
p1 <- ggplot(adult_bp, aes(x = age_group, y = systolic_mean)) +
  geom_boxplot(fill = "gray90", outlier.alpha = 0.3)

p2 <- p1 +
  coord_flip() +
  labs(
    title = "The Individual Values Still Vary Within Each Age Group",
    x = "Age group",
    y = "Mean systolic blood pressure"
  ) +
  theme_minimal()

p2
```

You'll notice here how the grouped box plots present the reader with a sense that the center of each group is closer together than our plot above. This is because of the x-axis. Be careful!

# Takeaways

- Decide what one plotted row should represent before plotting.
- Manual summaries make the statistical step visible.
- Dot-and-interval plots are often clearer than bars for comparing group means.
- Summary plots should be checked against the distributions behind them.
