---
title: "Practice: Annotating Data On Figures"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
```

# Goal

Practice adding labels and short annotation text to a figure.

Most of the code is complete. Your job is to decide what the reader should notice and revise the labels to support that message.

# 1. Load Data

```{r}
states <- readr::read_csv("data/real/state_population_ranks.csv")
```

# 2. Prepare A Focused Dataset

This keeps the 12 largest states and adds a shorter population label.

```{r}
top_states <- states |>
  slice_min(rank, n = 12) |>
  mutate(
    state_name = reorder(state_name, population),
    population_millions = population / 1000000,
    population_label = paste0(round(population_millions, 1), "M")
  )

top_states |>
  select(rank, state_name, population, population_label)
```

# 3. Start With A Plot

```{r}
p1 <- ggplot(top_states, aes(x = state_name, y = population_millions)) +
  geom_col(fill = "gray75") +
  coord_flip() +
  labs(
    title = "Largest States By Population",
    x = NULL,
    y = "Population (millions)"
  ) +
  theme_minimal()

p1
```

# 4. Add Data Labels

This labels every bar. It is readable here because there are only 12 bars.

```{r}
p2 <- p1 +
  geom_text(
    aes(label = population_label),
    hjust = -0.1,
    size = 3.5
  ) +
  coord_flip(ylim = c(0, 45))

p2
```

Try changing `size = 3.5` to `size = 3` or `size = 4`.

# 5. Highlight And Annotate One State

Now use color and annotation to direct attention to one state.

```{r}
annotated_states <- top_states |>
  mutate(focus = state_name == "Texas")

ggplot(annotated_states, aes(x = state_name, y = population_millions, fill = focus)) +
  geom_col() +
  geom_text(
    aes(label = population_label),
    hjust = -0.1,
    size = 3.5
  ) +
  annotate(
    geom = "label",
    x = "Texas",
    y = 36,
    label = "Texas is second\nin this dataset",
    hjust = 0,
    size = 3.5,
    fill = "white",
    color = "gray20"
  ) +
  coord_flip(ylim = c(0, 45)) +
  scale_fill_manual(values = c("TRUE" = "firebrick", "FALSE" = "gray75"), guide = "none") +
  labs(
    title = "Highlight One State And Label The Values",
    x = NULL,
    y = "Population (millions)"
  ) +
  theme_minimal()
```

Try changing the focus state from `"Texas"` to another state in the plot. Update the annotation text so it still makes sense.

# Interpretation

Write 1-2 sentences:

- What should the reader notice first?
- Which annotation helps them notice it?
- What would you remove if the figure felt cluttered?
