---
title: "Practice: Color Accessibility And Highlighting"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
```

# Goal

Practice using color to highlight one thing instead of coloring everything.

# 1. Load Data

```{r}
flu <- readr::read_csv("data/real/flu_vaccination_age_time_clean.csv")
```

# 2. Prepare A Focused Dataset

Use one geography and one age group so the practice is about color, not data wrangling.

```{r}
flu_us <- flu |>
  filter(
    geography == "United States",
    age_group == "18-49 Years",
    season_start_year >= 2018,
    !is.na(estimate)
  )
```

# 3. Too Many Colors

This plot gives every flu season its own color. The colors are doing too much work.

```{r}
ggplot(
  flu_us,
  aes(x = month, y = estimate, color = season, group = season)
) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = c(1, 3, 5, 7, 9, 11)) +
  labs(
    title = "Flu Vaccination Estimates By Season",
    subtitle = "United States, adults 18-49 years",
    x = "Month",
    y = "Vaccination estimate (%)",
    color = "Season"
  ) +
  theme_minimal()
```

# 4. Highlight One Season

Now color helps the reader focus on one season and keep the other seasons as context.

```{r}
focus_season <- max(flu_us$season_start_year)

flu_highlighted <- flu_us |>
  mutate(
    focus = if_else(season_start_year == focus_season, "Focus season", "Other seasons")
  )

ggplot(
  flu_highlighted,
  aes(x = month, y = estimate, group = season, color = focus)
) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = c("Focus season" = "steelblue", "Other seasons" = "gray80")) +
  scale_x_continuous(breaks = c(1, 3, 5, 7, 9, 11)) +
  labs(
    title = "Highlight One Season And Mute The Rest",
    subtitle = "United States, adults 18-49 years",
    x = "Month",
    y = "Vaccination estimate (%)",
    color = NULL
  ) +
  theme_minimal()
```

- Change `focus_season` to a specific year, such as `2020`.
- Change the focus color from `"steelblue"` to another high-contrast color.
- Change the age group in the filter and rerun the notebook.
