---
title: "Practice: Saving Visualizations"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
```

# What We Practiced

The module showed how to:

- save a plot object
- choose file type, width, height, and resolution
- check whether labels remain readable at the final size
- use `ggsave()` intentionally

# Task

Pick one dataset and create:

1. one polished plot object named `final_plot`
2. one saved PNG, PDF, or SVG file in a personal work folder
3. a short explanation of the file type and size choices

Suggested datasets:

- `data/real/state_population_ranks.csv`
- `data/real/flu_vaccination_age_time_clean.csv`
- `data/simulated/state_enrollment_ranks.csv`

Helpful module reference: `modules/10_communication-polish/03_saving-visualizations.qmd`.

# Work

```{r}
# Load data and build final_plot.
```

```{r}
#| eval: false
# Save final_plot with ggsave().
```

# Interpretation

Write briefly: why the file type and size were chosen and whether the saved version remains readable.
