---
title: "Geographic Comparison Without Maps"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# Goal

This notebook shows how to compare places without making a map.

The clinical examples did not include a full mapping workflow, so this module uses the same design principle as the scale and ranking examples: choose the comparison first, then choose the form.

# Analytic Questions

We will ask:

- Which counties have high current-smoking estimates?
- How do county estimates differ within a state?
- When might a ranked plot be better than a map?

# Load Data

```{r}
places_path <- "data/real/places_county_core_measures_long.csv"

places <- read_csv(places_path, show_col_types = FALSE)
```

# Prepare Data

```{r}
smoking_counties <- places |>
  filter(
    measure_key == "current_smoking",
    !is.na(estimate),
    !is.na(county_name),
    !is.na(state_name)
  )

smoking_counties |>
  select(state_name, county_name, estimate, ci_lower, ci_upper) |>
  head(10)
```

# Ranked Counties

```{r}
top_smoking <- smoking_counties |>
  slice_max(estimate, n = 20) |>
  mutate(place = paste(county_name, state, sep = ", ")) |>
  mutate(place = reorder(place, estimate))

p1 <- ggplot(top_smoking, aes(x = place, y = estimate)) +
  geom_col(fill = "steelblue")

p1
```

```{r}
p2 <- p1 +
  coord_flip() +
  labs(
    title = "Counties With High Current-Smoking Estimates",
    subtitle = "PLACES county estimates; top 20 in this file",
    x = NULL,
    y = "Current smoking estimate (%)"
  ) +
  theme_minimal()

p2
```

# Within-State Comparison

Maps are often useful, but a ranked plot can be clearer when the task is comparing values.

```{r}
maryland_smoking <- smoking_counties |>
  filter(state == "MD") |>
  mutate(county_name = reorder(county_name, estimate))

p1 <- ggplot(maryland_smoking, aes(x = county_name, y = estimate)) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2, color = "gray60") +
  geom_point(color = "steelblue", size = 2)

p2 <- p1 +
  coord_flip() +
  labs(
    title = "Current-Smoking Estimates Across Maryland Counties",
    subtitle = "Points show estimates; intervals show PLACES confidence intervals",
    x = NULL,
    y = "Current smoking estimate (%)"
  ) +
  theme_minimal()

p2
```

# Small Multiples By Measure

```{r}
selected_measures <- places |>
  filter(
    state == "MD",
    measure_key %in% c("current_smoking", "diabetes", "frequent_mental_distress"),
    !is.na(estimate)
  ) |>
  group_by(measure_key) |>
  slice_max(estimate, n = 8) |>
  ungroup() |>
  mutate(county_name = reorder(county_name, estimate))

p1 <- ggplot(selected_measures, aes(x = county_name, y = estimate)) +
  geom_col(fill = "gray70") +
  coord_flip() +
  facet_wrap(vars(measure), scales = "free_y")

p2 <- p1 +
  labs(
    title = "Top Maryland Counties Differ By Health Measure",
    x = NULL,
    y = "Estimate (%)"
  ) +
  theme_minimal()

p2
```

# Takeaways

- Spatial questions do not always require maps.
- Ranked plots are useful when the task is comparing place values.
- Include uncertainty when it is available.
- Maps emphasize geography; ranked plots emphasize order and difference.
