---
title: "Data Visualization: Day 3"
subtitle: "Continuous Data And Diagnostic Graphics"
author: "Erik Westlund"
format: revealjs
---

```{r}
#| include: false
if (basename(getwd()) == "slides") {
  knitr::opts_knit$set(root.dir = dirname(getwd()))
}

library(dplyr)
library(ggplot2)
```

# Welcome Back

## Today's Goals

1. Describe continuous variables using center, spread, shape, and unusual values.
2. Use histograms, density plots, boxplots, and violin plots to inspect distributions.
3. Use quantiles to describe position and spread without assuming symmetry.
4. Use QQ plots as a practical check for distributional assumptions.
5. Compare raw and transformed variables to see whether structure becomes clearer.

## Where We Are

Day 1: workflow and first plots.

Day 2: data preparation and categorical data.

Day 3: continuous variables.

## EDA Basics for Continuous Variables

For continuous variables, ask:

- What does one row represent?
- What unit is the variable measured in?
- What is typical?
- How much spread is there?
- What is the shape?
- Are there unusual values?
- Is the variable best analyzed in its native distribution, or should it be transformed?

# Continuous One-Variable EDA

## Describe The Distribution

We usually describe:

- center
- spread
- shape
- unusual values

## Center

Center asks what is the most typical value in a dataset. 

Common summaries:

- mean
- median
- mode

Plotting can help us see when the mean and median differ due to skew, or whether a distribution has more than one peak (e.g., it is bimodal).

## Spread

Spread asks how much variation there is.

Common summaries:

- range
- interquartile range
- standard deviation / variance

Plots help us see whether spread is symmetric, uneven, or dominated by a few values.

## Shape

Shape includes:

- skew
- tails
- gaps
- peaks
- heaping
- zero inflation
- multiple modes

Without plotting, it is often hard to catch these issues.

# Techniques

## Histograms

Histograms show how numeric values are distributed across bins.

The bin width changes what you can see.

- Small bins show more local detail.
- Large bins smooth over more detail.
- Neither is automatically correct.

## Example: Histogram

```{r}
#| echo: false
#| message: false
#| warning: false
davis <- readr::read_csv("data/real/davis_height_weight.csv") |>
  filter(
    !is.na(height_cm),
    !likely_swapped_measured_values
  )

ggplot(davis, aes(x = height_cm)) +
  geom_histogram(binwidth = 3, fill = "gray75", color = "white") +
  labs(
    title = "Measured Height",
    x = "Measured height, cm",
    y = "Number of people"
  ) +
  theme_minimal(base_size = 18)
```

## Density Plots

- Density plots smooth the distribution.
- They can make shape easier to see.
- They are more abstract than histograms.

Use them carefully with:

- small samples
- bounded variables
- variables with impossible regions
- conclusions that depend on small bumps

## Example: Histogram And Density

```{r}
#| echo: false
#| message: false
#| warning: false
ggplot(davis, aes(x = height_cm)) +
  geom_histogram(
    aes(y = after_stat(density)),
    binwidth = 3,
    fill = "gray85",
    color = "white"
  ) +
  geom_density(color = "steelblue", linewidth = 1.2) +
  labs(
    title = "Histogram With Density Overlay",
    x = "Measured height, cm",
    y = "Density"
  ) +
  theme_minimal(base_size = 18)
```

## Different Displays Show Different Parts

For one continuous variable, we can use several displays.

- Histogram: counts in bins
- Density plot: smoothed shape
- Boxplot: median, middle spread, and flagged points
- Violin plot: smoothed shape and spread together

Each of these plots highlight different features of data.

## Transition To Module

Open:

`modules/04_continuous-data/01_histograms-and-density.qmd`

Watch for:

- bin width
- `mean()`, `median()`, `IQR()`, and `sd()`
- center, spread, and shape
- what density emphasizes differently

## Practice 1: Histograms And Density

Open:

`practice/03-continuous-data/01_histograms-and-summaries-practice.qmd`

Goal:

- create histograms
- change one numeric variable
- change one bin width
- explain what changed

# Spread, Boxplots, And Violins

## Boxplots Emphasize Spread

A boxplot is a compact summary of a distribution.

It shows:

- the median
- the middle 50% of values
- the overall reach of typical values
- points that are unusual by the boxplot rule

## Example: Boxplot

```{r}
#| echo: false
#| message: false
#| warning: false
exercise <- readr::read_csv("data/real/blackmore_davis_exercise_eating_disorders.csv") |>
  filter(age_years == 14, !is.na(exercise_hours_per_week))

ggplot(exercise, aes(x = group, y = exercise_hours_per_week)) +
  geom_boxplot(fill = "gray90", outlier.alpha = 0.35) +
  labs(
    title = "Exercise Hours By Group At Age 14",
    x = NULL,
    y = "Exercise hours per week"
  ) +
  theme_minimal(base_size = 18)
```

## Violin Plots Emphasize Shape And Spread

A violin plot uses smoothing to show where values are concentrated.

It can help us see:

- wide versus narrow parts of the distribution
- skew
- multiple peaks
- differences in spread that a boxplot may hide

But it is still a smoothed display, so small bumps should not be overinterpreted.

## Example: Violin Plot

```{r}
#| echo: false
#| message: false
#| warning: false
ggplot(exercise, aes(x = group, y = exercise_hours_per_week)) +
  geom_violin(fill = "gray85", color = "gray40") +
  geom_boxplot(width = 0.12, outlier.shape = NA, fill = "white") +
  labs(
    title = "Violin Plus Boxplot",
    x = NULL,
    y = "Exercise hours per week"
  ) +
  theme_minimal(base_size = 18)
```

## Making Sense Of Unusual Points

Start with spread:

- How much do values vary?
- Are most values tightly clustered or widely spread?
- Are the tails short or long?
- Does one side stretch farther than the other?

## Sources Of Unusual Points

Unusual values might be:

- data entry problems
- rare but real observations
- important cases
- signs that units or transformations need attention
- values needing domain context

## What To Do With Unusual Points

Before removing anything, ask:

- Is the value possible?
- Does it influence results?
- Does it point to a data-quality issue?
- Does it suggest that a transformation might help?

Unusual values are not necessarily problematic.

## Transition To Module

Open:

`modules/04_continuous-data/02_boxplots-violin-plots-and-outliers.qmd`

Watch for:

- how boxplots connect to quartiles and IQR
- how changing the boxplot rule changes flagged points
- how jittered values change the interpretation
- how violin plots show shape differently from boxplots

## Practice 2: Boxplots And Unusual Values

Open:

`practice/03-continuous-data/02_boxplots-and-outliers-practice.qmd`

Goal:

- create a boxplot
- inspect flagged values
- change the grouping variable or boxplot rule
- explain whether the flagged values look isolated or part of a tail

# Quantiles, QQ Plots, And Transformations

## Quantiles

Quantiles describe positions in a distribution.

Examples:

- 0.25 quantile: 25% of values are at or below this point
- 0.50 quantile: the median
- 0.75 quantile: 75% of values are at or below this point

Quantiles help us describe data without assuming the distribution is symmetric. They are extremely useful for comparing distributions.

## More Than Quartiles

Quartiles are useful, but we can also inspect other positions.

Common choices:

- 10th percentile
- 25th percentile
- 50th percentile
- 75th percentile
- 90th percentile

These can show whether one tail stretches farther than the other.

## QQ Plots

- A QQ plot compares the quantiles of the observed data to the quantiles of a reference distribution.
- This makes them useful for assessing the plausibility that data come from similar "data-generating mechanisms"
- This often means asking: Is this data plausibly normally distributed? 
- Normality is not the only diagnostic QQ plots are useful for, but we'll focus there today.

## Example: QQ Plot

```{r}
#| echo: false
#| message: false
#| warning: false
recovery <- readr::read_csv("data/real/post_coma_recovery.csv") |>
  filter(!is.na(days_post_coma), days_post_coma > 0)

ggplot(recovery, aes(sample = days_post_coma)) +
  stat_qq() +
  stat_qq_line(color = "firebrick", linewidth = 1) +
  labs(
    title = "QQ Plot Of Days Post-Coma",
    x = "Theoretical normal quantiles",
    y = "Observed days post-coma quantiles"
  ) +
  theme_minimal(base_size = 18)
```

## Reading A QQ Plot

- Points roughly along the line: closer to the reference distribution.
- Curving away from the line: skew or tail differences.
- Strong departures at the ends: unusual tails or extreme values.

Do not treat a QQ plot as a pass/fail test. It is a diagnostic graphic.

## Transformations

Transformations can make structure easier to see.

Common reasons to try one:

- the variable is strongly right-skewed
- differences are multiplicative rather than additive
- a few large values dominate the scale
- a normal approximation is more plausible after transformation

## Log Transformations

Logs are useful for positive, right-skewed variables.

But:

- zeros need special care
- negative values cannot be logged directly
- interpretation changes after transformation

QQ plots can then be used to answer: "did the transformation make the structure easier to understand?"

## Example: Raw And Log Scales

```{r}
#| echo: false
#| message: false
#| warning: false
recovery_scales <- recovery |>
  transmute(
    `Raw days` = days_post_coma,
    `Log days` = log(days_post_coma)
  ) |>
  tidyr::pivot_longer(
    cols = everything(),
    names_to = "scale",
    values_to = "value"
  ) |>
  mutate(
    scale = factor(scale, levels = c("Raw days", "Log days"))
  )

ggplot(recovery_scales, aes(x = value)) +
  geom_histogram(bins = 30, fill = "gray75", color = "white") +
  facet_wrap(~ scale, scales = "free_x") +
  labs(
    title = "Raw And Log Days Post-Coma",
    x = NULL,
    y = "Number of assessments"
  ) +
  theme_minimal(base_size = 18)
```

## Transition To Module

Open:

`modules/04_continuous-data/03_quantiles-qq-plots-and-transformations.qmd`

Watch for:

- selected percentiles
- QQ plots with `stat_qq()` and `stat_qq_line()`
- raw versus log-transformed distributions
- how transformations change interpretation

## Practice 3: Diagnostics

Open:

`practice/03-continuous-data/03_quantiles-qq-transformations-practice.qmd`

Goal:

- create a QQ plot
- try one variable change
- compare raw and transformed displays
- write a cautious interpretation

# Wrap-Up

## What You Should Be Able To Do

After today, you should be able to:

- describe center, spread, shape, and unusual values
- explain why bin width changes a histogram
- connect boxplots to quartiles and IQR
- use quantiles to inspect a distribution
- read QQ plots as practical diagnostic graphics
- compare raw and transformed variables
- explain why transformations change interpretation

## Next Time

We will move from one-variable displays to group comparisons, differences, and associations between variables.
