---
title: "Data Visualization: Day 4"
subtitle: "Group Comparison And Association"
author: "Erik Westlund"
format: revealjs
---

```{r}
#| include: false
if (basename(getwd()) == "slides") {
  knitr::opts_knit$set(root.dir = dirname(getwd()))
}

library(dplyr)
library(ggplot2)
```

# Welcome Back

## Where We Are

- Day 1: workflow and first plots
- Day 2: data preparation and categorical data
- Day 3: continuous distributions and diagnostics

## Today's Goals

1. Compare groups using visualization.
2. Decide when the question is about a distribution, a summary, or a difference.
3. Plot differences directly when the gap between two values is the quantity of interest.
4. Visualize association between two numeric variables.
5. Add smoothing, grouping, or model outputs.

## Core Questions

- How do groups differ?
- Are we comparing raw values, summaries, or differences?
- How do two numeric variables move together?
- Does that association differ across groups?

# Group Comparison

## Two Variables

Yesterday we studied single continuous variables. Today we'll focus on how to visualize these quantities when grouping is of interest.

We continue to be concerned with:

- center
- spread
- shape
- unusual values
- sample size

## Grouped Distributions

Grouped distribution plots can show whether groups differ in:

- typical values
- variability
- tails
- overlap
- unusual values

## Example: Grouped Distribution

```{r}
#| echo: false
#| message: false
#| warning: false
places <- readr::read_csv("data/real/places_county_core_measures_long.csv")

county_smoking <- places |>
  filter(measure_key == "current_smoking") |>
  filter(state %in% c("MD", "VA", "DC")) |>
  filter(!is.na(estimate))

ggplot(county_smoking, aes(x = state_name, y = estimate)) +
  geom_boxplot(fill = "gray90", outlier.alpha = 0.35) +
  geom_jitter(width = 0.12, alpha = 0.25, size = 1.2) +
  labs(
    title = "County Smoking Estimates By State",
    x = NULL,
    y = "Current smoking estimate (%)"
  ) +
  theme_minimal(base_size = 18)
```

## Visualize Early

Inspect grouped distributions before alongside group summary statistics.

1. Check group sizes.
2. Plot grouped distributions.
3. Calculate summaries.

## Example: Grouped Summary

```{r}
#| echo: false
#| message: false
#| warning: false
places <- readr::read_csv("data/real/places_county_core_measures_long.csv")

state_obesity <- places |>
  filter(measure_key == "obesity") |>
  filter(state %in% c("MD", "VA", "DC")) |>
  filter(!is.na(estimate)) |>
  group_by(state_name) |>
  summarize(
    mean_estimate = mean(estimate),
    .groups = "drop"
  ) |>
  mutate(state_name = reorder(state_name, mean_estimate))

ggplot(state_obesity, aes(x = state_name, y = mean_estimate)) +
  geom_point(size = 4, color = "steelblue") +
  coord_flip() +
  labs(
    title = "Mean County Obesity Estimate By State",
    x = NULL,
    y = "Mean county estimate (%)"
  ) +
  theme_minimal(base_size = 18)
```

## Bar Charts For Means Are Often Misguided

Bars can show group means or other measures of central tendency, but they visually imply an amount starting at zero.

For group summaries focused on means or medians, dot plots are cleaner:

- less "ink"
- easier ordering
- easier to show intervals

## Uncertainty And Sample Size

When comparing groups, ask:

- How many observations are in each group?
- Are group sizes similar?
- Are intervals available or appropriate?
- Are differences large enough to matter relative to uncertainty and sample size?

The plot should not make small, unstable differences look definitive.

## Optional Practice: Grouped Distributions

If you want more practice with grouped distributions, you can later work with:

`practice/04-comparison-and-association/01_grouped-distributions-practice.qmd`

This will help you:

- compare full distributions before summaries
- check group sizes
- describe center, spread, and unusual values

## Transition To Module

Open:

`modules/05_group-comparison/01_grouped-summaries.qmd`

Watch for:

- deciding what one plotted row represents
- calculating summaries before plotting
- comparing bars with dot-and-interval plots
- checking summary plots against the underlying distributions

## Practice 1: Grouped Summaries

Open:

`practice/04-comparison-and-association/02_grouped-summaries-practice.qmd`

Goal:

- create a grouped-summary plot
- change the measure or states being compared
- explain what each point represents

# Difference Plots

## Comparing Differences

If the question is about a difference, we can lot the difference.

- measured minus reported value
- treatment minus comparison
- before minus after
- percentage-point gap
- observed minus expected

## Example: Difference Plot

```{r}
#| echo: false
#| message: false
#| warning: false
difference_example <- data.frame(
  group = c("Clinic A", "Clinic B", "Clinic C", "Clinic D", "Clinic E", "Clinic F"),
  difference = c(-8, -3, 1, 4, 7, 11)
) |>
  arrange(difference) |>
  mutate(group = factor(group, levels = group))

ggplot(difference_example, aes(x = group, y = difference)) +
  geom_hline(yintercept = 0, color = "gray60") +
  geom_segment(aes(xend = group, y = 0, yend = difference), color = "gray70") +
  geom_point(size = 4, color = "steelblue") +
  coord_flip() +
  labs(
    title = "Change In Follow-Up Rate By Clinic",
    x = NULL,
    y = "Percentage-point change"
  ) +
  theme_minimal(base_size = 18)
```

## Zero Values

- Difference plots need a clear zero.
- Zero means no difference.
- "Good" and "bad" values should be clear to the viewer.

## Difference Plot Checklist

Before plotting a difference:

- Are the two values comparable?
- Are units the same?
- Is subtraction meaningful?
- Does positive have a clear interpretation?
- Is there a clear zero reference line?

## Transition To Module

Open:

`modules/05_group-comparison/02_difference-plots.qmd`

Watch for:

- Cleveland-style dot plots
- connected dots for paired estimates
- percentage-point differences
- ordering by the difference
- labeling direction clearly

## Practice 2: Difference Plots

Open:

`practice/04-comparison-and-association/03_difference-plots-practice.qmd`

Goal:

- calculate one difference
- plot the difference directly
- explain what positive, negative, and zero mean

# Association

## Association Questions

Association asks how variables move together.

Common questions:

- Do higher values of one variable tend to go with higher values of another?
- Is the pattern linear or curved?
- How much spread is there around the pattern?
- Are there unusual points?
- Does the pattern differ by group?

## Scatterplots

Scatterplots show two numeric variables at once.

Look for:

- direction
- form
- strength
- spread
- unusual points
- clusters or groups

## Example: Scatterplot

```{r}
#| echo: false
#| message: false
#| warning: false
countries <- readr::read_csv("data/real/infant_mortality_countries.csv") |>
  filter(!is.na(income_usd_per_capita), !is.na(infant_mortality_per_1000))

ggplot(countries, aes(x = income_usd_per_capita, y = infant_mortality_per_1000)) +
  geom_point(alpha = 0.75) +
  scale_x_log10(labels = scales::label_dollar()) +
  labs(
    title = "Income And Infant Mortality Across Countries",
    x = "Per-capita income, log scale",
    y = "Infant mortality per 1,000 live births"
  ) +
  theme_minimal(base_size = 18)
```

## Overplotting

When many points overlap, the plot can hide density.

Common fixes:

- use transparency with `alpha`
- use smaller points
- add jitter when values are integers
- facet by group
- use summaries only after checking the raw points

## Smoothers

A smoother is a visual summary of the pattern. They can help show:

- general direction
- curvature
- group differences
- outliers/high-leverage points

There are several smoother options, such as loess, GAM-based smoothers, or linear model smoothers.

## Example: Loess Smoother

```{r}
#| echo: false
#| message: false
#| warning: false
ggplot(countries, aes(x = income_usd_per_capita, y = infant_mortality_per_1000)) +
  geom_point(alpha = 0.55) +
  geom_smooth(
    method = "loess",
    se = FALSE,
    color = "steelblue",
    linewidth = 1.1
  ) +
  scale_x_log10(labels = scales::label_dollar()) +
  labs(
    title = "Loess Smoother",
    subtitle = "Flexible local smoother",
    x = "Per-capita income, log scale",
    y = "Infant mortality per 1,000 live births"
  ) +
  theme_minimal(base_size = 18)
```

## Example: GAM Smoother

```{r}
#| echo: false
#| message: false
#| warning: false
ggplot(countries, aes(x = income_usd_per_capita, y = infant_mortality_per_1000)) +
  geom_point(alpha = 0.55) +
  geom_smooth(
    method = "gam",
    formula = y ~ s(x, bs = "cs"),
    se = FALSE,
    color = "steelblue",
    linewidth = 1.1
  ) +
  scale_x_log10(labels = scales::label_dollar()) +
  labs(
    title = "GAM Smoother",
    subtitle = "Flexible model-based smoother",
    x = "Per-capita income, log scale",
    y = "Infant mortality per 1,000 live births"
  ) +
  theme_minimal(base_size = 18)
```

## Example: Linear Model Smoother

```{r}
#| echo: false
#| message: false
#| warning: false
ggplot(countries, aes(x = income_usd_per_capita, y = infant_mortality_per_1000)) +
  geom_point(alpha = 0.55) +
  geom_smooth(
    method = "lm",
    se = FALSE,
    color = "firebrick",
    linewidth = 1.1,
    linetype = "dotted"
  ) +
  scale_x_log10(labels = scales::label_dollar()) +
  labs(
    title = "Linear Model Smoother",
    subtitle = "One straight-line trend",
    x = "Per-capita income, log scale",
    y = "Infant mortality per 1,000 live births"
  ) +
  theme_minimal(base_size = 18)
```

## Example: Smoothers Together

```{r}
#| echo: false
#| message: false
#| warning: false
ggplot(countries, aes(x = income_usd_per_capita, y = infant_mortality_per_1000)) +
  geom_point(alpha = 0.25) +
  geom_smooth(
    aes(color = "Loess"),
    method = "loess",
    se = FALSE,
    linewidth = 1.1
  ) +
  geom_smooth(
    aes(color = "GAM"),
    method = "gam",
    formula = y ~ s(x, bs = "cs"),
    se = FALSE,
    linewidth = 1.1,
    linetype = "dashed"
  ) +
  geom_smooth(
    aes(color = "Linear model"),
    method = "lm",
    se = FALSE,
    linewidth = 1.1,
    linetype = "dotted"
  ) +
  scale_x_log10(labels = scales::label_dollar()) +
  scale_color_manual(
    values = c(
      "Loess" = "steelblue",
      "GAM" = "darkorange",
      "Linear model" = "firebrick"
    )
  ) +
  labs(
    title = "Three Smoothers, Same Data",
    x = "Per-capita income, log scale",
    y = "Infant mortality per 1,000 live births",
    color = "Smoother"
  ) +
  theme_minimal(base_size = 18)
```

## Grouping In Association Plots

Grouping can clarify patterns, but it can also overload the plot. Ask:

- Is there enough data in each group?
- Are the groups visually separable?
- Would facets be clearer than color?
- Does the grouped pattern change the interpretation?

## Module 3: Association

Open:

`modules/06_association/01_scatterplots-smoothing-and-groups.qmd`

Watch for:

- transparency
- log scales for skewed numeric variables
- smoothers
- grouping and faceting
- grouping without causal overclaiming

## After The Module: Practice 3

Open:

`practice/04-comparison-and-association/04_scatterplots-smoothing-and-groups-practice.qmd`

Goal:

- build a scatterplot
- add one clarifying layer
- describe association without making a causal claim

# Plotting A Model Result

## From Model To Plot

- A fitted model gives predicted values.
- We can plot those predicted values as lines.
- The points are the raw data.
- The lines are the fitted model result.

## What The Lines Mean

A line can mean different things depending on the model.

- With `y ~ x`, it is the fitted two-variable relationship.
- With `y ~ x + group`, it can be shown as one fitted line per group.
- With `y ~ x * group`, the fitted slope for `x` can differ by group.
- With more variables, we have to choose what values the other variables are held at.

Plots of models need a clear explanation of how the plotted values were generated.

## Example: Prediction Lines

```{r}
#| echo: false
#| message: false
#| warning: false
davis <- readr::read_csv("data/real/davis_height_weight.csv")

body_size <- davis |>
  filter(!is.na(height_cm), !is.na(weight_kg), !is.na(sex), !likely_swapped_measured_values)

weight_model <- lm(weight_kg ~ height_cm + sex, data = body_size)

prediction_grid <- expand.grid(
  height_cm = seq(145, 195, by = 5),
  sex = sort(unique(body_size$sex))
)

prediction_data <- bind_cols(
  prediction_grid,
  as.data.frame(predict(weight_model, newdata = prediction_grid, interval = "confidence"))
)

ggplot(body_size, aes(x = height_cm, y = weight_kg)) +
  geom_point(aes(color = sex), alpha = 0.25) +
  geom_line(
    data = prediction_data,
    aes(x = height_cm, y = fit, color = sex),
    linewidth = 1.1
  ) +
  labs(
    title = "Fitted Weight Lines By Height And Sex",
    x = "Measured height, cm",
    y = "Measured weight, kg",
    color = "Sex"
  ) +
  theme_minimal(base_size = 18)
```

## Example: Interaction Lines

```{r}
#| echo: false
#| message: false
#| warning: false
interaction_model <- lm(weight_kg ~ height_cm * sex, data = body_size)

interaction_prediction_data <- bind_cols(
  prediction_grid,
  as.data.frame(predict(interaction_model, newdata = prediction_grid, interval = "confidence"))
)

ggplot(body_size, aes(x = height_cm, y = weight_kg)) +
  geom_point(aes(color = sex), alpha = 0.2) +
  geom_line(
    data = interaction_prediction_data,
    aes(x = height_cm, y = fit, color = sex),
    linewidth = 1.1
  ) +
  labs(
    title = "Interaction Model: Slope Can Differ By Group",
    subtitle = "Predictions come from weight_kg ~ height_cm * sex",
    x = "Measured height, cm",
    y = "Measured weight, kg",
    color = "Sex"
  ) +
  theme_minimal(base_size = 18)
```

## Optional Transition To Module

If time allows, open:

`modules/06_association/02_model-output-plots.qmd`

Watch for:

- plotting raw data first
- fitting a simple model
- creating prediction data
- knowing what values the predictions hold fixed
- visualizing model interactions as predicted lines
- plotting fitted model results with raw-data context

Optional practice:

`practice/04-comparison-and-association/05_model-output-plots-practice.qmd`

# Wrap-Up

## Summary

After today, you should be able to:

- distinguish grouped distributions from grouped summaries
- choose a summary statistic that matches the question
- plot differences directly when the gap is the quantity of interest
- describe association using direction, form, strength, and spread
- use smoothers and grouping carefully

## Next Time

We will cover annotating figures, visualizing change/flow, colors & accessibility, and saving figures.
