---
title: "Data Visualization: Day 2"
subtitle: "Data Preparation And Categorical Data"
author: "Erik Westlund"
format: revealjs
---

```{r}
#| include: false
if (basename(getwd()) == "slides") {
  knitr::opts_knit$set(root.dir = dirname(getwd()))
}

library(dplyr)
library(ggplot2)
```

## Day 2 Goals

1. Use tidyverse functions to inspect, filter, summarize, and arrange data before plotting.
2. Make basic bar plots, leveraging rank-order for clarity
3. Make three-variable bar plots, using faceting by small multiples for clarity
4. Explain why position on a common scale helps comparisons.
5. Practice changing one part of a data-preparation or plotting pipeline at a time.

## Where We Are

Yesterday we worked through:

- The basics of `ggplot2` and the grammar of graphics
- Quarto notebooks
- Git and the course project
- basic visualization workflow

Today, we'll slow down and look more carefully at the data before plotting.

## Today's Structure

We will work in two cycles:

1. Data cleaning: concept, module, practice
2. Categorical variables: concept, module, practice

Continuous variables move to Day 3 so we can give the mechanics more time.

## Day 1 Pattern

We will keep using the same build-and-check workflow:

```r
p1 <- ggplot(data, aes(...))
p1

p2 <- p1 + geom_*()
p2

p3 <- p2 + labs(...)
p3
```

This pattern helps us build plots step by step and see what each step changes.

# Data Preparation

## Visualization Depends On Data Preparation

Before plotting, we need to know:

- The unit of analysis: what does one row represent?
- Are observations:

  - raw records (e.g., patient-level data)?
  - summaries (e.g., from a `summarize()` command)?
  - percentages (e.g., from pre-aggregated data)?
  - or estimates (e.g., from a regression model)?

- Do we need to filter to the level that matches the question?
- Do we need to summarize before plotting?

## A Basic Data-Cleaning Pass

Start with inspection, not editing.

```r
glimpse(data)
head(data)
names(data)
```

Ask:

- What does one row represent?
- Which variables answer the question?
- Are the names, types, units, and missing values what we expect?

## Then Clean One Issue At A Time

Common first moves:

- rename unclear variables
- convert types that imported incorrectly
- recode categories with inconsistent labels
- filter to the population or time period for the question
- check missing values before summarizing

After each change, inspect again.

```r
clean_data <- raw_data |>
  rename(...) |>
  mutate(...) |>
  filter(...)

glimpse(clean_data)
head(clean_data)
```

## Make Small, Named Tables

It is fine to save intermediate data frames when they mark a real stage:

- `raw_data`: the file as read from disk
- `clean_data`: renamed, recoded, filtered, or type-corrected data
- `plot_data`: a count or summary table made for one plot

Avoid long chains of vague objects like `data2`, `data3`, and `final_final`.

If a table is simple to recreate, save the code rather than saving another object.

## The Pipe: `|>`

The pipe sends the result of one step into the next step.

```r
data |>
  filter(...) |>
  count(...) |>
  ggplot(aes(...)) +
  geom_bar(stat = "identity")
```

Read it as: "start with this data, then do this, then do this."

## `dplyr`

- `dplyr` is the primary `tidyverse` package for data manipulation
- It helps us make small, explicit changes to a data table before plotting.
- Most `dplyr` work is about keeping rows, keeping columns, creating columns, sorting, counting, or summarizing.

## Common `dplyr` Functions

- `select()`: keep variables we need
- `filter()`: keep observations that match a condition
- `count()`: count observations by category
- `group_by()` + `summarize()`: aggregate values by group
- `mutate()`: create or recode variables
- `arrange()`: sort a table before plotting

## Common Summary Functions

- `n()`: number of observations
- `sum()`: total amount
- `mean()`: average
- `median()`: middle value
- `min()` and `max()`: range endpoints
- `IQR()` and `sd()`: spread

## Data Cleaning Module

Open:

`modules/02_data-preparation/01_data-preparation.qmd`

Watch for:

- `glimpse()`, `head()`, and `names()` before editing
- saving a few named data frames without making a new object after every line
- using each common `dplyr` function before plotting

## Practice 1: Data Cleaning

Open:

`practice/02-one-variable-and-distributions/01_data-cleaning-practice.qmd`

Goal:

- select and rename variables
- create a new variable
- filter rows
- summarize by group
- practice reordering and releveling


# Categorical Data

Categorical variables put observations into groups.

Examples:

- gender
- state
- race/ethnicity category
- age group
- blood pressure category

## First Categorical Questions

Before choosing a plot, ask what the category is being used for.

Common questions:

- How many observations are in each category?
- What total or average belongs to each category?
- Which categories are largest or smallest?
- Does the category order have meaning?

These questions require different data tables before plotting.

# Categorical Counts And Summaries

## Categorical Question 1

How many observations are in each category?

This is a count question.

Common tools:

- observation counts
- `count()`
- bar charts
- dot plots

## Count What?

Counting observations is only meaningful if the observations mean what you think they mean.

- Is one observation a person?
- Is one observation a state?
- Is one observation an already-summarized estimate?

## Example: Count Bar Chart

```{r}
#| echo: false
#| message: false
#| warning: false
titanic <- readr::read_csv("data/real/titanic_passengers.csv")

titanic |>
  filter(!is.na(passenger_class)) |>
  count(passenger_class) |>
  mutate(
    passenger_class = reorder(passenger_class, n)
  ) |>
  ggplot(aes(x = passenger_class, y = n)) +
  geom_bar(stat = "identity", fill = "gray70") +
  coord_flip() +
  labs(
    title = "Titanic Passengers By Class",
    x = NULL,
    y = "Number of passengers"
  ) +
  theme_minimal(base_size = 18)
```

## Counts Versus Summaries

Counts in `dplyr` are just fancy `table()` calls

```r
data |>
  count(category)
```

Summaries apply a statistical/mathematical operation to the data.

```r
data |>
  group_by(category) |>
  summarize(total = sum(value))
```

The first counts records. The second aggregates a numeric variable.

## Ordering Matters

Alphabetical order is only sometimes the order you want in your visualization.

Sometimes you want:

- largest to smallest
- smallest to largest
- natural sequence
- grouped by concept

This requires judgment.

## Example: Ordered Bar Chart

```{r}
#| echo: false
#| message: false
#| warning: false
states <- readr::read_csv("data/simulated/state_enrollment_ranks.csv")

states |>
  filter(rank <= 10) |>
  mutate(state_name = reorder(state_name, population)) |>
  ggplot(aes(x = state_name, y = population)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  scale_y_continuous(labels = scales::label_number(scale_cut = scales::cut_short_scale())) +
  labs(
    title = "Ten Largest Simulated State Enrollments",
    x = NULL,
    y = "Enrollment"
  ) +
  theme_minimal(base_size = 18)
```

## Rank Is Just Ordering

A ranked plot is not a special chart type.

It is a categorical plot where categories are ordered by a value:

- count
- total
- average
- percentage

Ordering by rank helps comparison, but the plotted values still need to show magnitude.

## Don't Mindlessly Rank

- Not all data should be rank-ordered
- Sometimes competing analytic needs require a different ordering, especially with three-variable visualizations

## Transition To Module

Open:

`modules/03_categorical-data/01_counts-and-ordering.qmd`

Watch for:

- what one observation represents
- when to count observations versus summarize values
- how ordering changes readability

# Relative Position On A Common Scale

## Not Just A Design Principle, But Science!

People generally compare positions on a shared, aligned scale more accurately than they compare:

- angles
- areas
- volumes
- colors
- positions on different scales

This is backed up by experimental research on graphical perception.

Source: Cleveland and McGill's work on graphical perception (1984, 1987).

## Common Scale Examples

These all use position on a common scale:

- ordered bars with a shared baseline
- dot plots on one axis
- histograms along one measurement scale
- box plots aligned to one numeric scale
- reference lines at meaningful values

## Example: Lollipop Chart

```{r}
#| echo: false
#| message: false
#| warning: false
states |>
  filter(rank <= 10) |>
  mutate(state_name = reorder(state_name, population)) |>
  ggplot(aes(x = state_name, y = population)) +
  geom_segment(aes(xend = state_name, y = 0, yend = population), color = "gray70") +
  geom_point(size = 4, color = "steelblue") +
  coord_flip() +
  scale_y_continuous(labels = scales::label_number(scale_cut = scales::cut_short_scale())) +
  labs(
    title = "Same Values, Less Ink Than Bars",
    x = NULL,
    y = "Enrollment"
  ) +
  theme_minimal(base_size = 18)
```

## When Common Scales Break Down

Comparisons get harder when:

- facets use different axes
- stacked bar segments do not share a baseline
- too many colors carry numeric meaning
- area or bubble size encodes magnitude
- the axis is cropped or transformed without explanation

These choices are not always wrong, but they need a reason.

# Composition And Proportions

## Composition Questions

Composition asks what a whole is made of.

Examples:

- race/ethnicity groups within a state
- budget categories within a total
- causes within all reported cases
- response categories within a survey item

These are denominator questions: what is the whole, and what are its parts?

## Why Stacked Bars Get Hard

A stacked bar often puts three variables into one display:

- the whole or group on one axis
- the size or share on the other axis
- the component category in color

That can show overall composition, but most segments do not share a common baseline.

## Example: Stacked Composition

```{r}
#| echo: false
#| message: false
#| warning: false
race_state <- readr::read_csv("data/simulated/state_school_enrollment_groups_long.csv")

selected_composition <- race_state |>
  filter(state_name %in% c("Texas", "Minnesota", "Montana", "West Virginia")) |>
  mutate(race_ethnicity_proportion = share)

ggplot(
  selected_composition,
  aes(x = state_name, y = race_ethnicity_proportion, fill = race_ethnicity)
) +
  geom_bar(stat = "identity") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Simulated Enrollment Composition In Four States",
    x = NULL,
    y = "Proportion",
    fill = NULL
  ) +
  theme_minimal(base_size = 18) +
  theme(legend.position = "bottom")
```

## Small Multiples As An Alternative

Small multiples split the display into repeated panels.

This can make comparisons easier when:

- one stacked chart is doing too much
- each category deserves its own scale position
- the same comparison repeats across groups

The tradeoff is space. A useful question to then ask is: do we need all the facets?

## Example: Small Multiples

```{r}
#| echo: false
#| message: false
#| warning: false
selected_composition |>
  mutate(race_ethnicity = reorder(race_ethnicity, race_ethnicity_proportion)) |>
  ggplot(aes(x = race_ethnicity, y = race_ethnicity_proportion)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  facet_wrap(~ state_name) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Same Composition Data As Small Multiples",
    x = NULL,
    y = "Proportion"
  ) +
  theme_minimal(base_size = 18)
```

## Transition To Module

Open:

`modules/03_categorical-data/02_composition-and-proportions.qmd`

Watch for:

- raw counts versus proportions
- why stacked bars can be hard to read
- how faceting creates small multiples for the same data

## Practice 2: Categorical Plotting

Open:

`practice/02-one-variable-and-distributions/02_categorical-practice.qmd`

Goal:

- load a state-level dataset
- reorder categories by a value
- make a simple ordered bar chart
- explain what one bar represents

## If You Finish Early

Revise one plot:

- reorder categories
- add clearer labels
- check the codebook
- write a more precise interpretation

# Wrap-Up

## What You Should Be Able To Do

After today, you should be able to:

- distinguish observation counts from summarized values
- use common functions to prepare data for plotting
- explain why common scales help comparisons
- order categories by rank while keeping magnitude visible
- explain why stacked bars can be hard to read
- change one part of a working pipeline and explain what changed

## Tomorrow

We will spend the day on continuous variables: histograms, density plots, boxplots, violins, quantiles, QQ plots, unusual values, and transformations.
