---
title: "Data Visualization: Day 5"
subtitle: "Communication, Polish, And Project Studio"
author: "Erik Westlund"
format: revealjs
---

```{r}
#| include: false
if (basename(getwd()) == "slides") {
  knitr::opts_knit$set(root.dir = dirname(getwd()))
}

library(dplyr)
library(ggplot2)
```

# Welcome Back

## Today's Goals

1. Add useful annotations directly to figures.
2. Show one time-series example.
3. Preview DAGs and Mermaid charts.
4. Use color and highlighting intentionally.
5. Check accessibility before calling a figure finished.
6. Save figures at useful sizes and formats.
7. Make progress on final project visualizations.

## Where We Are

This week moved from setup to visual reasoning:

- workflow
- categorical data
- continuous data
- group comparison
- association

Today has two jobs:

- practice annotating figures for communication
- quickly preview a few other visual forms
- finish with practical polish: accessibility, color, and saving

# Annotating Figures

## Annotation Directs Attention

Annotation is not decoration.

It tells the reader what to notice.

Useful annotation can mark:

- an important value
- a benchmark
- a visible change
- a group to compare
- a limitation or caution

## Example: Direct Labels

```{r}
#| echo: false
#| message: false
#| warning: false
flu <- readr::read_csv("data/real/flu_vaccination_age_time_clean.csv")

us_may <- flu |>
  filter(
    geography == "United States",
    month == 5,
    age_group %in% c("18-49 Years", "50-64 Years", ">=65 Years"),
    !is.na(estimate)
  )

latest_may <- us_may |>
  filter(season_start_year == max(season_start_year)) |>
  mutate(label = paste0(age_group, ": ", round(estimate), "%"))

ggplot(us_may, aes(x = season_start_year, y = estimate, color = age_group)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.5) +
  geom_text(
    data = latest_may,
    aes(label = label),
    hjust = 0,
    nudge_x = 0.15,
    show.legend = FALSE
  ) +
  annotate(
    geom = "label",
    x = 2011,
    y = 74,
    label = "Text should point\nto the reading task",
    hjust = 0,
    size = 4,
    fill = "white",
    color = "gray20"
  ) +
  coord_cartesian(xlim = c(2010, 2025.5), ylim = c(25, 80)) +
  labs(
    title = "Put Text Near The Data It Explains",
    x = "Season start year",
    y = "Vaccination estimate (%)"
  ) +
  theme_minimal(base_size = 18) +
  theme(legend.position = "none")
```

## Transition To Module And Practice

Open:

`modules/10_communication-polish/00_annotating-data-on-figures.qmd`

Then practice:

`practice/05-communication-and-studio/00_annotating-data-on-figures-practice.qmd`

# Quick Explanation Demos

## One Time-Series Plot

Time belongs on the x-axis when order is part of the question.

```{r}
#| echo: false
#| message: false
#| warning: false
ggplot(us_may, aes(x = season_start_year, y = estimate, color = age_group)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.5) +
  geom_text(
    data = latest_may,
    aes(label = age_group),
    hjust = 0,
    nudge_x = 0.15,
    show.legend = FALSE
  ) +
  coord_cartesian(xlim = c(2010, 2025.5), ylim = c(25, 80)) +
  labs(
    title = "A Time Series Shows Change Over Ordered Time",
    subtitle = "United States flu vaccination estimates, May (month 5)",
    x = "Season start year",
    y = "Vaccination estimate (%)"
  ) +
  theme_minimal(base_size = 18) +
  theme(legend.position = "none")
```

This is a preview only. We are not doing a time-series practice block today.

## DAG-Style Diagrams And Mermaid

Not every visual is a data plot.

Sometimes the goal is to show:

- workflow
- process
- conceptual relationships
- assumptions behind a comparison

Open:

`modules/11_extra-topics/04_dag_structures_with_mermaid.qmd`

Also useful for process charts:

`modules/09_flow/02_process-diagrams.qmd`

Today these are quick demos, not practice sections.

# Color And Highlighting

## Color Has A Job

Color should do something:

- distinguish groups
- show order
- highlight a focus
- encode magnitude
- support accessibility

If color has no job, use less of it.

## Too Many Colors

Many categorical colors can make a plot harder to read.

Consider:

- filtering
- faceting
- direct labels
- highlighting one group
- using gray for context

## Accessibility

Do not rely only on color.

Also use:

- labels
- position
- shape or line type when needed
- readable contrast
- colorblind-safe palettes
- direct labels when a legend is hard to track
- ordering that supports the comparison

## Highlighting

Highlighting is useful when there is a focus.

The goal is not to make everything colorful.

The goal is to guide attention.

## Example: Highlighting

```{r}
#| echo: false
#| message: false
#| warning: false
states <- readr::read_csv("data/real/state_population_ranks.csv")

states |>
  filter(rank <= 15) |>
  mutate(
    state_name = reorder(state_name, population),
    focus = state_name == "Texas"
  ) |>
  ggplot(aes(x = state_name, y = population, fill = focus)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "firebrick", "FALSE" = "gray75"), guide = "none") +
  scale_y_continuous(labels = scales::label_number(scale_cut = scales::cut_short_scale())) +
  labs(
    title = "Use Color To Direct Attention",
    x = NULL,
    y = "Population"
  ) +
  theme_minimal(base_size = 18)
```

## Transition To Module

Open:

`modules/10_communication-polish/02_color-accessibility-and-highlighting.qmd`

Watch for:

- default colors versus purposeful color
- highlighting one comparison
- reducing visual noise

# Saving Figures

## Export Is Part Of Design

A figure that works in the RStudio plot pane may fail in:

- slides
- a report
- a paper
- a website
- an assignment submission

Size and format matter.

## Save The Plot Object

Good workflow:

```r
final_plot <- ggplot(...) +
  ...

ggsave("practice/work/figure.png", final_plot, width = 7, height = 5, dpi = 300)
```

Build the object.

Then save it intentionally.

## Common Formats

- PNG: slides, websites, most course submissions
- PDF: vector output, print, manuscripts
- SVG: scalable web graphics
- JPEG: usually not ideal for statistical graphics

## Transition To Module

Open:

`modules/10_communication-polish/03_saving-visualizations.qmd`

Watch for:

- plot size
- file format
- saving from a named plot object

# Critique

## Critique Questions

When reviewing a figure, ask:

- What is the question?
- What should I compare?
- What is one row or unit?
- Are scales honest?
- Are labels enough?
- What uncertainty or limitation is missing?

## Common Fixes

- clearer title
- better axis labels
- fewer colors
- reordered categories
- direct highlighting
- appropriate scale
- caption with interpretation and caution

# Final Project Studio

## Final Project Logic

The final project has two phases:

1. exploratory visualization
2. focused final visualizations

The final figures should grow out of the exploration.

They should answer questions that emerged from your exploration, not be added only because they look polished.

## Studio Goals

During project time, aim to identify:

- dataset
- rough question
- 2-4 exploratory plots
- one promising final direction
- what you need help with

## Peer Feedback Prompts

When looking at a classmate's plot, ask:

- What do you think the plot is trying to show?
- What is clear?
- What is confusing?
- What comparison did your eye make first?
- What would you revise next?

# Practice

## Extra Topic Modules

These are available for later review or project-specific needs:

- `modules/11_extra-topics/01_reusable-plot-styling.qmd`
- `modules/11_extra-topics/02_iterative-refinement-aggregation-and-faceting.qmd`
- `modules/11_extra-topics/03_correlation-patterns-and-coefficient-plots.qmd`
- `modules/11_extra-topics/04_dag_structures_with_mermaid.qmd`
- `modules/11_extra-topics/05_cleveland_style_diagnostic_plots.qmd`

They are optional. Today, prioritize communication polish, accessibility, saving, and project progress.

## Today's Practice Options

Use one or more practice notebooks from:

- `practice/05-communication-and-studio/`

Or work directly on your final project.

For today's studio work, use:

`practice/work/`

## Minimum Practice Goal

By the end of class, aim to have:

1. one revised/polished figure
2. one written interpretation
3. one limitation or caution
4. one concrete next step for the final project

# Wrap-Up

## What You Should Be Able To Do

After today, you should be able to:

- revise plots for communication
- justify scale choices
- use color intentionally
- save figures deliberately
- critique a figure constructively
- connect exploratory plots to final figures

## Final Reminder

A good visualization is not just a pretty chart.

It is a visual answer to a clear question, built from appropriate data, with honest interpretation.
