---
title: "Data Visualization: Day 5"
subtitle: "Communication, Polish, And Lightning Round"
author: "Erik Westlund"
format: revealjs
---

```{r}
#| include: false
if (basename(getwd()) == "slides") {
  knitr::opts_knit$set(root.dir = dirname(getwd()))
}

library(dplyr)
library(ggplot2)
```

# Welcome Back

## Today's Goals

1. Add useful annotations directly to figures.
2. Use color, highlighting, and accessibility checks intentionally.
3. Save figures at useful sizes and formats.
4. Preview six resource topics in a lightning round.

## Where We Are

This week moved from setup to visual reasoning:

- workflow
- categorical data
- continuous data
- group comparison
- association

Today has two parts:

- first two hours: annotation, color, and saving images
- last hour: lightning round of visualization examples and wrap-up

# Annotating Figures

## Annotation Directs Attention

Annotation of a figure is one of the most powerful ways to add impact to a visualization.

Useful annotation can mark:

- an important value
- a benchmark
- a visible change
- a group to compare
- a limitation or caution

## Example: Annotation

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-height: 4.8
flu <- readr::read_csv("data/real/flu_vaccination_age_time_clean.csv")

us_may <- flu |>
  filter(
    geography == "United States",
    month == 5,
    age_group %in% c("18-49 Years", "50-64 Years", ">=65 Years"),
    !is.na(estimate)
  )

latest_may <- us_may |>
  filter(season_start_year == max(season_start_year)) |>
  mutate(
    short_age_group = recode(
      age_group,
      "18-49 Years" = "18-49",
      "50-64 Years" = "50-64",
      ">=65 Years" = "65+"
    ),
    label = paste0(short_age_group, ": ", round(estimate), "%")
  )

ggplot(us_may, aes(x = season_start_year, y = estimate, color = age_group)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  geom_vline(xintercept = 2016.5, linetype = "dashed", color = "gray35") +
  annotate(
    geom = "label",
    x = 2016.65,
    y = 77,
    label = "Something\nhappened here",
    hjust = 0,
    size = 3.2,
    fill = "white",
    color = "gray20"
  ) +
  coord_cartesian(xlim = c(2010, 2023.5), ylim = c(25, 80)) +
  labs(
    title = "Put Text Near The Data It Explains",
    x = "Season start year",
    y = "Vaccination estimate (%)",
    color = "Age group"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "right")
```

## Transition To Module And Practice

Open:

`modules/07_annotation/01_annotating-data-on-figures.qmd`

Then practice:

`practice/05-communication-and-studio/01_annotating-data-on-figures-practice.qmd`

# Color And Highlighting

## Color Needs Purpose

Color should do something:

- distinguish groups
- show order
- highlight a focus
- encode magnitude
- support accessibility

If color has no job, do not use it.

## Too Many Colors

Many categorical colors can make a plot harder to read.

Consider:

- filtering
- faceting
- direct labels
- highlighting one group
- using gray for context

## Accessibility

- Be conscious of people with vision impairments.
- When using color to differentiate groups or represent a change in magnitude, ensure you use colors that can be distinguished from each other.

## Highlighting

Highlighting a single or small group of standout entities in a figure can be an effective way to draw attention to a key finding.

## Example: Highlighting

```{r}
#| echo: false
#| message: false
#| warning: false
states <- readr::read_csv("data/real/state_population_ranks.csv")

states |>
  filter(rank <= 15) |>
  mutate(
    state_name = reorder(state_name, population),
    focus = state_name == "Texas"
  ) |>
  ggplot(aes(x = state_name, y = population, fill = focus)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "firebrick", "FALSE" = "gray75"), guide = "none") +
  scale_y_continuous(labels = scales::label_number(scale_cut = scales::cut_short_scale())) +
  labs(
    title = "Use Color To Direct Attention",
    x = NULL,
    y = "Population"
  ) +
  theme_minimal(base_size = 18)
```

## Transition To Module

Open:

`modules/08_color/01_color-accessibility-and-highlighting.qmd`

Then practice:

`practice/05-communication-and-studio/02_color-accessibility-and-highlighting-practice.qmd`

# Saving Figures

## Exporting Your Visualizations

A figure that works in the RStudio plot pane may fail in:

- slides
- a report
- a paper
- a website
- an assignment submission

Size and format matter.

## Save The Plot Object

Good workflow:

```r
final_plot <- ggplot(...) +
  ...

ggsave("practice/work/figure.png", final_plot, width = 7, height = 5, dpi = 300)
```

- Build the object and then save it with ggsave.
- Pay attention to width, height, and DPI.

## Common Formats

- PNG: slides, websites, most course submissions
- PDF: vector output, print, manuscripts
- SVG: scalable web graphics
- JPEG: almost never appropriate for data visualization

## Transition To Module

Open:

`modules/09_saving/01_saving-visualizations.qmd`

Then practice:

`practice/05-communication-and-studio/03_saving-visualizations-practice.qmd`


# Lightning Round

## Lightning 1: Time Series

Question: how does a value change over ordered time?

Resource: `modules/11_grab-bag/01_time-trends-and-small-multiples.qmd`

## Lightning 2: Before/After

Question: did something change around a reference point?

Resource: `modules/11_grab-bag/02_before-after-and-trajectories.qmd`

## Lightning 3: Process Diagrams

Use R to make it easier to draw diagrams with arrows and boxes.

Resource: `modules/11_grab-bag/03_process-diagrams.qmd`

## Lightning 4: Causal DAGs

Use R to draw directed acylic graphs (DAGs) and assess covariate adjustment sets for causal inference.

Resource: `modules/11_grab-bag/04_causal-dag-graphs.qmd`

## Lightning 5: Maps And Alternatives

Use R to draw maps; consider alternatives to choropleths, which are often misleading.

Resource: `modules/11_grab-bag/05_choropleth-maps-and-alternatives.qmd`

## Lightning 6: Cleveland-Style Diagnostics

Advanced exploratory data analysis following Cleveland.

Resource: `modules/11_grab-bag/06_cleveland_style_diagnostic_plots.qmd`

# Wrap-Up

## Data Visualization Checklist

Before you call a figure finished, check:

- **Question**: does the title or caption make the point clear?
- **Data**: is the unit of observation clear?
- **Variables**: are labels and units readable?
- **Scale**: does the axis support the claim honestly?

## Data Visualization Checklist (cont.)

- **Color**: does color have a job? Is it used effectively? 
- **Accessibility**: is the plot readable for all people in all mediums?
- **Annotation**: does text guide attention without clutter?
- **Export**: is the saved file readable at its final size?
- **Destination**: does the file format match where it is going?

## Thank You

Please feel free to reach out with any questions: ewestlund@jhu.edu