---
title: "Annotating Data On Figures"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
```

# Goal

This notebook shows how to add data labels and short annotations to a figure.

Annotation is one of the most important ways to move from an exploratory plot to a communication plot. It tells the reader what to notice.

# Analytic Questions

- What value or comparison should the reader notice first?
- Which labels help, and which labels add clutter?
- How can text on the figure reduce legend work?

# Load Data

```{r}
flu <- readr::read_csv("data/real/flu_vaccination_age_time_clean.csv")
```

# Prepare Data

<!-- data-prep-anatomy-start -->
::: {.data-prep-anatomy}
| Data preparation step | Functions in this chunk |
|---|---|
| Keep or remove rows | `filter()` |
:::
<!-- data-prep-anatomy-end -->

```{r}
us_may <- flu |>
  filter(
    geography == "United States",
    month == 5,
    age_group %in% c("18-49 Years", "50-64 Years", ">=65 Years"),
    !is.na(estimate)
  )
```

# Start With The Plot

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Aesthetic mappings | `x`; `y`; `color` |
| Geoms / stats | `geom_line()`; `geom_point()` |
| `geom_line()` properties | `linewidth` |
| `geom_point()` properties | `size` |
| Scales / coordinates / facets / labels | `coord_cartesian()`; `labs()`; `theme_minimal()` |
:::
<!-- plot-anatomy-end -->

```{r}
p1 <- ggplot(us_may, aes(x = season_start_year, y = estimate, color = age_group)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  coord_cartesian(ylim = c(25, 80)) +
  labs(
    title = "Flu Vaccination Estimates By Age Group",
    subtitle = "United States, May (month 5)",
    x = "Season start year",
    y = "Vaccination estimate (%)",
    color = "Age group"
  ) +
  theme_minimal()

p1
```

# Label The Last Values

Direct labels can reduce legend work. A common strategy is to label the end of each line instead of labeling every point.

<!-- data-prep-anatomy-start -->
::: {.data-prep-anatomy}
| Data preparation step | Functions in this chunk |
|---|---|
| Keep or remove rows | `filter()` |
| Create or change columns | `mutate()` |
:::
<!-- data-prep-anatomy-end -->

```{r}
latest_may <- us_may |>
  filter(season_start_year == max(season_start_year)) |>
  mutate(label = paste0(age_group, ": ", round(estimate), "%"))
```

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Aesthetic mappings | `label` |
| Geoms / stats | `geom_text()` |
| Scales / coordinates / facets / labels | `coord_cartesian()`; `theme()` |
:::
<!-- plot-anatomy-end -->

```{r}
p2 <- p1 +
  geom_text(
    data = latest_may,
    aes(label = label),
    hjust = 0,
    nudge_x = 0.15,
    show.legend = FALSE
  ) +
  coord_cartesian(xlim = c(2010, 2025.5), ylim = c(25, 80)) +
  theme(legend.position = "none")

p2
```

# Add A Short Note

Use annotation text sparingly. A good annotation names the point of the plot, not every fact in the plot.

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Geoms / stats | `annotate()` |
| `annotate()` properties | `geom = "label"`; `x`; `y`; `label`; `hjust`; `size`; `fill`; `color` |
:::
<!-- plot-anatomy-end -->

```{r}
p3 <- p2 +
  annotate(
    geom = "label",
    x = 2011,
    y = 74,
    label = "Older adults have the\nhighest estimates across seasons",
    hjust = 0,
    size = 3.5,
    fill = "white",
    color = "gray20"
  )

p3
```

# What To Annotate

Good annotations often mark:

- a key value
- a visible change
- a benchmark or reference line
- a group the reader should compare
- a limitation or missing context

Avoid annotating everything. If every point is labeled, the reader has to read the plot instead of seeing the pattern.

# Takeaways

- Annotation should direct attention, not decorate the plot.
- Direct labels can reduce legend work.
- Put text close to the data it explains.
- Use fewer labels than you think you need.
