---
title: "Color, Accessibility, And Highlighting"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
```

# Goal

This notebook shows how color can encode groups, highlight a comparison, or create clutter.

# Analytic Questions

- When should color encode data?
- When should color highlight one category?
- How can filtering or highlighting reduce legend work?

# Load Data

```{r}
census <- readr::read_csv("data/real/census_state_race_ethnicity_long.csv")
```

# Prepare Data

<!-- data-prep-anatomy-start -->
::: {.data-prep-anatomy}
| Data preparation step | Functions in this chunk |
|---|---|
| Keep or remove rows | `filter()` |
| Create or change columns | `mutate()` |
:::
<!-- data-prep-anatomy-end -->

```{r}
selected_states <- census |>
  filter(state %in% c("CA", "FL", "MA", "MD", "TX", "WA")) |>
  mutate(proportion_percent = share * 100)
```

# Color Encodes Categories

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Aesthetic mappings | `x`; `y`; `fill` |
| Geoms / stats | `geom_col()` |
| Scales / coordinates / facets / labels | `labs()`; `theme_minimal()` |
:::
<!-- plot-anatomy-end -->

```{r}
p1 <- ggplot(
  selected_states,
  aes(x = state_name, y = proportion_percent, fill = race_ethnicity)
) +
  geom_col()

p2 <- p1 +
  labs(
    title = "Population Composition In Selected States",
    x = NULL,
    y = "Population share (%)",
    fill = "Race/ethnicity"
  ) +
  theme_minimal()

p2
```

This uses color to encode several categories. It works, but the reader has to use the legend constantly.

# Highlight One Category

When the question is about one category, highlight that category and mute the rest.

<!-- data-prep-anatomy-start -->
::: {.data-prep-anatomy}
| Data preparation step | Functions in this chunk |
|---|---|
| Create or change columns | `mutate()`; `if_else()` |
| Count or summarize | `group_by()`; `summarize()` |
:::
<!-- data-prep-anatomy-end -->

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Scales / coordinates / facets / labels | `scale_fill_manual()`; `labs()`; `theme_minimal()` |
:::
<!-- plot-anatomy-end -->

```{r}
asian_share <- selected_states |>
  mutate(
    highlight = if_else(race_ethnicity == "Asian", "Asian", "Other categories")
  ) |>
  group_by(state_name, highlight) |>
  summarize(
    proportion_percent = sum(proportion_percent),
    .groups = "drop"
  )

p1 <- ggplot(asian_share, aes(x = state_name, y = proportion_percent, fill = highlight)) +
  geom_col(position = "stack")

p2 <- p1 +
  scale_fill_manual(values = c("Asian" = "steelblue", "Other categories" = "gray85")) +
  labs(
    title = "Highlighting One Category Changes The Reading Task",
    subtitle = "Asian population share highlighted within selected states",
    x = NULL,
    y = "Population share (%)",
    fill = NULL
  ) +
  theme_minimal()

p2
```

# Direct Comparison Without Stacking

If the task is comparing one category across states, plot only that category.

<!-- data-prep-anatomy-start -->
::: {.data-prep-anatomy}
| Data preparation step | Functions in this chunk |
|---|---|
| Keep or remove rows | `filter()` |
| Create or change columns | `mutate()` |
:::
<!-- data-prep-anatomy-end -->

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Aesthetic mappings | `x`; `y` |
| `geom_col()` properties | `fill` |
| Scales / coordinates / facets / labels | `coord_flip()`; `labs()`; `theme_minimal()` |
:::
<!-- plot-anatomy-end -->

```{r}
asian_only <- selected_states |>
  filter(race_ethnicity == "Asian") |>
  mutate(state_name = reorder(state_name, proportion_percent))

p1 <- ggplot(asian_only, aes(x = state_name, y = proportion_percent)) +
  geom_col(fill = "steelblue")

p2 <- p1 +
  coord_flip() +
  labs(
    title = "Asian Population Share In Selected States",
    x = NULL,
    y = "Population share (%)"
  ) +
  theme_minimal()

p2
```

# Match Palette To Data Type

Different data types need different color jobs.

- Qualitative palettes distinguish categories.
- Sequential palettes show low-to-high values.
- Diverging palettes show values around a meaningful middle point.

# Sequential Color

Sequential color works for ordered numeric values.

<!-- data-prep-anatomy-start -->
::: {.data-prep-anatomy}
| Data preparation step | Functions in this chunk |
|---|---|
| Keep or remove rows | `filter()` |
| Create or change columns | `mutate()` |
:::
<!-- data-prep-anatomy-end -->

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Aesthetic mappings | `x`; `y`; `fill` |
| Geoms / stats | `geom_bar()` |
| Scales / coordinates / facets / labels | `coord_flip()`; `scale_fill_viridis_c()`; `scale_y_continuous()`; `labs()`; `theme_minimal()` |
:::
<!-- plot-anatomy-end -->

```{r}
state_population <- readr::read_csv("data/real/state_population_ranks.csv") |>
  filter(rank <= 20) |>
  mutate(state_name = reorder(state_name, population))

ggplot(state_population, aes(x = state_name, y = population, fill = population)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_viridis_c(
    option = "C",
    labels = scales::label_number(scale_cut = scales::cut_short_scale())
  ) +
  scale_y_continuous(labels = scales::label_number(scale_cut = scales::cut_short_scale())) +
  labs(
    title = "Sequential Color Shows Larger And Smaller Values",
    x = NULL,
    y = "Population",
    fill = "Population"
  ) +
  theme_minimal()
```

Sequential palettes are useful when color supports a numeric scale. They are not a substitute for position; here, bar length still does most of the work.

# Diverging Color

Diverging color is useful when zero or another middle value matters.

<!-- data-prep-anatomy-start -->
::: {.data-prep-anatomy}
| Data preparation step | Functions in this chunk |
|---|---|
| Create or change columns | `mutate()` |
| Sort rows | `arrange()` |
:::
<!-- data-prep-anatomy-end -->

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Aesthetic mappings | `x`; `y`; `color`; `xend`; `yend` |
| Geoms / stats | `geom_hline()`; `geom_segment()`; `geom_point()` |
| `geom_hline()` properties | `color` |
| `geom_segment()` properties | `color` |
| `geom_point()` properties | `size` |
| Scales / coordinates / facets / labels | `coord_flip()`; `scale_color_gradient2()`; `labs()`; `theme_minimal()` |
:::
<!-- plot-anatomy-end -->

```{r}
difference_example <- tibble::tibble(
  clinic = c("Clinic A", "Clinic B", "Clinic C", "Clinic D", "Clinic E", "Clinic F"),
  difference = c(-8, -3, 1, 4, 7, 11)
) |>
  arrange(difference) |>
  mutate(clinic = factor(clinic, levels = clinic))

ggplot(difference_example, aes(x = clinic, y = difference, color = difference)) +
  geom_hline(yintercept = 0, color = "gray60") +
  geom_segment(aes(xend = clinic, y = 0, yend = difference), color = "gray75") +
  geom_point(size = 3) +
  coord_flip() +
  scale_color_gradient2(
    low = "firebrick",
    mid = "gray60",
    high = "steelblue",
    midpoint = 0
  ) +
  labs(
    title = "Diverging Color Supports A Difference Plot",
    x = NULL,
    y = "Percentage-point difference",
    color = "Difference"
  ) +
  theme_minimal()
```

Use diverging color only when the middle point has a real interpretation.

# Accessibility Checks

Before calling a figure done, ask:

- Does the plot still work in grayscale?
- Are important differences also shown by position, labels, shape, or line type?
- Is the contrast strong enough for projection or printing?
- Is the legend doing too much work?
- Would direct labels or highlighting be clearer?

# Takeaways

- Use color when it carries information or directs attention.
- Too many colors can make the legend do too much work.
- Highlighting is useful when one category is central.
- Sequential, diverging, and qualitative palettes have different jobs.
- Do not rely on color alone; use labels, ordering, contrast, or filtering so the figure still works without color.
- Sometimes the clearest color choice is a single color plus careful filtering.
