---
title: "Practice: Flow And Attrition"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# Goal

Practice showing how observations move through filters or stages.

# Suggested Data

Choose one:

- `data/real/nhanes_2021_2023_analysis.csv`
- `data/real/titanic_passengers.csv`
- `data/simulated/neighborhood_health_survey_analysis.csv`

Helpful module reference: `modules/08_flow/01_flow-and-attrition.qmd`.

# Load And Inspect

```{r}
# Load your dataset here.
```

```{r}
# Inspect variables that could define inclusion stages.
```

# Define Stages

Create a small table with one row per stage and a count for each stage.

```{r}
# Create your stage-count table here.
```

# Plot

Use a bar chart, dot plot, or connected step plot to show retention or loss.

```{r}
# Build your flow or attrition plot here.
```

# Interpret

> TODO: Where are most observations lost? What does the final analytic sample represent?
