---
title: "Process Diagrams"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
```

# Goal

This notebook shows how diagrams can clarify analytic structure before plotting data.

# Why Diagrams?

Some visualization work is about explaining a process, not plotting a measured variable.

Diagrams can help show:

- how data become an analysis file
- how participants move through stages
- how variables are conceptually related
- which comparison a final figure is meant to support

# Mermaid Flow Diagram

Mermaid is a text-based syntax for drawing diagrams. Quarto can render Mermaid diagrams directly.

```{mermaid}
flowchart LR
  A[Raw data] --> B[Clean variables]
  B --> C[Define analysis sample]
  C --> D[Explore patterns]
  D --> E[Choose final question]
  E --> F[Polish final figures]
```

# Simple Causal Structure

This is not a formal causal analysis. The arrows show assumptions we would need to defend; they do not prove that one variable causes another.

```{mermaid}
flowchart LR
  Age[Age] --> BP[Blood pressure]
  Diet[Dietary intake] --> BP
  Income[Income] --> Diet
  Income --> BP
  Sex[Sex] --> BP
```

# Build A Static Process Plot

When a diagram needs custom labels or counts, `ggplot2` can draw boxes and arrows.

Do not memorize the coordinate numbers here; the main idea is that boxes and arrows can be stored as small data frames and adjusted by previewing.

```{r}
process_steps <- tibble::tibble(
  step = c("Load", "Clean", "Explore", "Focus", "Communicate"),
  x = 1:5,
  y = 1,
  label = c(
    "Load data",
    "Clean variables",
    "Explore plots",
    "Choose question",
    "Final figures"
  )
)

arrows <- process_steps |>
  mutate(xend = lead(x), yend = lead(y)) |>
  filter(!is.na(xend))

p1 <- ggplot() +
  geom_segment(
    data = arrows,
    aes(x = x + 0.28, y = y, xend = xend - 0.28, yend = yend),
    arrow = grid::arrow(length = grid::unit(0.15, "inches")),
    color = "gray40"
  ) +
  geom_label(
    data = process_steps,
    aes(x = x, y = y, label = label),
    fill = "gray95",
    label.size = 0.3,
    size = 3.5
  )

p2 <- p1 +
  coord_cartesian(xlim = c(0.5, 5.5), ylim = c(0.7, 1.3), clip = "off") +
  labs(title = "A Visualization Workflow Diagram") +
  theme_void() +
  theme(plot.title = element_text(face = "bold"))

p2
```

# Takeaways

- Diagrams are useful when the structure is the message.
- Keep diagrams simple enough to explain verbally.
- Use formal DAG tools only when the goal is causal reasoning.
- For many classroom and report needs, Mermaid or simple `ggplot2` diagrams are enough.
