---
title: "First Visualization"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# Goal

This notebook demonstrates the basic Day 1 visualization workflow.

We will build every plot piece by piece:

1. create a plot object
2. print it so we can see what it does
3. add one layer or change
4. print it again
5. repeat

This is the workflow we will use throughout the course.

# Analytic Questions

We will use a cleaned public PRAMS dataset to ask three basic questions:

- What is the distribution of reported binge drinking before pregnancy?
- Which subgroup categories appear most often in the selected PRAMS data?
- How are reported alcohol use and binge drinking related across state/subgroup rows?

# Load Data

```{r}
prams_path <- "data/real/prams_2011_selected.csv"

prams <- read_csv(prams_path, show_col_types = FALSE)

prams |>
  glimpse()
```

Each row is a pre-aggregated percentage for one location, subgroup category, and subgroup.

This is not person-level data.

# Plot 1: Distribution

Question: what is the distribution of reported binge drinking before pregnancy?

First, filter to rows where binge drinking is observed.

```{r}
prams_binge <- prams |>
  filter(!is.na(binge_drinking_within_3_months_birth))

prams_binge |>
  select(location_abbr, subgroup_cat, subgroup, binge_drinking_within_3_months_birth) |>
  head()
```

Start with data and an aesthetic mapping.

```{r}
p1 <- ggplot(
  prams_binge,
  aes(x = binge_drinking_within_3_months_birth)
)

p1
```

Add a geom.

```{r}
p2 <- p1 +
  geom_histogram(binwidth = 5)

p2
```

Add labels.

```{r}
p3 <- p2 +
  labs(
    title = "Distribution of Reported Binge Drinking Before Pregnancy",
    subtitle = "Rows are PRAMS state/subgroup estimates",
    x = "Percent reporting binge drinking",
    y = "Number of state/subgroup rows"
  )

p3
```

Add a theme.

```{r}
p4 <- p3 +
  theme_minimal()

p4
```

Interpretation: most observed subgroup estimates are below 30 percent, but the distribution has a long right tail with some much higher estimates.

# Plot 2: Counts By Category

Question: which subgroup categories appear most often?

Start by counting rows by subgroup category.

```{r}
subgroup_counts <- prams |>
  count(subgroup_cat, sort = TRUE)

subgroup_counts
```

Start the plot with data and mappings.

```{r}
p1 <- ggplot(
  subgroup_counts,
  aes(x = subgroup_cat, y = n)
)

p1
```

Add columns.

```{r}
p2 <- p1 +
  geom_col()

p2
```

This works, but the category names are hard to read. Reorder the categories before plotting.

```{r}
subgroup_counts_ordered <- subgroup_counts |>
  mutate(subgroup_cat = reorder(subgroup_cat, n))

p3 <- ggplot(
  subgroup_counts_ordered,
  aes(x = subgroup_cat, y = n)
) +
  geom_col()

p3
```

Flip the coordinates.

```{r}
p4 <- p3 +
  coord_flip()

p4
```

Add labels and a theme.

```{r}
p5 <- p4 +
  labs(
    title = "PRAMS Rows By Subgroup Category",
    x = NULL,
    y = "Number of rows"
  ) +
  theme_minimal()

p5
```

Interpretation: subgroup categories with more subgroup levels contribute more rows to this selected PRAMS dataset.

# Plot 3: Association

Question: how are reported alcohol use and binge drinking related?

Filter to rows where both measures are observed.

```{r}
prams_alcohol <- prams |>
  filter(
    !is.na(alcohol_use_within_3_months_birth),
    !is.na(binge_drinking_within_3_months_birth)
  )

prams_alcohol |>
  select(
    location_abbr,
    subgroup_cat,
    subgroup,
    alcohol_use_within_3_months_birth,
    binge_drinking_within_3_months_birth
  ) |>
  head()
```

Start with data and mappings.

```{r}
p1 <- ggplot(
  prams_alcohol,
  aes(
    x = alcohol_use_within_3_months_birth,
    y = binge_drinking_within_3_months_birth
  )
)

p1
```

Add points.

```{r}
p2 <- p1 +
  geom_point()

p2
```

Add transparency so overlapping points are easier to see.

```{r}
p3 <- p1 +
  geom_point(alpha = 0.6)

p3
```

Add a trend line.

```{r}
p4 <- p3 +
  geom_smooth(method = "lm", se = FALSE)

p4
```

Add labels and a theme.

```{r}
p5 <- p4 +
  labs(
    title = "Alcohol Use And Binge Drinking Before Pregnancy",
    subtitle = "Each point is a PRAMS state/subgroup estimate",
    x = "Percent reporting any alcohol use",
    y = "Percent reporting binge drinking"
  ) +
  theme_minimal()

p5
```

Interpretation: rows with higher reported alcohol use tend to have higher reported binge drinking. This is an association across pre-aggregated state/subgroup estimates, not a person-level causal claim.

# Day 1 Pattern

The object-building pattern is:

```r
p1 <- ggplot(data, aes(...))
p1

p2 <- p1 + geom_*()
p2

p3 <- p2 + labs(...)
p3
```

This makes it easier to see what each line of code contributes to the figure.

# Render Check

Render this notebook and confirm that each intermediate plot appears in the HTML output.
