---
title: "First Visualization And ggplot Basics"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
```

# Goal

This notebook introduces the basic Day 1 visualization workflow.

We will build plots piece by piece:

1. load data
2. inspect the data
3. start a plot object
4. add one layer or change
5. print the plot after each step
6. interpret what the plot shows

This is the pattern we will use throughout the course.

# The Grammar Of Graphics

A visualization combines:

- data
- aesthetic mappings
- geometric objects
- statistical transformations
- scales
- coordinates
- facets
- labels and themes

`ggplot2` gives us a language for building those pieces in layers.

# Load Data

We will use a cleaned public PRAMS dataset.

```{r}
prams <- readr::read_csv("data/real/prams_2011_selected.csv")

prams |>
  glimpse()
```

Each observation in this file is a pre-aggregated percentage for one location, subgroup category, and subgroup.

This is not person-level data.

# Explore The Data Structure

Visualization starts before the first plot. First, use tables to understand what the plotted observations would represent.

**How many records are in the file, how many locations are represented, and how many subgroup categories are included?**

<!-- data-prep-anatomy-start -->
::: {.data-prep-anatomy}
| Data preparation step | Functions in this chunk |
|---|---|
| Count or summarize | `summarize()`; `n()`; `n_distinct()` |
:::
<!-- data-prep-anatomy-end -->

```{r}
prams |>
  summarize(
    records = n(),
    locations = n_distinct(location),
    subgroup_categories = n_distinct(subgroup_cat)
  )
```

**Which subgroup categories appear in the file?**

<!-- data-prep-anatomy-start -->
::: {.data-prep-anatomy}
| Data preparation step | Functions in this chunk |
|---|---|
| Count or summarize | `count()` |
:::
<!-- data-prep-anatomy-end -->

```{r}
subgroup_counts <- prams |>
  count(subgroup_cat, sort = TRUE)

subgroup_counts
```

**How many estimates does each location contribute?**

<!-- data-prep-anatomy-start -->
::: {.data-prep-anatomy}
| Data preparation step | Functions in this chunk |
|---|---|
| Count or summarize | `count()` |
:::
<!-- data-prep-anatomy-end -->

```{r}
prams |>
  count(location, sort = TRUE) |>
  head(10)
```

Each location appears many times because the file includes overall estimates and subgroup estimates. A plot of all records would mix those levels together.

The `None` subgroup category gives the overall estimate for each location. That is the right level for a first state/location comparison.

<!-- data-prep-anatomy-start -->
::: {.data-prep-anatomy}
| Data preparation step | Functions in this chunk |
|---|---|
| Keep or remove rows | `filter()` |
| Choose or rename columns | `select()` |
:::
<!-- data-prep-anatomy-end -->

```{r}
prams_overall <- prams |>
  filter(subgroup_cat == "None")

prams_overall |>
  select(
    location,
    location_abbr,
    binge_drinking_within_3_months_birth,
    alcohol_use_within_3_months_birth
  ) |>
  head()
```

For plots about overall location estimates, we will use `prams_overall`. For plots about the structure of the PRAMS file, we will use the full `prams` data.

This kind of data preparation is part of visualization. A plot can only answer the right question if the data are suited to the questions of interest. On Day 2 we will spend more time on the basic functions used to inspect, filter, summarize, and arrange data before plotting one variable at a time.

# Plot 1: Distribution

Question: what is the distribution of reported binge drinking in the 3 months before birth?

For this first plot, use the overall location estimates. That way each plotted observation represents one state/location, not a subgroup slice.

<!-- data-prep-anatomy-start -->
::: {.data-prep-anatomy}
| Data preparation step | Functions in this chunk |
|---|---|
| Keep or remove rows | `filter()` |
| Choose or rename columns | `select()` |
:::
<!-- data-prep-anatomy-end -->

```{r}
prams_binge <- prams_overall |>
  filter(!is.na(binge_drinking_within_3_months_birth))

prams_binge |>
  select(location, location_abbr, binge_drinking_within_3_months_birth) |>
  head()
```

Start with data only.

```{r}
p1 <- ggplot(data = prams_binge)

p1
```

Add an aesthetic mapping.

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Aesthetic mappings | `x` |
:::
<!-- plot-anatomy-end -->

```{r}
p2 <- ggplot(
  data = prams_binge,
  mapping = aes(x = binge_drinking_within_3_months_birth)
)

p2
```

Add a geom.

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Geoms / stats | `geom_histogram()` |
:::
<!-- plot-anatomy-end -->

```{r}
p3 <- p2 +
  geom_histogram(binwidth = 5)

p3
```

Add labels.

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Scales / coordinates / facets / labels | `labs()` |
:::
<!-- plot-anatomy-end -->

```{r}
p4 <- p3 +
  labs(
    title = "Reported Binge Drinking In The 3 Months Before Birth",
    subtitle = "Overall PRAMS state/location estimates",
    x = "Percent reporting binge drinking",
    y = "Number of states/locations"
  )

p4
```

Add a theme.

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Scales / coordinates / facets / labels | `theme_minimal()` |
:::
<!-- plot-anatomy-end -->

```{r}
p5 <- p4 +
  theme_minimal()

p5
```

Most overall state/location estimates are below 30 percent, but the distribution has a right tail with some higher estimates.

# Plot 2: Compare Categories

Question: which states or locations have higher overall reported binge drinking estimates?

First, make a small table for the comparison.

<!-- data-prep-anatomy-start -->
::: {.data-prep-anatomy}
| Data preparation step | Functions in this chunk |
|---|---|
| Choose or rename columns | `select()` |
| Sort rows | `arrange()`; `desc()` |
:::
<!-- data-prep-anatomy-end -->

```{r}
prams_binge_ordered <- prams_binge |>
  arrange(desc(binge_drinking_within_3_months_birth)) |>
  select(location, binge_drinking_within_3_months_birth)

prams_binge_ordered |>
  head(10)
```

Start with the ordered data and mappings.

<!-- data-prep-anatomy-start -->
::: {.data-prep-anatomy}
| Data preparation step | Functions in this chunk |
|---|---|
| Create or change columns | `mutate()` |
:::
<!-- data-prep-anatomy-end -->

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Aesthetic mappings | `x`; `y` |
:::
<!-- plot-anatomy-end -->

```{r}
prams_binge_ordered <- prams_binge_ordered |>
  mutate(location = reorder(location, binge_drinking_within_3_months_birth))

p1 <- ggplot(
  prams_binge_ordered,
  aes(x = location, y = binge_drinking_within_3_months_birth)
)

p1
```

Use `geom_col()` because the bar height is already in the data.

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Geoms / stats | `geom_col()` |
:::
<!-- plot-anatomy-end -->

```{r}
p2 <- p1 +
  geom_col()

p2
```

Flip the coordinates so the location names are readable.

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Scales / coordinates / facets / labels | `coord_flip()` |
:::
<!-- plot-anatomy-end -->

```{r}
p3 <- p2 +
  coord_flip()

p3
```

Add labels and a theme.

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Scales / coordinates / facets / labels | `labs()`; `theme_minimal()` |
:::
<!-- plot-anatomy-end -->

```{r}
p4 <- p3 +
  labs(
    title = "Reported Binge Drinking By Location",
    subtitle = "Overall PRAMS state/location estimates",
    x = NULL,
    y = "Percent reporting binge drinking"
  ) +
  theme_minimal()

p4
```

The plot compares overall location estimates, not individual people. Ordering the locations makes the higher and lower estimates easier to see.

# Plot 3: Association

Question: how are reported alcohol use and binge drinking related?

Use the same overall location estimates, then filter to places where both measures are observed.

<!-- data-prep-anatomy-start -->
::: {.data-prep-anatomy}
| Data preparation step | Functions in this chunk |
|---|---|
| Keep or remove rows | `filter()` |
| Choose or rename columns | `select()` |
:::
<!-- data-prep-anatomy-end -->

```{r}
prams_alcohol <- prams_overall |>
  filter(
    !is.na(alcohol_use_within_3_months_birth),
    !is.na(binge_drinking_within_3_months_birth)
  )

prams_alcohol |>
  select(
    location,
    location_abbr,
    alcohol_use_within_3_months_birth,
    binge_drinking_within_3_months_birth
  ) |>
  head()
```

Start with data and mappings.

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Aesthetic mappings | `x`; `y` |
:::
<!-- plot-anatomy-end -->

```{r}
p1 <- ggplot(
  prams_alcohol,
  aes(
    x = alcohol_use_within_3_months_birth,
    y = binge_drinking_within_3_months_birth
  )
)

p1
```

Add points.

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Geoms / stats | `geom_point()` |
:::
<!-- plot-anatomy-end -->

```{r}
p2 <- p1 +
  geom_point()

p2
```

Add transparency so overlapping points are easier to see.

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| `geom_point()` properties | `alpha` |
:::
<!-- plot-anatomy-end -->

```{r}
p3 <- p1 +
  geom_point(alpha = 0.6)

p3
```

Add a trend line.

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Geoms / stats | `geom_smooth()` |
| `geom_smooth()` properties | `method = "lm"`; `se` |
:::
<!-- plot-anatomy-end -->

```{r}
p4 <- p3 +
  geom_smooth(method = "lm", se = FALSE)

p4
```

Add labels and a theme.

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Scales / coordinates / facets / labels | `labs()`; `theme_minimal()` |
:::
<!-- plot-anatomy-end -->

```{r}
p5 <- p4 +
  labs(
    title = "Alcohol Use And Binge Drinking In The 3 Months Before Birth",
    subtitle = "Each point is one overall PRAMS state/location estimate",
    x = "Percent reporting any alcohol use",
    y = "Percent reporting binge drinking"
  ) +
  theme_minimal()

p5
```

State/location estimates with higher reported alcohol use tend to have higher reported binge drinking. This is an association across pre-aggregated location estimates, not a person-level causal claim.

# Create A Flag Before Plotting

Sometimes the variable we want to plot is not already in the data. Here, we can mark which locations are above the median binge-drinking estimate.

<!-- data-prep-anatomy-start -->
::: {.data-prep-anatomy}
| Data preparation step | Functions in this chunk |
|---|---|
| Choose or rename columns | `select()` |
| Create or change columns | `mutate()`; `if_else()` |
:::
<!-- data-prep-anatomy-end -->

```{r}
binge_median <- median(prams_binge$binge_drinking_within_3_months_birth)

prams_binge_flagged <- prams_binge |>
  mutate(
    binge_group = if_else(
      binge_drinking_within_3_months_birth > binge_median,
      "Above median",
      "At or below median"
    ),
    location = reorder(location, binge_drinking_within_3_months_birth)
  )

prams_binge_flagged |>
  select(location, binge_drinking_within_3_months_birth, binge_group) |>
  head(10)
```

Build a plot that uses the new flag for highlighting.

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Aesthetic mappings | `x`; `y`; `fill` |
:::
<!-- plot-anatomy-end -->

```{r}
p1 <- ggplot(
  prams_binge_flagged,
  aes(
    x = location,
    y = binge_drinking_within_3_months_birth,
    fill = binge_group
  )
)

p1
```

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Geoms / stats | `geom_col()` |
:::
<!-- plot-anatomy-end -->

```{r}
p2 <- p1 +
  geom_col()

p2
```

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Scales / coordinates / facets / labels | `coord_flip()` |
:::
<!-- plot-anatomy-end -->

```{r}
p3 <- p2 +
  coord_flip()

p3
```

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Scales / coordinates / facets / labels | `scale_fill_manual()`; `labs()`; `theme_minimal()` |
:::
<!-- plot-anatomy-end -->

```{r}
p4 <- p3 +
  scale_fill_manual(
    values = c("Above median" = "steelblue", "At or below median" = "gray75")
  ) +
  labs(
    title = "Reported Binge Drinking Above And Below The Median",
    subtitle = "Overall PRAMS state/location estimates",
    x = NULL,
    y = "Percent reporting binge drinking",
    fill = NULL
  ) +
  theme_minimal()

p4
```

The new variable does not change the estimates. It changes what the figure emphasizes by separating locations above the median from the rest.

# Day 1 Pattern

The object-building pattern is:

```r
p1 <- ggplot(data, aes(...))
p1

p2 <- p1 + geom_*()
p2

p3 <- p2 + labs(...)
p3
```

This makes it easier to see what each line of code contributes to the figure.

Tomorrow we will use this same pattern for counts, summaries, ordered categories, histograms, density plots, and unusual values.

# Render Check

Render this notebook and confirm that each intermediate plot appears in the HTML output.
