---
title: "First Visualization And ggplot Basics"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# Goal

This notebook introduces the basic Day 1 visualization workflow.

We will build plots piece by piece:

1. load data
2. inspect the data
3. start a plot object
4. add one layer or change
5. print the plot after each step
6. interpret what the plot shows

This is the pattern we will use throughout the course.

# The Grammar Of Graphics

A visualization combines:

- data
- aesthetic mappings
- geometric objects
- statistical transformations
- scales
- coordinates
- facets
- labels and themes

`ggplot2` gives us a language for building those pieces in layers.

# Load Data

We will use a cleaned public PRAMS dataset.

```{r}
prams_path <- "data/real/prams_2011_selected.csv"

prams <- read_csv(prams_path, show_col_types = FALSE)

prams |>
  glimpse()
```

Each row is a pre-aggregated percentage for one location, subgroup category, and subgroup.

This is not person-level data.

# Plot 1: Distribution

Question: what is the distribution of reported binge drinking before pregnancy?

First, filter to rows where binge drinking is observed.

```{r}
prams_binge <- prams |>
  filter(!is.na(binge_drinking_within_3_months_birth))

prams_binge |>
  select(location_abbr, subgroup_cat, subgroup, binge_drinking_within_3_months_birth) |>
  head()
```

Start with data only.

```{r}
p1 <- ggplot(data = prams_binge)

p1
```

Add an aesthetic mapping.

```{r}
p2 <- ggplot(
  data = prams_binge,
  mapping = aes(x = binge_drinking_within_3_months_birth)
)

p2
```

Add a geom.

```{r}
p3 <- p2 +
  geom_histogram(binwidth = 5)

p3
```

Add labels.

```{r}
p4 <- p3 +
  labs(
    title = "Distribution of Reported Binge Drinking Before Pregnancy",
    subtitle = "Rows are PRAMS state/subgroup estimates",
    x = "Percent reporting binge drinking",
    y = "Number of state/subgroup rows"
  )

p4
```

Add a theme.

```{r}
p5 <- p4 +
  theme_minimal()

p5
```

Interpretation: most observed subgroup estimates are below 30 percent, but the distribution has a long right tail with some much higher estimates.

# Plot 2: Counts By Category

Question: which subgroup categories appear most often?

`geom_bar()` counts rows inside `ggplot2`.

```{r}
p1 <- ggplot(prams, aes(x = subgroup_cat))

p1
```

```{r}
p2 <- p1 +
  geom_bar()

p2
```

That works, but the category names are hard to read.

Often it is clearer to count first, then plot the summary.

```{r}
subgroup_counts <- prams |>
  count(subgroup_cat, sort = TRUE)

subgroup_counts
```

Now one row is one subgroup category.

```{r}
p1 <- ggplot(
  subgroup_counts,
  aes(x = subgroup_cat, y = n)
)

p1
```

Use `geom_col()` because the bar height is already in the data.

```{r}
p2 <- p1 +
  geom_col()

p2
```

Reorder the categories before plotting.

```{r}
subgroup_counts_ordered <- subgroup_counts |>
  mutate(subgroup_cat = reorder(subgroup_cat, n))

p3 <- ggplot(
  subgroup_counts_ordered,
  aes(x = subgroup_cat, y = n)
) +
  geom_col()

p3
```

Flip the coordinates.

```{r}
p4 <- p3 +
  coord_flip()

p4
```

Add labels and a theme.

```{r}
p5 <- p4 +
  labs(
    title = "PRAMS Rows By Subgroup Category",
    x = NULL,
    y = "Number of rows"
  ) +
  theme_minimal()

p5
```

Interpretation: subgroup categories with more subgroup levels contribute more rows to this selected PRAMS dataset.

# Plot 3: Association

Question: how are reported alcohol use and binge drinking related?

Filter to rows where both measures are observed.

```{r}
prams_alcohol <- prams |>
  filter(
    !is.na(alcohol_use_within_3_months_birth),
    !is.na(binge_drinking_within_3_months_birth)
  )

prams_alcohol |>
  select(
    location_abbr,
    subgroup_cat,
    subgroup,
    alcohol_use_within_3_months_birth,
    binge_drinking_within_3_months_birth
  ) |>
  head()
```

Start with data and mappings.

```{r}
p1 <- ggplot(
  prams_alcohol,
  aes(
    x = alcohol_use_within_3_months_birth,
    y = binge_drinking_within_3_months_birth
  )
)

p1
```

Add points.

```{r}
p2 <- p1 +
  geom_point()

p2
```

Add transparency so overlapping points are easier to see.

```{r}
p3 <- p1 +
  geom_point(alpha = 0.6)

p3
```

Map another variable to color.

```{r}
p4 <- ggplot(
  prams_alcohol,
  aes(
    x = alcohol_use_within_3_months_birth,
    y = binge_drinking_within_3_months_birth,
    color = subgroup_cat
  )
) +
  geom_point(alpha = 0.6)

p4
```

This adds information, but it may be too much information for this plot. That is a design judgment.

Return to the simpler plot and add a trend line.

```{r}
p5 <- p3 +
  geom_smooth(method = "lm", se = FALSE)

p5
```

Add labels and a theme.

```{r}
p6 <- p5 +
  labs(
    title = "Alcohol Use And Binge Drinking Before Pregnancy",
    subtitle = "Each point is a PRAMS state/subgroup estimate",
    x = "Percent reporting any alcohol use",
    y = "Percent reporting binge drinking"
  ) +
  theme_minimal()

p6
```

Interpretation: rows with higher reported alcohol use tend to have higher reported binge drinking. This is an association across pre-aggregated state/subgroup estimates, not a person-level causal claim.

# Transform Before Plotting

Sometimes the variable we want to plot is not already in the data.

```{r}
prams_difference <- prams_alcohol |>
  mutate(
    alcohol_minus_binge = alcohol_use_within_3_months_birth -
      binge_drinking_within_3_months_birth
  )

prams_difference |>
  select(
    location_abbr,
    subgroup_cat,
    alcohol_use_within_3_months_birth,
    binge_drinking_within_3_months_birth,
    alcohol_minus_binge
  ) |>
  head()
```

Build a plot of the transformed variable.

```{r}
p1 <- ggplot(prams_difference, aes(x = alcohol_minus_binge))

p1
```

```{r}
p2 <- p1 +
  geom_histogram(binwidth = 5)

p2
```

```{r}
p3 <- p2 +
  labs(
    title = "Difference Between Any Alcohol Use And Binge Drinking",
    subtitle = "Rows are PRAMS state/subgroup estimates",
    x = "Alcohol use percentage minus binge drinking percentage",
    y = "Number of rows"
  ) +
  theme_minimal()

p3
```

# Day 1 Pattern

The object-building pattern is:

```r
p1 <- ggplot(data, aes(...))
p1

p2 <- p1 + geom_*()
p2

p3 <- p2 + labs(...)
p3
```

This makes it easier to see what each line of code contributes to the figure.

# Render Check

Render this notebook and confirm that each intermediate plot appears in the HTML output.
