---
title: "Grammar Of Graphics With ggplot2"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# Goal

This notebook introduces `ggplot2` through the grammar of graphics.

The teaching pattern is deliberate:

- create a plot object
- print it
- add one grammar component
- print it again

This lets us see what each piece contributes.

# The Grammar Of Graphics

A visualization combines:

- data
- aesthetic mappings
- geometric objects
- statistical transformations
- scales
- coordinates
- facets
- labels and themes

`ggplot2` gives us a language for building those pieces in layers.

# Load Data

```{r}
prams_path <- "data/real/prams_2011_selected.csv"

prams <- read_csv(prams_path, show_col_types = FALSE)

prams |>
  glimpse()
```

Each row is a pre-aggregated PRAMS estimate for one location/subgroup combination.

# Prepare Example Data

For the first grammar example, use rows with observed binge drinking estimates.

```{r}
prams_binge <- prams |>
  filter(!is.na(binge_drinking_within_3_months_birth))

prams_binge |>
  select(location_abbr, subgroup_cat, subgroup, binge_drinking_within_3_months_birth) |>
  head()
```

# Grammar Part 1: Data

Every `ggplot()` starts with data.

```{r}
p1 <- ggplot(data = prams_binge)

p1
```

This gives `ggplot2` a table, but it does not say what variables to use or what marks to draw.

# Grammar Part 2: Aesthetic Mappings

Aesthetic mappings tell `ggplot2` how variables map to visual features.

```{r}
p2 <- ggplot(
  data = prams_binge,
  mapping = aes(x = binge_drinking_within_3_months_birth)
)

p2
```

Now `ggplot2` knows what variable belongs on the x-axis, but it still has no geom.

# Grammar Part 3: Geometric Objects

Geoms draw marks.

```{r}
p3 <- p2 +
  geom_histogram(binwidth = 5)

p3
```

This plot has data, an aesthetic mapping, and a geom.

# Add Labels

Labels help the reader understand what the plot is about.

```{r}
p4 <- p3 +
  labs(
    title = "Binge Drinking Before Pregnancy",
    subtitle = "Distribution across PRAMS state/subgroup estimates",
    x = "Percent reporting binge drinking",
    y = "Number of state/subgroup rows"
  )

p4
```

# Add A Theme

Themes change the non-data styling.

```{r}
p5 <- p4 +
  theme_minimal()

p5
```

# A Different Mapping: Two Variables

Now use two continuous variables.

```{r}
prams_alcohol <- prams |>
  filter(
    !is.na(alcohol_use_within_3_months_birth),
    !is.na(binge_drinking_within_3_months_birth)
  )
```

Start with data only.

```{r}
p1 <- ggplot(prams_alcohol)

p1
```

Add mappings.

```{r}
p2 <- ggplot(
  prams_alcohol,
  aes(
    x = alcohol_use_within_3_months_birth,
    y = binge_drinking_within_3_months_birth
  )
)

p2
```

Add a point geom.

```{r}
p3 <- p2 +
  geom_point()

p3
```

# More Aesthetic Mappings

Map another variable to color.

```{r}
p4 <- ggplot(
  prams_alcohol,
  aes(
    x = alcohol_use_within_3_months_birth,
    y = binge_drinking_within_3_months_birth,
    color = subgroup_cat
  )
) +
  geom_point()

p4
```

This adds information, but it may be too much information for this plot. That is a design judgment.

# Layers

Layers are added with `+`.

```{r}
p5 <- p3 +
  geom_smooth(method = "lm", se = FALSE)

p5
```

The first layer draws points. The second layer draws a fitted line.

# Statistical Transformations Inside ggplot2

Some geoms calculate summaries internally.

`geom_bar()` counts rows.

```{r}
p1 <- ggplot(prams, aes(x = subgroup_cat))

p1
```

```{r}
p2 <- p1 +
  geom_bar()

p2
```

The bars are counts of rows in the original data.

# Count Yourself, Then Plot

Sometimes it is clearer to calculate the summary first.

```{r}
subgroup_counts <- prams |>
  count(subgroup_cat, sort = TRUE)

subgroup_counts
```

Now one row is one subgroup category.

```{r}
p1 <- ggplot(subgroup_counts, aes(x = subgroup_cat, y = n))

p1
```

Use `geom_col()` because the bar height is already in the data.

```{r}
p2 <- p1 +
  geom_col()

p2
```

Reorder and flip the coordinates.

```{r}
subgroup_counts <- subgroup_counts |>
  mutate(subgroup_cat = reorder(subgroup_cat, n))

p3 <- ggplot(subgroup_counts, aes(x = subgroup_cat, y = n)) +
  geom_col()

p3
```

```{r}
p4 <- p3 +
  coord_flip()

p4
```

# Transformations Before Plotting

Sometimes the variable we want to plot is not already in the data.

```{r}
prams_transformed <- prams_alcohol |>
  mutate(
    alcohol_minus_binge = alcohol_use_within_3_months_birth -
      binge_drinking_within_3_months_birth
  )

prams_transformed |>
  select(
    location_abbr,
    subgroup_cat,
    alcohol_use_within_3_months_birth,
    binge_drinking_within_3_months_birth,
    alcohol_minus_binge
  ) |>
  head()
```

Build a plot of the transformed variable.

```{r}
p1 <- ggplot(prams_transformed, aes(x = alcohol_minus_binge))

p1
```

```{r}
p2 <- p1 +
  geom_histogram(binwidth = 5)

p2
```

# Aggregation Before Plotting

Many plots summarize rows into groups.

```{r}
state_summary <- prams_binge |>
  group_by(location_abbr, location) |>
  summarise(
    avg_binge = mean(binge_drinking_within_3_months_birth, na.rm = TRUE),
    n_rows = n(),
    .groups = "drop"
  ) |>
  arrange(desc(avg_binge))

state_summary |>
  head()
```

Now one row is one location.

```{r}
p1 <- ggplot(state_summary, aes(x = location, y = avg_binge))

p1
```

```{r}
p2 <- p1 +
  geom_col()

p2
```

Reorder, flip, and label.

```{r}
state_summary <- state_summary |>
  mutate(location = reorder(location, avg_binge))

p3 <- ggplot(state_summary, aes(x = location, y = avg_binge)) +
  geom_col()

p3
```

```{r}
p4 <- p3 +
  coord_flip()

p4
```

```{r}
p5 <- p4 +
  labs(
    title = "Average Binge Drinking Estimate By Location",
    subtitle = "Average across observed PRAMS subgroup rows",
    x = NULL,
    y = "Average percent"
  ) +
  theme_minimal()

p5
```

# Scales

Scales control how data values become visual values.

```{r}
p1 <- ggplot(
  prams_alcohol,
  aes(
    x = alcohol_use_within_3_months_birth,
    y = binge_drinking_within_3_months_birth,
    color = subgroup_cat == "None"
  )
) +
  geom_point(alpha = 0.7)

p1
```

```{r}
p2 <- p1 +
  scale_color_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "gray70"),
    labels = c("FALSE" = "Subgroup row", "TRUE" = "Overall row"),
    name = "Row type"
  )

p2
```

# Facets

Facets create small multiples.

```{r}
p1 <- prams_alcohol |>
  filter(subgroup_cat %in% c("None", "Maternal Age (3 Levels)", "Maternal Education")) |>
  ggplot(
    aes(
      x = alcohol_use_within_3_months_birth,
      y = binge_drinking_within_3_months_birth
    )
  ) +
  geom_point(alpha = 0.7)

p1
```

```{r}
p2 <- p1 +
  facet_wrap(~ subgroup_cat)

p2
```

# Day 1 Takeaways

- Build plots as objects so you can inspect each step.
- `ggplot2` is a grammar, not a chart menu.
- Data defines rows and variables.
- Aesthetic mappings connect variables to visual features.
- Geoms draw marks.
- Stats can summarize data inside the plot.
- Transformations and aggregations often happen before plotting.
- Scales, coordinates, facets, labels, and themes refine the result.
