---
title: "Practice: Data Preparation"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
```

# Goal

Practice the mechanics of basic data preparation one step at a time.

Most chunks already run. Your job is to run the code, inspect the output, and change one small part.

You will practice:

- `select()`
- `rename()`
- `mutate()`
- `filter()`
- `count()`
- `group_by()` and `summarize()`
- `reorder()`
- `factor(..., levels = c(...))`

# Data

Use NHANES. This file has one row per NHANES participant in the teaching dataset.

```{r}
dat_raw <- readr::read_csv("data/real/nhanes_2021_2023_analysis.csv")

dat_raw |>
  glimpse()
```

# 1. Select Columns

This keeps a smaller set of columns.

```{r}
dat <- dat_raw |>
  select(
    respondent_id,
    gender,
    age_years,
    race_ethnicity,
    pulse_1,
    pulse_2,
    pulse_3,
    adult_bp_category_simplified
  )

dat |>
  head()
```

Try adding `systolic_mean` to the selected columns.

# 2. Rename Columns

Notice that `rename()` uses `new_name = old_name`.

```{r}
dat_named_example <- dat |>
  rename(
    id = respondent_id,
    bp_category = adult_bp_category_simplified
  )

dat_named_example |>
  names()
```

Try adding `age = age_years` to the `rename()` call.

For the rest of the practice, use the shorter names below.

```{r}
dat <- dat_raw |>
  select(
    respondent_id,
    gender,
    age_years,
    race_ethnicity,
    pulse_1,
    pulse_2,
    pulse_3,
    adult_bp_category_simplified
  ) |>
  rename(
    id = respondent_id,
    bp_category = adult_bp_category_simplified
  )
```

# 3. Create Variables

This creates a mean pulse variable and a simple age group.

```{r}
dat <- dat |>
  mutate(
    pulse_mean = rowMeans(
      pick(pulse_1, pulse_2, pulse_3),
      na.rm = TRUE
    ),
    age_group = case_when(
      age_years < 18 ~ "Under 18",
      age_years >= 18 & age_years <= 44 ~ "18-44",
      age_years >= 45 & age_years <= 64 ~ "45-64",
      age_years >= 65 ~ "65+",
      TRUE ~ NA_character_
    )
  )

dat |>
  select(id, age_years, age_group, pulse_mean) |>
  head()
```

Try changing the first adult group from `18-44` to `18-39`. You will also need to change the next group so the categories do not overlap.

# 4. Filter Rows

This keeps adults ages 18-44 with non-missing pulse values.

```{r}
young_adults <- dat |>
  filter(age_years >= 18) |>
  filter(age_years <= 44) |>
  filter(!is.na(pulse_mean))

young_adults |>
  select(id, age_years, gender, pulse_mean) |>
  head()
```

Try changing the upper age cutoff from `44` to `64`.

# 5. Count Categories

This counts observations by age group.

```{r}
dat |>
  count(age_group)
```

Try counting `gender` instead of `age_group`.

# 6. Group And Summarize

This summarizes average pulse by age group.

```{r}
pulse_by_age <- dat |>
  group_by(age_group) |>
  summarize(
    mean_pulse = mean(pulse_mean, na.rm = TRUE),
    median_pulse = median(pulse_mean, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )

pulse_by_age
```

Try summarizing by `gender` instead of `age_group`.

# 7. Reorder Categories

This orders race/ethnicity categories by average pulse.

```{r}
pulse_by_race_ethnicity <- dat |>
  group_by(race_ethnicity) |>
  summarize(
    mean_pulse = mean(pulse_mean, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) |>
  mutate(
    race_ethnicity = reorder(race_ethnicity, mean_pulse)
  )

ggplot(
  pulse_by_race_ethnicity,
  aes(x = race_ethnicity, y = mean_pulse)
) +
  geom_point(size = 3) +
  coord_flip() +
  labs(
    title = "Average Pulse By Race/Ethnicity",
    x = NULL,
    y = "Average pulse"
  ) +
  theme_minimal()
```

Try ordering by `n` instead of `mean_pulse`.

# 8. Relevel Categories

This puts blood pressure categories in a meaningful order.

```{r}
dat <- dat |>
  mutate(
    bp_category = factor(
      bp_category,
      levels = c(
        "normal_range",
        "elevated_range",
        "stage_1_range",
        "stage_2_range"
      )
    )
  )

dat |>
  count(bp_category)
```

Try moving `"stage_2_range"` to the front of the levels. What changes in the output order?

# On Your Own

Choose one small task and try it with less scaffolding:

- Make a summary table of average pulse by blood pressure category.
- Filter to adults only and count blood pressure categories.
- Make a dot plot from one of your summary tables.

Write 1-2 sentences: what did one row represent before summarizing, and what does one row represent after summarizing?
