---
title: "Practice: Categorical Plots"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
```

# Goal

Practice categorical plotting by running working code and changing one small part.

You will practice:

- counting categories
- ordering bars
- plotting precomputed values
- comparing raw counts and proportions
- using facets for small multiples

# 1. Load State Population Data

```{r}
states <- readr::read_csv("data/real/state_population_ranks.csv")

states |>
  head()
```

# 2. Ordered Bars

This keeps the ten largest states and orders them by population.

```{r}
largest_states <- states |>
  filter(rank <= 10) |>
  mutate(
    state_name = reorder(state_name, population)
  )

ggplot(
  largest_states,
  aes(x = state_name, y = population)
) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(
    title = "Ten Largest States By Population",
    x = NULL,
    y = "Population"
  ) +
  theme_minimal()
```

Try changing `rank <= 10` to `rank <= 15`.

# 3. Smallest States

This uses the same pattern for the ten smallest states.

```{r}
smallest_states <- states |>
  filter(rank >= 42) |>
  mutate(
    state_name = reorder(state_name, population)
  )

ggplot(
  smallest_states,
  aes(x = state_name, y = population)
) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(
    title = "Ten Smallest States By Population",
    x = NULL,
    y = "Population"
  ) +
  theme_minimal()
```

Try changing the filter so the plot shows only the five smallest states.

# 4. Count A Category

Now switch to NHANES and count blood pressure categories.

```{r}
nhanes <- readr::read_csv("data/real/nhanes_2021_2023_analysis.csv")

bp_counts <- nhanes |>
  filter(!is.na(adult_bp_category_simplified)) |>
  count(adult_bp_category_simplified) |>
  mutate(
    adult_bp_category_simplified = reorder(adult_bp_category_simplified, n)
  )

bp_counts
```

Try counting `age_group` instead of `adult_bp_category_simplified`.

# 5. Plot Counts

This plots the precomputed counts from the previous step.

```{r}
ggplot(
  bp_counts,
  aes(x = adult_bp_category_simplified, y = n)
) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(
    title = "Adult Blood Pressure Categories",
    x = NULL,
    y = "Number of participants"
  ) +
  theme_minimal()
```

Try removing `coord_flip()`. Which version is easier to read?

# 6. Composition And Proportion

This calculates race/ethnicity proportions within selected states.

```{r}
race_state <- readr::read_csv("data/real/census_state_race_ethnicity_long.csv")

selected_state_composition <- race_state |>
  filter(state_name %in% c("California", "Texas", "Florida", "New York")) |>
  mutate(
    race_ethnicity_proportion = share
  )

ggplot(
  selected_state_composition,
  aes(x = state_name, y = race_ethnicity_proportion, fill = race_ethnicity)
) +
  geom_bar(stat = "identity") +
  labs(
    title = "Race/Ethnicity Composition In Four States",
    x = NULL,
    y = "Proportion",
    fill = NULL
  ) +
  theme_minimal()
```

Try replacing `"Florida"` with another state.

# 7. Facets

This uses small multiples instead of one stacked bar chart.

```{r}
ggplot(
  selected_state_composition,
  aes(x = race_ethnicity, y = race_ethnicity_proportion)
) +
  geom_bar(stat = "identity") +
  coord_flip() +
  facet_wrap(~ state_name) +
  labs(
    title = "Race/Ethnicity Composition In Four States",
    x = NULL,
    y = "Proportion"
  ) +
  theme_minimal()
```

Try adding one more state to `selected_state_composition` and compare the faceted plot.

# On Your Own

Choose one small task:

- Make an ordered bar chart for the 12 largest states.
- Make an ordered bar chart for one race/ethnicity category across states.
- Compare a stacked bar chart and a faceted chart using the same data.

Write 1-2 sentences: what does one bar represent in your plot?
