---
title: "Time Trends And Small Multiples"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
```

# Goal

Show change over time for one group and several groups.

# Load Data

```{r}
flu <- readr::read_csv("data/real/flu_vaccination_age_time_clean.csv")
```

# Prepare Data

Use the same month for every season.

<!-- data-prep-anatomy-start -->
::: {.data-prep-anatomy}
| Data preparation step | Functions in this chunk |
|---|---|
| Keep or remove rows | `filter()` |
| Count or summarize | `count()` |
:::
<!-- data-prep-anatomy-end -->

```{r}
us_may <- flu |>
  filter(
    geography == "United States",
    month == 5,
    !is.na(estimate)
  )

us_may |>
  count(age_group)
```

# One Group Over Time

<!-- data-prep-anatomy-start -->
::: {.data-prep-anatomy}
| Data preparation step | Functions in this chunk |
|---|---|
| Keep or remove rows | `filter()` |
:::
<!-- data-prep-anatomy-end -->

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Aesthetic mappings | `x`; `y` |
| Geoms / stats | `geom_line()`; `geom_point()` |
:::
<!-- plot-anatomy-end -->

```{r}
older_adults <- us_may |>
  filter(age_group == ">=65 Years")

p1 <- ggplot(older_adults, aes(x = season_start_year, y = estimate)) +
  geom_line() +
  geom_point()

p1
```

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Scales / coordinates / facets / labels | `labs()`; `theme_minimal()` |
:::
<!-- plot-anatomy-end -->

```{r}
p2 <- p1 +
  labs(
    title = "Flu Vaccination Estimates Over Time",
    subtitle = "United States, May (month 5), ages 65 and older",
    x = "Season start year",
    y = "Vaccination estimate (%)"
  ) +
  theme_minimal()

p2
```

# Several Groups With Color

<!-- data-prep-anatomy-start -->
::: {.data-prep-anatomy}
| Data preparation step | Functions in this chunk |
|---|---|
| Keep or remove rows | `filter()` |
:::
<!-- data-prep-anatomy-end -->

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Aesthetic mappings | `x`; `y`; `color` |
| Geoms / stats | `geom_line()`; `geom_point()` |
| `geom_line()` properties | `linewidth` |
| `geom_point()` properties | `size` |
:::
<!-- plot-anatomy-end -->

```{r}
selected_age_groups <- us_may |>
  filter(age_group %in% c("6 Months - 4 Years", "18-49 Years", "50-64 Years", ">=65 Years"))

p1 <- ggplot(
  selected_age_groups,
  aes(x = season_start_year, y = estimate, color = age_group)
) +
  geom_line(linewidth = 1) +
  geom_point(size = 2)

p2 <- p1 +
  labs(
    title = "Flu Vaccination Time Trends Differ By Age Group",
    subtitle = "United States, May (month 5)",
    x = "Season start year",
    y = "Vaccination estimate (%)",
    color = "Age group"
  ) +
  theme_minimal()

p2
```

# Small Multiples

Facets separate groups into repeated panels.

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Aesthetic mappings | `x`; `y` |
| `geom_line()` properties | `color`; `linewidth` |
| `geom_point()` properties | `color`; `size` |
| Scales / coordinates / facets / labels | `facet_wrap()`; `labs()`; `theme_minimal()` |
:::
<!-- plot-anatomy-end -->

```{r}
p1 <- ggplot(selected_age_groups, aes(x = season_start_year, y = estimate)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 1.8) +
  facet_wrap(vars(age_group))

p2 <- p1 +
  labs(
    title = "Small Multiples Separate The Age-Group Patterns",
    subtitle = "United States, May (month 5)",
    x = "Season start year",
    y = "Vaccination estimate (%)"
  ) +
  theme_minimal()

p2
```

# Takeaways

- Line plots work well when the x-axis has a meaningful order.
- Use the same month or comparable time point when comparing seasons.
- Color works for a few groups; facets often work better for many groups.

