---
title: "Before/After"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
```

# Goal

Show change around a reference day.

# Load Data

```{r}
diary <- readr::read_csv("data/real/migraine_headache_diary.csv")
```

# Prepare Data

Day 0 is the reference day. Keep days through 50 so the noisy tail does not dominate the plot.

<!-- data-prep-anatomy-start -->
::: {.data-prep-anatomy}
| Data preparation step | Functions in this chunk |
|---|---|
| Keep or remove rows | `filter()` |
| Create or change columns | `mutate()` |
| Count or summarize | `group_by()`; `summarize()`; `n()` |
:::
<!-- data-prep-anatomy-end -->

```{r}
daily_summary <- diary |>
  filter(!is.na(day), day <= 50, !is.na(headache_reported)) |>
  mutate(period = if_else(day < 0, "Before day 0", "Day 0 and after")) |>
  group_by(day, period) |>
  summarize(
    n = n(),
    headache_percent = mean(headache_reported) * 100,
    .groups = "drop"
  )

daily_summary |>
  head(10)
```

Check `n`; small counts make percentages noisy.

# Before And After Day 0

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Aesthetic mappings | `x`; `y`; `color` |
| Geoms / stats | `geom_line()`; `geom_point()` |
:::
<!-- plot-anatomy-end -->

```{r}
p1 <- ggplot(daily_summary, aes(x = day, y = headache_percent, color = period)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2)

p1
```

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Geoms / stats | `geom_vline()` |
| `geom_vline()` properties | `color`; `linetype` |
| Scales / coordinates / facets / labels | `labs()`; `theme_minimal()`; `theme()` |
:::
<!-- plot-anatomy-end -->

```{r}
p2 <- p1 +
  geom_vline(xintercept = 0, color = "gray50", linetype = "dashed") +
  labs(
    title = "Headache Reporting Before And After Day 0",
    subtitle = "Days after 50 removed",
    x = "Diary day",
    y = "Headache reported (%)",
    color = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

p2
```

This could represent a null finding: headache reporting looks similar before and after day 0.

# Takeaways

- Mark reference points directly on time plots.
- A before/after plot can show little or no change.
