---
title: "Process Diagrams"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
```

# Goal

Show simple process diagrams. R and `ggplot2` can be an effective way to declaratively create figures without point-and-click frustration.

# Mermaid Flow Diagram

Quarto can render Mermaid diagrams directly.

```{mermaid}
flowchart LR
  A[Raw data] --> B[Clean variables]
  B --> C[Define analysis sample]
  C --> D[Explore patterns]
  D --> E[Choose final question]
  E --> F[Polish final figures]
```

# Simple Causal Structure

```{mermaid}
flowchart LR
  Age[Age] --> BP[Blood pressure]
  Diet[Dietary intake] --> BP
  Income[Income] --> Diet
  Income --> BP
  Sex[Sex] --> BP
```

# Build A Static Process Plot

Use `ggplot2` when you want custom labels, counts, or styling.

<!-- data-prep-anatomy-start -->
::: {.data-prep-anatomy}
| Data preparation step | Functions in this chunk |
|---|---|
| Keep or remove rows | `filter()` |
| Create or change columns | `mutate()` |
:::
<!-- data-prep-anatomy-end -->

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Aesthetic mappings | `x`; `y`; `xend`; `yend`; `label` |
| Geoms / stats | `geom_segment()`; `geom_label()` |
| `geom_segment()` properties | `color` |
| `geom_label()` properties | `fill`; `linewidth`; `size` |
| Scales / coordinates / facets / labels | `coord_cartesian()`; `labs()`; `theme_void()`; `theme()` |
:::
<!-- plot-anatomy-end -->

```{r}
#| fig-width: 12
#| fig-height: 2
#| out-width: "100%"
process_steps <- tibble::tibble(
  step = c("Load", "Clean", "Explore", "Focus", "Communicate"),
  x = c(1, 3, 5, 7, 9),
  y = 1,
  label = c(
    "Load data",
    "Clean variables",
    "Explore plots",
    "Choose question",
    "Final figures"
  )
)

arrows <- process_steps |>
  mutate(xend = lead(x), yend = lead(y)) |>
  filter(!is.na(xend))

p1 <- ggplot() +
  geom_segment(
    data = arrows,
    aes(x = x + 0.65, y = y, xend = xend - 0.65, yend = yend),
    arrow = grid::arrow(length = grid::unit(0.15, "inches")),
    color = "gray40"
  ) +
  geom_label(
    data = process_steps,
    aes(x = x, y = y, label = label),
    fill = "gray95",
    linewidth = 0.3,
    size = 3.5
  )

p2 <- p1 +
  coord_cartesian(xlim = c(0.2, 9.8), ylim = c(0.85, 1.15), clip = "off") +
  labs(title = "A Visualization Workflow Diagram") +
  theme_void() +
  theme(plot.title = element_text(face = "bold"))

p2
```

This is compressed to fit in the notebook; typically you'd want it bigger.

# A Vertical Data Process Diagram

Use this when each step updates the analysis dataset.

The main path shows the updated dataset. Side boxes show records removed at each step.

```{mermaid}
flowchart TB
  X0["X0: All records<br/>n = 11933"] --> F1{"Keep adults age 18+"}
  F1 --> X1["X1: Adults<br/>n = 8153"]
  F1 --> R1["Removed: age < 18<br/>n = 3780"]

  X1 --> F2{"Require systolic BP"}
  F2 --> X2["X2: Has systolic BP<br/>n = 6123"]
  F2 --> R2["Removed: missing BP<br/>n = 2030"]

  X2 --> F3{"Require diet summary"}
  F3 --> X3["X3: Has diet summary<br/>n = 4868"]
  F3 --> R3["Removed: missing diet<br/>n = 1255"]

  X3 --> F4{"Require income-to-poverty"}
  F4 --> X4["X4: Final analysis dataset<br/>n = 4303"]
  F4 --> R4["Removed: missing income-to-poverty<br/>n = 565"]
```


