---
title: "Practice: Geographic Comparison Without Maps"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# What We Practiced

The module showed how to:

- compare places without defaulting to a map
- filter to one measure or region
- use ranked or interval plots for place comparisons
- include uncertainty when available

# Task

Pick one dataset and create:

1. a focused place-comparison table
2. one non-map geographic comparison plot
3. a short interpretation of which places stand out

Suggested datasets:

- `data/real/places_county_core_measures_long.csv`
- `data/real/epa_air_quality_county_2025_long.csv`
- `data/simulated/county_wellness_indicators_long.csv`

Helpful module reference: `modules/07_space/01_geographic-comparison-without-maps.qmd`.

# Work

```{r}
# Load data, focus the place comparison, and build one non-map plot.
```

# Interpretation

Write briefly: which places stand out, what the plot makes easy to compare, and what context is missing without a map.
