---
title: "Practice: Geographic Comparison Without Maps"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# Goal

Practice comparing places without making a map.

# Suggested Data

Choose one:

- `data/real/places_county_core_measures_long.csv`
- `data/real/epa_air_quality_county_2025_long.csv`
- `data/simulated/county_wellness_indicators_long.csv`

Helpful module reference: `modules/07_space/01_geographic-comparison-without-maps.qmd`.

# Load And Inspect

```{r}
# Load your dataset here.
```

```{r}
# Identify place names, state/county variables, measures, and estimates.
```

# Focus The Comparison

Choose one measure and a manageable set of places.

```{r}
# Filter to one measure and selected places here.
```

# Plot

Use a ranked dot plot, interval plot, or small multiples.

```{r}
# Build your geographic comparison plot here.
```

# Interpret

> TODO: Which places stand out? What context is missing without a map?
