---
title: "Counts And Ordering"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# Goal

This notebook shows how to visualize categorical counts and order categories so the plot answers a clearer question.

It also introduces the data verbs we will use throughout the course: `select()`, `count()`, `group_by()`, `summarize()`, `arrange()`, and `mutate()`.

# Analytic Questions

We will use a cleaned Census race/ethnicity dataset to ask:

- How many state-category observations appear for each race/ethnicity category?
- Which race/ethnicity categories account for the largest total population in this file?
- How does ordering change what the reader can see?

# Load Data

```{r}
census_path <- "data/real/census_state_race_ethnicity_long.csv"

census <- read_csv(census_path, show_col_types = FALSE)

census |>
  glimpse()
```

Each observation is one race/ethnicity category within one state.

Here, "included" means the race/ethnicity categories present in this teaching file. Check the codebook before treating them as every possible category.

# Explore The Structure

Before plotting, use a small table to check the structure of the file.

```{r}
census |>
  summarize(
    observations = n(),
    states = n_distinct(state),
    categories = n_distinct(race_ethnicity)
  )
```

Use `select()` when you want to inspect only the variables relevant to the next question.

```{r}
census |>
  select(state_name, race_ethnicity, race_ethnicity_population, share) |>
  head()
```

# Count Observations

First, count how many observations appear for each race/ethnicity category.

```{r}
category_counts <- census |>
  count(race_ethnicity, sort = TRUE)

category_counts
```

Start with data and mappings.

```{r}
p1 <- ggplot(
  category_counts,
  aes(x = race_ethnicity, y = n)
)

p1
```

Add columns.

```{r}
p2 <- p1 +
  geom_col()

p2
```

This plot is not very informative because every category appears once per state.

It is still useful as a reminder: before plotting counts, ask what the observations mean.

# Summarize Population By Category

A categorical summary that better matches the question is total included population by race/ethnicity category.

This is aggregation: `group_by()` defines the groups, and `summarize()` reduces many state-category observations into one value per category.

```{r}
category_population <- census |>
  group_by(race_ethnicity) |>
  summarize(
    total_population = sum(race_ethnicity_population),
    .groups = "drop"
  ) |>
  arrange(desc(total_population))

category_population
```

Now each summarized observation is one race/ethnicity category.

```{r}
p1 <- ggplot(
  category_population,
  aes(x = race_ethnicity, y = total_population)
)

p1
```

```{r}
p2 <- p1 +
  geom_col()

p2
```

# Order The Categories

The labels are hard to read, and alphabetical order does not help answer the ranking question.

Reorder by total population. This uses `mutate()` to change the category order before plotting.

```{r}
category_population_ordered <- category_population |>
  mutate(race_ethnicity = reorder(race_ethnicity, total_population))

p3 <- ggplot(
  category_population_ordered,
  aes(x = race_ethnicity, y = total_population)
) +
  geom_col()

p3
```

Flip the coordinates so the category labels are readable.

```{r}
p4 <- p3 +
  coord_flip()

p4
```

Add labels and a theme.

```{r}
p5 <- p4 +
  labs(
    title = "Included Population By Race/Ethnicity Category",
    subtitle = "Census 2020 state race/ethnicity population counts in the course data",
    x = NULL,
    y = "Population count"
  ) +
  theme_minimal()

p5
```

Interpretation: the largest category in this file is White, followed by Hispanic or Latino and Black. The plot is a count of people in the included categories, not a count of states.

# Add A Design Choice

Columns are common for counts, but a dot plot uses less ink and can make rank comparisons easier once the categories are summarized.

```{r}
p1 <- ggplot(
  category_population_ordered,
  aes(x = total_population, y = race_ethnicity)
)

p1
```

```{r}
p2 <- p1 +
  geom_point(size = 3)

p2
```

```{r}
p3 <- p2 +
  labs(
    title = "Included Population By Race/Ethnicity Category",
    x = "Population count",
    y = NULL
  ) +
  theme_minimal()

p3
```

# Takeaways

- Categorical plots depend on what one observation represents.
- `count()` and `geom_bar()` count observations.
- `group_by()` + `summarize()` aggregate values by group.
- `geom_col()` uses heights already calculated in the data.
- Ordering categories often makes the plot easier to interpret.
- A count of observations is not the same thing as a count of people, events, or cases.
