---
title: "Composition And Proportions"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# Goal

This notebook shows how to visualize composition: what a whole is made of.

Composition questions usually require proportions, not only counts.

# Analytic Questions

We will ask:

- What is the race/ethnicity composition of selected states?
- How do proportions differ from raw counts?
- When does a stacked bar help, and when does it make comparison harder?

# Load Data

```{r}
census_path <- "data/real/census_state_race_ethnicity_long.csv"

census <- read_csv(census_path, show_col_types = FALSE)

census |>
  glimpse()
```

Each row is one race/ethnicity category within one state.

# Choose A Few States

Use a small number of states so the composition plot is readable.

```{r}
selected_states <- census |>
  filter(state %in% c("CA", "FL", "MD", "MA", "TX", "WA"))

selected_states |>
  select(state_name, race_ethnicity, race_ethnicity_population, share) |>
  head()
```

# Raw Counts

Start with counts.

```{r}
p1 <- ggplot(
  selected_states,
  aes(
    x = state_name,
    y = race_ethnicity_population,
    fill = race_ethnicity
  )
)

p1
```

```{r}
p2 <- p1 +
  geom_col()

p2
```

Add labels and a theme.

```{r}
p3 <- p2 +
  labs(
    title = "Race/Ethnicity Counts In Selected States",
    subtitle = "Raw counts mix composition with state size",
    x = NULL,
    y = "Population count",
    fill = "Category"
  ) +
  theme_minimal()

p3
```

Interpretation: California and Texas have larger bars partly because they have larger populations. That makes raw counts a poor choice if the question is composition.

# Proportions

The `share` variable gives each category's share within the included race/ethnicity categories for that state.

```{r}
p1 <- ggplot(
  selected_states,
  aes(
    x = state_name,
    y = share,
    fill = race_ethnicity
  )
)

p1
```

```{r}
p2 <- p1 +
  geom_col()

p2
```

Format the y-axis as percentages.

```{r}
p3 <- p2 +
  scale_y_continuous(
    labels = function(x) paste0(round(x * 100), "%")
  )

p3
```

Add labels and a theme.

```{r}
p4 <- p3 +
  labs(
    title = "Race/Ethnicity Composition In Selected States",
    subtitle = "Each bar sums to 100% of included categories",
    x = NULL,
    y = "Share of included race/ethnicity population",
    fill = "Category"
  ) +
  theme_minimal()

p4
```

# Make One Comparison Easier

Stacked bars show overall composition, but they make it hard to compare categories that do not share a common baseline.

If the question is about one category, filter to that category.

```{r}
hispanic_share <- census |>
  filter(race_ethnicity == "Hispanic or Latino") |>
  mutate(state_name = reorder(state_name, share))

hispanic_share |>
  select(state_name, share) |>
  head()
```

```{r}
p1 <- ggplot(
  hispanic_share,
  aes(x = state_name, y = share)
)

p1
```

```{r}
p2 <- p1 +
  geom_col(fill = "steelblue")

p2
```

```{r}
p3 <- p2 +
  coord_flip() +
  scale_y_continuous(
    labels = function(x) paste0(round(x * 100), "%")
  ) +
  labs(
    title = "Hispanic Or Latino Share By State",
    subtitle = "Share of included race/ethnicity population",
    x = NULL,
    y = "Share"
  ) +
  theme_minimal()

p3
```

Interpretation: filtering to one category makes the cross-state comparison easier than reading that category out of every stacked bar.

# Takeaways

- Composition asks what a whole is made of.
- Raw counts and proportions answer different questions.
- Stacked bars are useful for showing a whole, but weak for comparing many individual segments.
- When the question is about one category, plot that category directly.
