---
title: "Ranked Comparisons"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# Goal

This notebook shows how to make ranked categorical comparisons.

Ranked plots are useful when the question is about largest, smallest, highest, lowest, or relative position.

# Analytic Questions

We will ask:

- Which states have the largest populations?
- How does a ranked plot differ from an alphabetical plot?
- How can we show the top 15 while making clear they are a subset of all states/DC?

# Load Data

```{r}
population_path <- "data/real/state_population_ranks.csv"

state_population <- read_csv(population_path, show_col_types = FALSE)

state_population |>
  glimpse()
```

Each row is one state or the District of Columbia.

# Start With An Alphabetical Plot

```{r}
p1 <- ggplot(
  state_population,
  aes(x = state_name, y = population)
)

p1
```

```{r}
p2 <- p1 +
  geom_col()

p2
```

This plot has too many labels, and alphabetical order does not fit the ranking question.

# Keep The Top 15

For a focused ranking, keep the 15 largest states.

```{r}
top_states <- state_population |>
  arrange(desc(population)) |>
  slice_head(n = 15) |>
  mutate(
    state_name = reorder(state_name, population),
    population_millions = population / 1000000
  )

top_states
```

```{r}
p1 <- ggplot(
  top_states,
  aes(x = state_name, y = population_millions)
)

p1
```

```{r}
p2 <- p1 +
  geom_col(fill = "gray60")

p2
```

Flip the coordinates.

```{r}
p3 <- p2 +
  coord_flip()

p3
```

Add labels and a theme.

```{r}
p4 <- p3 +
  labs(
    title = "Largest States By 2020 Population",
    subtitle = "Top 15 of 51 state/DC rows in the course population file",
    x = NULL,
    y = "Population, in millions"
  ) +
  theme_minimal()

p4
```

# Use A Dot Plot For Rank

A dot plot can work well when precise bar length is less important than relative position.

```{r}
p1 <- ggplot(
  top_states,
  aes(x = population_millions, y = state_name)
)

p1
```

```{r}
p2 <- p1 +
  geom_point(size = 3)

p2
```

```{r}
p3 <- p2 +
  labs(
    title = "Largest States By 2020 Population",
    subtitle = "Top 15 of 51 state/DC rows",
    x = "Population, in millions",
    y = NULL
  ) +
  theme_minimal()

p3
```

# Highlight A Category

Highlighting can guide attention without removing the comparison group.

```{r}
top_states <- top_states |>
  mutate(is_california = state == "CA")

p1 <- ggplot(
  top_states,
  aes(
    x = population_millions,
    y = state_name,
    color = is_california
  )
) +
  geom_point(size = 3)

p1
```

```{r}
p2 <- p1 +
  scale_color_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "gray65"),
    labels = c("FALSE" = "Other top-15 state", "TRUE" = "California"),
    name = NULL
  )

p2
```

```{r}
p3 <- p2 +
  labs(
    title = "California Compared With Other Large States",
    x = "Population, in millions",
    y = NULL
  ) +
  theme_minimal()

p3
```

# Takeaways

- Ranking is a statistical choice and a design choice.
- Sort by the value that matches the question.
- Limit to a top or bottom group when all categories would be unreadable.
- Say clearly when the plot shows only a subset.
- Highlighting can focus attention while preserving context.
