---
title: "Practice: Before-After And Trajectories"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
```

# What We Practiced

The module showed how to:

- mark a reference time point
- summarize repeated observations over time
- compare group-level trajectories
- distinguish individual paths from group summaries

# Task

Pick one dataset and create:

1. a prepared before-after or trajectory table
2. one plot with a clear reference point or shared time scale
3. a short interpretation of what changes and for whom

Suggested datasets:

- `data/real/migraine_headache_diary.csv`
- `data/real/post_coma_recovery.csv`
- `data/simulated/toothache_diary.csv`

Helpful module reference: `modules/07_change/02_before-after-and-trajectories.qmd`.

# Work

```{r}
# Load data, prepare before-after or trajectory data, and build one change plot.
```

# Interpretation

Write briefly: what changes, the reference point, and whether the plot shows individual change, group change, or both.
