---
title: "Practice: Scatterplots, Smoothing, And Groups"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# Goal

Practice using scatterplots to examine association between two numeric variables.

# Suggested Data

Choose one:

- `data/real/davis_height_weight.csv`
- `data/real/nhanes_2021_2023_analysis.csv`
- `data/simulated/measured_vs_self_reported_exercise_minutes.csv`

Helpful module reference: `modules/05_association/01_scatterplots-smoothing-and-groups.qmd`.

# Load And Inspect

```{r}
# Load your dataset here.
```

```{r}
# Inspect numeric variables and possible grouping variables.
```

# Prepare Variables

Pick two numeric variables. Optionally choose a grouping variable for color or facets.

```{r}
# Filter missing values and prepare variables here.
```

# Plot

Make a scatterplot. Then try adding transparency, a smooth line, or color by group.

```{r}
# Build your scatterplot here.
```

# Interpret

> TODO: Describe direction, form, strength, groups, and any unusual points. Avoid causal language unless your design supports it.
