---
title: "Practice: Scatterplots, Smoothing, And Groups"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# What We Practiced

The module showed how to:

- plot association between two numeric variables
- use transparency for overplotting
- add a smoother as a visual summary
- use grouping only when it serves the question

# Task

Pick one dataset and create:

1. a prepared table with two numeric variables
2. one scatterplot
3. one revision using transparency, smoothing, grouping, or faceting
4. a short interpretation using non-causal language

Suggested datasets:

- `data/real/davis_height_weight.csv`
- `data/real/nhanes_2021_2023_analysis.csv`
- `data/simulated/measured_vs_self_reported_exercise_minutes.csv`

Helpful module reference: `modules/05_association/01_scatterplots-smoothing-and-groups.qmd`.

# Work

```{r}
# Load data, prepare two numeric variables, and build one or two association plots.
```

# Interpretation

Write briefly: direction, form, strength, groups, unusual points, and one causal claim to avoid.
