---
title: "Practice: Model Output Plots"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# What We Practiced

The module showed how to:

- fit a simple linear model
- create values for prediction
- plot model-based summaries
- keep model output connected to the observed data

# Task

Pick one dataset and create:

1. a simple model
2. a small table of predictions or model summaries
3. one plot of the model output
4. a short interpretation that names what the model does and does not show

Suggested datasets:

- `data/real/nhanes_2021_2023_analysis.csv`
- `data/real/davis_height_weight.csv`
- `data/simulated/neighborhood_health_survey_analysis.csv`

Helpful module reference: `modules/05_association/02_model-output-plots.qmd`.

# Work

```{r}
# Load data, fit a simple model, create model output, and plot it.
```

# Interpretation

Write briefly: what the plotted model output shows, what assumptions it inherits, and what raw data context is missing.
