---
title: "Practice: Model Output Plots"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# Goal

Practice turning simple model output into a plot readers can understand.

# Suggested Data

Choose one:

- `data/real/nhanes_2021_2023_analysis.csv`
- `data/real/davis_height_weight.csv`
- `data/simulated/neighborhood_health_survey_analysis.csv`

Helpful module reference: `modules/05_association/02_model-output-plots.qmd`.

# Load And Inspect

```{r}
# Load your dataset here.
```

```{r}
# Inspect variables that could go into a simple linear model.
```

# Fit A Simple Model

Choose one outcome and one or two predictors. Keep the model simple.

```{r}
# Fit your model here with lm().
```

# Create Values To Plot

Make a small prediction dataset or coefficient summary.

```{r}
# Create model output or predictions to plot here.
```

# Plot

```{r}
# Plot the model output here.
```

# Interpret

> TODO: What does the plotted model output show? What does it not show?
