---
title: "Difference Plots"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# Goal

This notebook compares paired estimates first, then plots the calculated gap directly.

# Analytic Questions

We will ask:

- How do PRAMS alcohol-use estimates differ from binge-drinking estimates?
- Which states have the largest gaps between the two estimates?
- Why might a difference plot be clearer than two bars per state?

# Load Data

```{r}
prams_path <- "data/real/prams_2011_selected.csv"

prams <- read_csv(prams_path, show_col_types = FALSE)
```

# Prepare Data

PRAMS values are already aggregated percentages for state/subgroup rows.

In this dataset, `subgroup_cat == "None"` marks the overall state row.

```{r}
state_overall <- prams |>
  filter(
    subgroup_cat == "None",
    !is.na(alcohol_use_within_3_months_birth),
    !is.na(binge_drinking_within_3_months_birth)
  ) |>
  transmute(
    location,
    alcohol_use = alcohol_use_within_3_months_birth,
    binge_drinking = binge_drinking_within_3_months_birth,
    difference = alcohol_use - binge_drinking
  ) |>
  arrange(difference)

state_overall
```

# Two Bars Per State

First, compare the two estimates using paired bars.

```{r}
paired_estimates <- state_overall |>
  select(location, alcohol_use, binge_drinking) |>
  tidyr::pivot_longer(
    cols = c(alcohol_use, binge_drinking),
    names_to = "measure",
    values_to = "estimate"
  ) |>
  mutate(
    measure = recode(
      measure,
      alcohol_use = "Alcohol use",
      binge_drinking = "Binge drinking"
    ),
    location = factor(location, levels = state_overall$location)
  )

p1 <- ggplot(paired_estimates, aes(x = location, y = estimate, fill = measure)) +
  geom_col(position = "dodge")

p1
```

```{r}
p2 <- p1 +
  coord_flip() +
  labs(
    title = "Two Related PRAMS Estimates By State",
    x = NULL,
    y = "Estimate (%)",
    fill = NULL
  ) +
  theme_minimal()

p2
```

# Plot The Difference Directly

If the question is about the gap, calculate and plot the gap.

```{r}
difference_data <- state_overall |>
  mutate(location = reorder(location, difference))

p1 <- ggplot(difference_data, aes(x = location, y = difference)) +
  geom_hline(yintercept = 0, color = "gray60") +
  geom_col(fill = "steelblue")

p1
```

```{r}
p2 <- p1 +
  coord_flip() +
  labs(
    title = "Gap Between Alcohol Use And Binge Drinking Estimates",
    subtitle = "PRAMS state overall rows; values are percentage-point differences",
    x = NULL,
    y = "Alcohol use minus binge drinking"
  ) +
  theme_minimal()

p2
```

Interpretation: the difference plot makes the size of the gap easier to compare across states than the paired bar plot. The values are state-level PRAMS percentages, not individual-level outcomes.

# Takeaways

- If the question is about a difference, plot the difference.
- Difference plots reduce visual clutter when paired comparisons are the focus.
- Always label whether differences are percentage points, raw units, ratios, or percent changes.
