---
title: "Effective And Honest Scales"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# Goal

This notebook shows how scale choices affect interpretation.

# Analytic Questions

We will ask:

- How can axis limits exaggerate or reduce apparent differences?
- When should a scale start at zero?
- How do we show a focused comparison without misleading the reader?

# Load Data

```{r}
flu_path <- "data/real/flu_vaccination_age_time_clean.csv"

flu <- read_csv(flu_path, show_col_types = FALSE)
```

# Prepare Data

```{r}
us_may <- flu |>
  filter(
    geography == "United States",
    month == 5,
    age_group %in% c("18-49 Years", "50-64 Years", ">=65 Years"),
    !is.na(estimate)
  )
```

# A Focused Axis

`coord_cartesian()` zooms the displayed y-axis without removing rows from the data.

```{r}
p1 <- ggplot(us_may, aes(x = season_start_year, y = estimate, color = age_group)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  coord_cartesian(ylim = c(25, 80))

p2 <- p1 +
  labs(
    title = "Focused Axis For Comparing Seasonal Changes",
    subtitle = "United States, May (month 5)",
    x = "Season start year",
    y = "Vaccination estimate (%)",
    color = "Age group"
  ) +
  theme_minimal()

p2
```

The focused axis makes year-to-year changes easier to see.

# A Full Percent Scale

```{r}
p1 <- ggplot(us_may, aes(x = season_start_year, y = estimate, color = age_group)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  coord_cartesian(ylim = c(0, 100))

p2 <- p1 +
  labs(
    title = "Full Percent Scale Shows The Size Of The Estimates",
    subtitle = "United States, May (month 5)",
    x = "Season start year",
    y = "Vaccination estimate (%)",
    color = "Age group"
  ) +
  theme_minimal()

p2
```

The full scale makes the absolute level clearer, but compresses small changes.

# Bars Usually Need Zero

```{r}
latest <- us_may |>
  filter(season_start_year == max(season_start_year)) |>
  mutate(age_group = reorder(age_group, estimate))

p1 <- ggplot(latest, aes(x = age_group, y = estimate)) +
  geom_col(fill = "steelblue") +
  coord_flip(ylim = c(0, 100))

p2 <- p1 +
  labs(
    title = "Bars Encode Length From A Baseline",
    subtitle = "Latest available season, United States, May (month 5)",
    x = "Age group",
    y = "Vaccination estimate (%)"
  ) +
  theme_minimal()

p2
```

# Takeaways

- For line plots, a narrowed y-axis can help compare change, but the title, subtitle, or caption should make the axis range obvious.
- Percent scales often need context for the full 0-100 range.
- Bars encode length from a baseline, so zero usually matters.
- Scale choices should match the comparison the figure asks the reader to make.
