---
title: "Color, Accessibility, And Highlighting"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# Goal

This notebook shows how color can encode groups, highlight a comparison, or create clutter.

# Analytic Questions

We will ask:

- When should color encode data?
- When should color highlight one category?
- How can filtering or highlighting reduce legend work?

# Load Data

```{r}
census_path <- "data/real/census_state_race_ethnicity_long.csv"

census <- read_csv(census_path, show_col_types = FALSE)
```

# Prepare Data

```{r}
selected_states <- census |>
  filter(state %in% c("CA", "FL", "MA", "MD", "TX", "WA")) |>
  mutate(share_percent = share * 100)
```

# Color Encodes Categories

```{r}
p1 <- ggplot(
  selected_states,
  aes(x = state_name, y = share_percent, fill = race_ethnicity)
) +
  geom_col()

p2 <- p1 +
  labs(
    title = "Population Composition In Selected States",
    x = NULL,
    y = "Population share (%)",
    fill = "Race/ethnicity"
  ) +
  theme_minimal()

p2
```

This uses color to encode several categories. It works, but the reader has to use the legend constantly.

# Highlight One Category

When the question is about one category, highlight that category and mute the rest.

```{r}
asian_share <- selected_states |>
  mutate(
    highlight = if_else(race_ethnicity == "Asian", "Asian", "Other categories")
  ) |>
  group_by(state_name, highlight) |>
  summarise(
    share_percent = sum(share_percent),
    .groups = "drop"
  )

p1 <- ggplot(asian_share, aes(x = state_name, y = share_percent, fill = highlight)) +
  geom_col(position = "stack")

p2 <- p1 +
  scale_fill_manual(values = c("Asian" = "steelblue", "Other categories" = "gray85")) +
  labs(
    title = "Highlighting One Category Changes The Reading Task",
    subtitle = "Asian population share highlighted within selected states",
    x = NULL,
    y = "Population share (%)",
    fill = NULL
  ) +
  theme_minimal()

p2
```

# Direct Comparison Without Stacking

If the task is comparing one category across states, plot only that category.

```{r}
asian_only <- selected_states |>
  filter(race_ethnicity == "Asian") |>
  mutate(state_name = reorder(state_name, share_percent))

p1 <- ggplot(asian_only, aes(x = state_name, y = share_percent)) +
  geom_col(fill = "steelblue")

p2 <- p1 +
  coord_flip() +
  labs(
    title = "Asian Population Share In Selected States",
    x = NULL,
    y = "Population share (%)"
  ) +
  theme_minimal()

p2
```

# Takeaways

- Use color when it carries information or directs attention.
- Too many colors can make the legend do too much work.
- Highlighting is useful when one category is central.
- Do not rely on color alone; use labels, ordering, contrast, or filtering so the figure still works without color.
- Sometimes the clearest color choice is a single color plus careful filtering.
