---
title: "Saving Visualizations"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# Goal

This notebook shows how to save a `ggplot` with a chosen file type, size, and resolution.

The file-writing code is disabled so the module renders without creating extra files.

# Why Saving Choices Matter

When you export a plot, you are deciding:

- file format
- width and height
- resolution
- background
- whether the output is for a notebook, slides, a paper, or the web

# Build A Plot

```{r}
state_path <- "data/real/state_population_ranks.csv"

states <- read_csv(state_path, show_col_types = FALSE)

top_states <- states |>
  slice_max(population, n = 10) |>
  mutate(state_name = reorder(state_name, population))

p1 <- ggplot(top_states, aes(x = state_name, y = population)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Largest States In The Population Rank Dataset",
    x = NULL,
    y = "Population"
  ) +
  theme_minimal()

p1
```

# Save With `ggsave()`

The main `ggplot2` function for saving plots is `ggsave()`.

These chunks are not evaluated when rendering. To run them manually, first create a `figures/` folder at the project root, or change the filename to an existing project-root folder.

```{r}
#| eval: false
ggsave(
  filename = "figures/top-states.png",
  plot = p1,
  width = 7,
  height = 5,
  dpi = 300
)
```

# Common Formats

Use formats based on where the plot is going.

| Format | Common Use |
|---|---|
| PNG | slides, websites, documents |
| PDF | manuscripts, vector graphics, print workflows |
| SVG | web graphics that need to scale |
| JPEG | photos or image-heavy files; less common for statistical graphics |

# Save Several Versions

```{r}
#| eval: false
ggsave("figures/top-states.png", p1, width = 7, height = 5, dpi = 300)
ggsave("figures/top-states.pdf", p1, width = 7, height = 5)
ggsave("figures/top-states.svg", p1, width = 7, height = 5)
```

# Size Is Part Of Design

A plot that is readable in a notebook can become unreadable on a slide if the final display size makes labels too small.

```{r}
p2 <- p1 +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.text = element_text(size = 11),
    axis.title = element_text(size = 12)
  )

p2
```

# Takeaways

- Export settings are part of the communication workflow.
- Use PNG for most course submissions and slides.
- Use PDF or SVG when vector output matters.
- Choose plot size based on where the figure will be read.
