---
title: "Practice: Color Accessibility And Highlighting"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# Goal

Practice using color to guide attention rather than decorate every category equally.

# Suggested Data

Choose one:

- `data/real/census_state_race_ethnicity_long.csv`
- `data/real/places_county_core_measures_long.csv`
- `data/simulated/state_school_enrollment_groups_long.csv`

Helpful module reference: `modules/09_communication-polish/02_color-accessibility-and-highlighting.qmd`.

# Load And Inspect

```{r}
# Load your dataset here.
```

```{r}
# Inspect categories and possible highlight targets.
```

# Build A First Plot

```{r}
# Build a first plot with default or many colors.
```

# Add Highlighting

Choose one group, place, or category to highlight. Make the rest visually quieter.

```{r}
# Build a highlighted version here.
```

# Interpret

> TODO: What should the reader notice first? Is the plot still understandable without relying only on color?
