---
title: "Practice: Color Accessibility And Highlighting"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# What We Practiced

The module showed how to:

- use color to encode groups or guide attention
- reduce legend burden
- highlight one category while muting context
- avoid relying on color alone

# Task

Pick one dataset and create:

1. a first plot with default or many colors
2. a revised plot that filters, highlights, or simplifies color
3. a short interpretation of what the reader should notice first

Suggested datasets:

- `data/real/census_state_race_ethnicity_long.csv`
- `data/real/places_county_core_measures_long.csv`
- `data/simulated/state_school_enrollment_groups_long.csv`

Helpful module reference: `modules/09_communication-polish/02_color-accessibility-and-highlighting.qmd`.

# Work

```{r}
# Load data, build a first plot, then build a revised highlighted or simplified version.
```

# Interpretation

Write briefly: what the reader should notice first and whether the plot still works without relying only on color.
