---
title: "Practice: Saving Visualizations"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# Goal

Practice creating a final plot object and saving it intentionally.

# Suggested Data

Choose one:

- `data/real/state_population_ranks.csv`
- `data/real/flu_vaccination_age_time_clean.csv`
- `data/simulated/state_enrollment_ranks.csv`

Helpful module reference: `modules/09_communication-polish/03_saving-visualizations.qmd`.

# Load And Inspect

```{r}
# Load your dataset here.
```

```{r}
# Inspect variables you could use in a final plot.
```

# Build A Final Plot Object

Create one polished plot and save it as an object such as `final_plot`.

```{r}
# Build final_plot here.
```

# Save The Plot

Choose a filename, width, height, and units. Save into your own work folder.

```{r}
# Save your plot with ggsave() here.
```

# Reflect

> TODO: Why did you choose this size and file format? Does the saved version remain readable?
