---
title: "Composition And Proportions"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
```

# Goal

This notebook demonstrates how to visualize composition.

# Data

We will continue with the census race/ethnicity data.

# Analytic Questions

- What is the race/ethnicity composition of selected states?
- How do proportions differ from raw counts?
- Why are stacked bars hard to read when they combine several variables?
- How can small multiples make category comparisons easier?

# Load Data

```{r}
census <- readr::read_csv("data/real/census_state_race_ethnicity_long.csv")

census |>
  glimpse()
```

Each row is one race/ethnicity category within one state.

# Prepare The Data

We will pick a smaller number of states so the composition plot is readable.

The original data column is named `share`. We will use the `rename` function to change the name of `share` to `race_ethnicity_proportion`, which is more explicit.

<!-- data-prep-anatomy-start -->
::: {.data-prep-anatomy}
| Data preparation step | Functions in this chunk |
|---|---|
| Keep or remove rows | `filter()` |
| Choose or rename columns | `select()`; `rename()` |
| Create or change columns | `mutate()` |
:::
<!-- data-prep-anatomy-end -->

```{r}
selected_states <- census |>
  filter(state %in% c("CA", "FL", "MD", "MA", "TX", "WA")) |>
  rename(race_ethnicity_proportion = share) |>
  mutate(
    state_name = factor(
      state_name,
      levels = c(
        "California",
        "Texas",
        "Florida",
        "Washington",
        "Maryland",
        "Massachusetts"
      )
    )
  )

selected_states |>
  select(state_name, race_ethnicity, race_ethnicity_population, race_ethnicity_proportion) |>
  head()
```

# Raw Counts

We will start with counts.  We will also assign race_ethnicity to the `fill` argument in the aesthetic mapping function.

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Aesthetic mappings | `x`; `y`; `fill` |
:::
<!-- plot-anatomy-end -->

```{r}
p1 <- ggplot(
  selected_states,
  aes(
    x = state_name,
    y = race_ethnicity_population,
    fill = race_ethnicity
  )
)

p1
```

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Geoms / stats | `geom_bar()` |
:::
<!-- plot-anatomy-end -->

```{r}
p2 <- p1 +
  geom_bar(stat = "identity")

p2
```

Because we have fill, this is a three variable visualization. This is very hard to understand for a number of reasons:

- The fill colors have insufficient contrast.
- Only the bottom pink category (`white`) has a common scale; every other category is offset by it.

We can nevertheless add labels and a theme.

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Scales / coordinates / facets / labels | `scale_y_continuous()`; `labs()`; `theme_minimal()` |
:::
<!-- plot-anatomy-end -->

```{r}
p3 <- p2 +
  scale_y_continuous(
    breaks = seq(0, 40000000, by = 10000000),
    labels = function(x) ifelse(x == 0, "0", paste0(x / 1000000, "M"))
  ) +
  labs(
    title = "Race/Ethnicity Counts In Selected States",
    x = NULL,
    y = "Population count",
    fill = "Category"
  ) +
  theme_minimal()

p3
```

California and Texas have larger bars because they have larger populations.

This makes raw counts a poor choice if the question is composition.

# Proportions

The variable gives each category's proportion within the included race/ethnicity categories for that state.

In this file, the denominator is the sum of the race/ethnicity categories included in the teaching dataset for that state.

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Aesthetic mappings | `x`; `y`; `fill` |
:::
<!-- plot-anatomy-end -->

```{r}
p1 <- ggplot(
  selected_states,
  aes(
    x = state_name,
    y = race_ethnicity_proportion,
    fill = race_ethnicity
  )
)

p1
```

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Geoms / stats | `geom_bar()` |
:::
<!-- plot-anatomy-end -->

```{r}
p2 <- p1 +
  geom_bar(stat = "identity")

p2
```

Format the y-axis as percentages.

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Scales / coordinates / facets / labels | `scale_y_continuous()` |
:::
<!-- plot-anatomy-end -->

```{r}
p3 <- p2 +
  scale_y_continuous(
    labels = function(x) paste0(round(x * 100), "%")
  )

p3
```

Add labels and a theme.

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Scales / coordinates / facets / labels | `labs()`; `theme_minimal()` |
:::
<!-- plot-anatomy-end -->

```{r}
p4 <- p3 +
  labs(
    title = "Race/Ethnicity Composition In Selected States",
    x = NULL,
    y = "Proportion of included race/ethnicity population",
    fill = "Category"
  ) +
  theme_minimal()

p4
```

This plot is trying to show three things at once:

- state
- proportion of population
- race/ethnicity category

It shows the overall composition of each state, but -- as above -- it is hard to compare most individual categories across states because only the bottom segment has a shared baseline.

# Small Multiples

Another option is to give each state its own panel.

This uses more space, but it matches the denominator better: within each state, the race/ethnicity proportions add to 100%.

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Aesthetic mappings | `x`; `y` |
| Geoms / stats | `geom_bar()` |
| Scales / coordinates / facets / labels | `facet_wrap()`; `scale_x_continuous()`; `labs()`; `theme_minimal()`; `theme()` |
:::
<!-- plot-anatomy-end -->

```{r}
ggplot(
  selected_states,
  aes(x = race_ethnicity_proportion, y = race_ethnicity)
) +
  geom_bar(stat = "identity") +
  facet_wrap(~ state_name, ncol = 2) +
  scale_x_continuous(
    labels = function(x) paste0(round(x * 100), "%")
  ) +
  labs(
    title = "Race/Ethnicity Composition Within Selected States",
    x = "Proportion of included race/ethnicity population",
    y = NULL
  ) +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"))
```

Small multiples are often a good compromise when a single stacked bar chart is doing too much. Here, each panel keeps one state together instead of scattering one state's parts across many panels. It's easier to comprehend.

However, we notice that our race/ethnicity labels are now ordered in reverse alphabetical order, which is confusing. If we relevel by the overall proportion, it is easier to grasp -- especially when a state breaks the trend.

We can also order the state panels by total population. Adding the population to the panel title gives useful context without overloading the chart itself.

<!-- data-prep-anatomy-start -->
::: {.data-prep-anatomy}
| Data preparation step | Functions in this chunk |
|---|---|
| Keep or remove rows | `distinct()` |
| Create or change columns | `mutate()` |
| Count or summarize | `group_by()`; `summarize()` |
| Sort rows | `arrange()`; `desc()` |
| Combine data frames | `left_join()` |
:::
<!-- data-prep-anatomy-end -->

```{r}
race_ethnicity_order <- selected_states |>
  group_by(race_ethnicity) |>
  summarize(
    overall_population = sum(race_ethnicity_population),
    .groups = "drop"
  ) |>
  arrange(overall_population) |>
  pull(race_ethnicity)

state_labels <- selected_states |>
  distinct(state_name, total_population) |>
  arrange(desc(total_population)) |>
  mutate(
    state_label = paste0(state_name, " (", round(total_population / 1000000, 1), "M people)")
  )

selected_states_ordered <- selected_states |>
  left_join(state_labels, by = c("state_name", "total_population")) |>
  mutate(
    race_ethnicity = factor(race_ethnicity, levels = race_ethnicity_order),
    state_label = factor(state_label, levels = state_labels$state_label)
  )

ggplot(
  selected_states_ordered,
  aes(x = race_ethnicity_proportion, y = race_ethnicity)
) +
  geom_bar(stat = "identity") +
  facet_wrap(~ state_label, ncol = 2) +
  scale_x_continuous(
    labels = function(x) paste0(round(x * 100), "%")
  ) +
  labs(
    title = "Race/Ethnicity Composition Within Selected States",
    x = "Proportion of included race/ethnicity population",
    y = NULL
  ) +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"))
```

In the version above, the race/ethnicity labels only appear in the left column. That is not wrong, but it may be easier to read if each panel has its own y-axis labels.

```{r}
ggplot(
  selected_states_ordered,
  aes(x = race_ethnicity_proportion, y = race_ethnicity)
) +
  geom_bar(stat = "identity") +
  facet_wrap(~ state_label, ncol = 2, axes = "all_y") +
  scale_x_continuous(
    labels = function(x) paste0(round(x * 100), "%")
  ) +
  labs(
    title = "Race/Ethnicity Composition Within Selected States",
    x = "Proportion of included race/ethnicity population",
    y = NULL
  ) +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"))
```

Nope -- duplicating labels is worse.

# Make One Comparison Easier

Stacked bars show overall composition, but they make it hard to compare categories that do not share a common baseline.

If the question is about one category, filter to that category.

<!-- data-prep-anatomy-start -->
::: {.data-prep-anatomy}
| Data preparation step | Functions in this chunk |
|---|---|
| Keep or remove rows | `filter()` |
| Choose or rename columns | `select()`; `rename()` |
| Create or change columns | `mutate()` |
:::
<!-- data-prep-anatomy-end -->

```{r}
hispanic_proportion <- census |>
  rename(race_ethnicity_proportion = share) |>
  filter(race_ethnicity == "Hispanic or Latino") |>
  mutate(state_name = reorder(state_name, race_ethnicity_proportion))

hispanic_proportion |>
  select(state_name, race_ethnicity_proportion) |>
  head()
```

```{r}
p1 <- ggplot(
  hispanic_proportion,
  aes(x = state_name, y = race_ethnicity_proportion)
)

p1
```

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Geoms / stats | `geom_bar()` |
| `geom_bar()` properties | `fill` |
:::
<!-- plot-anatomy-end -->

```{r}
p2 <- p1 +
  geom_bar(stat = "identity", fill = "steelblue")

p2
```

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Scales / coordinates / facets / labels | `coord_flip()`; `scale_y_continuous()`; `labs()`; `theme_minimal()` |
:::
<!-- plot-anatomy-end -->

```{r}
p3 <- p2 +
  coord_flip() +
  scale_y_continuous(
    labels = function(x) paste0(round(x * 100), "%")
  ) +
  labs(
    title = "Hispanic Or Latino Proportion By State",
    x = NULL,
    y = "Proportion"
  ) +
  theme_minimal()

p3
```

One option is to keep the axis labels and add direct labels to the bars. This is a matter of taste, but with many states, the direct labels can make the chart easier to scan.

Question: do the annotations help?

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Aesthetic mappings | `label` |
| Geoms / stats | `geom_text()` |
| `geom_text()` properties | `size` |
:::
<!-- plot-anatomy-end -->

```{r}
p4 <- p2 +
  geom_text(
    aes(label = paste0(round(race_ethnicity_proportion * 100), "%")),
    hjust = -0.2,
    size = 3
  ) +
  coord_flip() +
  scale_y_continuous(
    labels = function(x) paste0(round(x * 100), "%"),
    limits = c(0, 0.55)
  ) +
  labs(
    title = "Hispanic Or Latino Proportion By State",
    x = NULL,
    y = "Proportion"
  ) +
  theme_minimal()

p4
```

The state names are still crowded. Often the fix is not more code inside the plot. It is simply giving the figure more vertical space.

```{r}
#| fig-height: 10
p4
```

We can also convert the bars to a lollipop plot. This keeps the ranking and values, but reduces the visual weight of the bars.

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Aesthetic mappings | `x`; `y`; `xend`; `yend`; `label` |
| Geoms / stats | `geom_segment()`; `geom_point()`; `geom_text()` |
| `geom_segment()` properties | `color`; `linewidth` |
| `geom_point()` properties | `color`; `size` |
| Scales / coordinates / facets / labels | `scale_x_continuous()`; `labs()`; `theme_minimal()`; `theme()` |
:::
<!-- plot-anatomy-end -->

```{r}
#| fig-height: 10
p5 <- ggplot(
  hispanic_proportion,
  aes(x = race_ethnicity_proportion, y = state_name)
) +
  geom_segment(
    aes(x = 0, xend = race_ethnicity_proportion, yend = state_name),
    color = "gray70",
    linewidth = 0.8
  ) +
  geom_point(color = "steelblue", size = 2.5) +
  geom_text(
    aes(label = paste0(round(race_ethnicity_proportion * 100), "%")),
    nudge_x = 0.015,
    hjust = 0,
    size = 3
  ) +
  scale_x_continuous(
    labels = function(x) paste0(round(x * 100), "%"),
    limits = c(0, 0.55)
  ) +
  labs(
    title = "Hispanic Or Latino Proportion By State",
    x = "Proportion",
    y = NULL
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank()
  )

p5
```

One more scale choice: should the x-axis run all the way to 100%? That uses the full possible range for a proportion, but it also compresses the differences among states.

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Scales / coordinates / facets / labels | `scale_x_continuous()` |
:::
<!-- plot-anatomy-end -->

```{r}
#| fig-height: 10
p5 +
  scale_x_continuous(
    labels = function(x) paste0(round(x * 100), "%"),
    limits = c(0, 1),
    breaks = seq(0, 1, by = 0.25)
  )
```

Filtering to one category makes the cross-state comparison easier than reading that category out of every stacked bar.  This is indicative of a more general principle: less is more.

# Takeaways

- Raw counts and proportions answer different questions.
- Stacked bars are useful for showing a whole, but weak for comparing many individual segments.
- Small multiples can make those individual category comparisons easier.
- When your research question focused on one category, plot that category directly.  Less is often more.
