---
title: "Practice: Grouped Summaries"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# Goal

Practice calculating and plotting one summary for each group.

# Suggested Data

Choose one:

- `data/real/nhanes_2021_2023_analysis.csv`
- `data/real/places_county_core_measures_long.csv`
- `data/simulated/neighborhood_health_survey_analysis.csv`

Helpful module reference: `modules/04_group-comparison/01_grouped-summaries.qmd`.

# Load And Inspect

```{r}
# Load your dataset here.
```

```{r}
# Inspect groups and numeric variables.
```

# Summarize By Group

Choose a grouping variable and a numeric variable. Calculate a group-level summary.

```{r}
# Create one row per group with summaries such as n, mean, median, or standard error.
```

# Plot

Try a dot plot or bar chart. Add uncertainty intervals if you calculate them.

```{r}
# Build your grouped summary plot here.
```

# Interpret

> TODO: Which groups are higher or lower? Is the summary appropriate for the variable and question?
