---
title: "Practice: Grouped Summaries"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# What We Practiced

The module showed how to:

- choose a statistic to represent each group
- calculate group summaries with `group_by()` and `summarize()`
- compare bars with dot-and-interval plots
- check summaries against the underlying distributions

# Task

Pick one dataset and create:

1. a grouped summary table
2. one grouped summary plot
3. a short interpretation explaining why the statistic fits the question

Suggested datasets:

- `data/real/nhanes_2021_2023_analysis.csv`
- `data/real/places_county_core_measures_long.csv`
- `data/simulated/neighborhood_health_survey_analysis.csv`

Helpful module reference: `modules/04_group-comparison/01_grouped-summaries.qmd`.

# Work

```{r}
# Load data, create a grouped summary table, and build one grouped summary plot.
```

# Interpretation

Write briefly: which groups are higher or lower, what statistic was used, and whether the summary is appropriate for the variable.
