---
title: "Practice: Difference Plots"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# Goal

Practice plotting differences rather than only plotting two values side by side.

# Suggested Data

Choose one:

- `data/real/prams_2011_selected.csv`
- `data/real/davis_height_weight.csv`
- `data/simulated/new_parent_wellness_survey.csv`

Helpful module reference: `modules/04_group-comparison/02_difference-plots.qmd`.

# Load And Inspect

```{r}
# Load your dataset here.
```

```{r}
# Inspect variables that can be compared or subtracted.
```

# Define A Difference

Choose two related measures or two groups to compare. Calculate a difference where positive and negative values have clear meanings.

```{r}
# Create your difference variable or difference table here.
```

# Plot

Make a plot centered around the difference.

```{r}
# Build your difference plot here.
```

# Interpret

> TODO: What does zero mean? What does a positive value mean? What does a negative value mean?
