---
title: "Flow And Attrition"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
```

# Goal

This notebook shows how to visualize movement through stages or narrowing through filters.

# Analytic Questions

- How many NHANES respondents remain after common analytic restrictions?
- Where do observations leave the analysis dataset?
- How can a simple attrition plot communicate workflow decisions?

# Load Data

```{r}
nhanes <- readr::read_csv("data/real/nhanes_2021_2023_analysis.csv")
```

# Build Attrition Counts

Each row below is cumulative: every stage keeps the previous restrictions and adds one more.

<!-- data-prep-anatomy-start -->
::: {.data-prep-anatomy}
| Data preparation step | Functions in this chunk |
|---|---|
| Keep or remove rows | `filter()` |
| Create or change columns | `mutate()` |
:::
<!-- data-prep-anatomy-end -->

```{r}
attrition <- tibble::tibble(
  stage = c(
    "All rows",
    "Adults age 18+",
    "Has systolic BP",
    "Has diet summary",
    "Has income-to-poverty"
  ),
  n = c(
    nrow(nhanes),
    nhanes |> filter(age_years >= 18) |> nrow(),
    nhanes |> filter(age_years >= 18, !is.na(systolic_mean)) |> nrow(),
    nhanes |> filter(age_years >= 18, !is.na(systolic_mean), !is.na(mean_daily_energy_kcal)) |> nrow(),
    nhanes |> filter(age_years >= 18, !is.na(systolic_mean), !is.na(mean_daily_energy_kcal), !is.na(family_income_to_poverty)) |> nrow()
  )
) |>
  mutate(
    stage = factor(stage, levels = stage),
    retained_percent = n / first(n) * 100
  )

attrition
```

# Attrition Bars

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Aesthetic mappings | `x`; `y` |
| Geoms / stats | `geom_col()` |
| `geom_col()` properties | `fill` |
:::
<!-- plot-anatomy-end -->

```{r}
p1 <- ggplot(attrition, aes(x = stage, y = n)) +
  geom_col(fill = "steelblue")

p1
```

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Scales / coordinates / facets / labels | `coord_flip()`; `labs()`; `theme_minimal()` |
:::
<!-- plot-anatomy-end -->

```{r}
p2 <- p1 +
  coord_flip() +
  labs(
    title = "Rows Retained Through An Analysis Workflow",
    x = NULL,
    y = "Number of rows"
  ) +
  theme_minimal()

p2
```

# Add Retention Labels

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Aesthetic mappings | `x`; `y`; `label` |
| Geoms / stats | `geom_col()`; `geom_text()` |
| `geom_col()` properties | `fill` |
| `geom_text()` properties | `size` |
| Scales / coordinates / facets / labels | `coord_flip()`; `labs()`; `theme_minimal()`; `theme()` |
:::
<!-- plot-anatomy-end -->

```{r}
p1 <- ggplot(attrition, aes(x = stage, y = n)) +
  geom_col(fill = "gray75") +
  geom_text(
    aes(label = paste0(round(retained_percent), "% retained")),
    hjust = -0.05,
    size = 3.5
  )

p2 <- p1 +
  coord_flip() +
  labs(
    title = "Analytic Restrictions Change The Available Sample",
    x = NULL,
    y = "Number of rows"
  ) +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank())

p2
```

# Stage-To-Stage Loss

<!-- data-prep-anatomy-start -->
::: {.data-prep-anatomy}
| Data preparation step | Functions in this chunk |
|---|---|
| Keep or remove rows | `filter()` |
| Create or change columns | `mutate()` |
:::
<!-- data-prep-anatomy-end -->

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Aesthetic mappings | `x`; `y` |
| Geoms / stats | `geom_col()` |
| Scales / coordinates / facets / labels | `coord_flip()`; `labs()`; `theme_minimal()` |
:::
<!-- plot-anatomy-end -->

```{r}
stage_loss <- attrition |>
  mutate(
    previous_n = lag(n),
    lost = previous_n - n
  ) |>
  filter(!is.na(lost))

p1 <- ggplot(stage_loss, aes(x = stage, y = lost)) +
  geom_col(fill = "darkorange3")

p2 <- p1 +
  coord_flip() +
  labs(
    title = "Rows Removed When Each New Restriction Is Added",
    x = NULL,
    y = "Rows lost from previous step"
  ) +
  theme_minimal()

p2
```

# Takeaways

- Flow is about movement through stages, not only Sankey diagrams.
- Attrition plots make analytic restrictions visible.
- Counts and percentages answer different questions and often belong together.
- Simple static plots are often enough for analysis workflow communication.
