---
title: "Scatterplots, Smoothing, And Groups"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# Goal

This notebook shows how to move from a basic scatterplot to grouped and smoothed association plots.

It adapts the clinical-course association examples while using a small course dataset.

# Analytic Questions

We will ask:

- How closely do self-reported and measured weight agree?
- Does reporting error differ by measured weight?
- How does grouping change what we notice?

# Load Data

```{r}
davis_path <- "data/real/davis_height_weight.csv"

davis <- read_csv(davis_path, show_col_types = FALSE)
```

# Prepare Data

```{r}
weight_data <- davis |>
  filter(
    !missing_reported_weight,
    !is.na(weight_kg),
    !is.na(reported_weight_kg),
    !likely_swapped_measured_values
  )

weight_data |>
  select(participant_id, sex, weight_kg, reported_weight_kg, weight_reporting_error_kg) |>
  head(10)
```

# Basic Scatterplot

```{r}
p1 <- ggplot(weight_data, aes(x = weight_kg, y = reported_weight_kg)) +
  geom_point()

p1
```

Add a reference line for perfect agreement.

```{r}
p2 <- p1 +
  geom_abline(slope = 1, intercept = 0, color = "gray50") +
  labs(
    title = "Reported And Measured Weight Are Closely Related",
    subtitle = "The diagonal line shows perfect agreement",
    x = "Measured weight (kg)",
    y = "Reported weight (kg)"
  ) +
  theme_minimal()

p2
```

# Add A Smoother

```{r}
p1 <- ggplot(weight_data, aes(x = weight_kg, y = reported_weight_kg)) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, color = "gray60") +
  geom_smooth(se = FALSE, color = "steelblue")

p2 <- p1 +
  labs(
    title = "A Smoother Summarizes The Pattern",
    x = "Measured weight (kg)",
    y = "Reported weight (kg)"
  ) +
  theme_minimal()

p2
```

The smoother is a summary. The points still matter because they show spread around the summary.

# Reporting Error

Sometimes the more useful association uses a derived variable.

```{r}
p1 <- ggplot(weight_data, aes(x = weight_kg, y = weight_reporting_error_kg)) +
  geom_hline(yintercept = 0, color = "gray60") +
  geom_point(alpha = 0.6) +
  geom_smooth(se = FALSE, color = "steelblue")

p2 <- p1 +
  labs(
    title = "Weight Reporting Error By Measured Weight",
    subtitle = "Negative values mean reported weight was lower than measured weight",
    x = "Measured weight (kg)",
    y = "Reported minus measured weight (kg)"
  ) +
  theme_minimal()

p2
```

# Add Grouping

```{r}
p1 <- ggplot(
  weight_data,
  aes(x = weight_kg, y = weight_reporting_error_kg, color = sex)
) +
  geom_hline(yintercept = 0, color = "gray70") +
  geom_point(alpha = 0.7) +
  geom_smooth(se = FALSE)

p2 <- p1 +
  labs(
    title = "Reporting Error Can Be Compared Across Groups",
    x = "Measured weight (kg)",
    y = "Reported minus measured weight (kg)",
    color = "Sex"
  ) +
  theme_minimal()

p2
```

Interpretation: grouping helps check whether the association looks similar across groups. The plot is exploratory and should not be read as a causal explanation for reporting error.

# Takeaways

- Scatterplots show association and spread at the same time.
- Reference lines can encode a meaningful benchmark.
- Smoothers summarize patterns but should not replace the points.
- Derived variables can make the analytic question clearer.
