---
title: "Model Output Plots"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# Goal

This notebook shows how to visualize simple model output without turning the notebook into a modeling course.

# Analytic Questions

We will ask:

- How is age associated with systolic blood pressure?
- How can a fitted line be shown with the observed data?
- How can predicted values help communicate a model result?

# Load Data

```{r}
nhanes_path <- "data/real/nhanes_2021_2023_analysis.csv"

nhanes <- read_csv(nhanes_path, show_col_types = FALSE)
```

# Prepare Data

```{r}
adult_bp <- nhanes |>
  filter(
    age_years >= 18,
    !is.na(age_years),
    !is.na(systolic_mean),
    !is.na(gender)
  )

adult_bp |>
  count(gender)
```

# Data First

```{r}
p1 <- ggplot(adult_bp, aes(x = age_years, y = systolic_mean)) +
  geom_point(alpha = 0.15)

p1
```

```{r}
p2 <- p1 +
  geom_smooth(method = "lm", se = TRUE, color = "steelblue") +
  labs(
    title = "Age And Systolic Blood Pressure",
    subtitle = "Points show observations; the line summarizes a linear fit",
    x = "Age (years)",
    y = "Mean systolic blood pressure"
  ) +
  theme_minimal()

p2
```

# Fit A Simple Model

We are not interpreting every coefficient here. The table is shown only to motivate why prediction plots can be easier to read.

```{r}
bp_model <- lm(systolic_mean ~ age_years + gender, data = adult_bp)

summary(bp_model)
```

This model assumes the age slope is the same for each gender group. If we wanted different slopes by gender, we would need an interaction: `age_years * gender`.

# Plot Predictions

Predicted values can be easier to read than a coefficient table.

```{r}
prediction_grid <- expand.grid(
  age_years = seq(20, 80, by = 5),
  gender = sort(unique(adult_bp$gender))
)

predictions <- predict(
  bp_model,
  newdata = prediction_grid,
  interval = "confidence"
)

prediction_data <- bind_cols(
  prediction_grid,
  as.data.frame(predictions)
)

head(prediction_data)
```

```{r}
p1 <- ggplot(prediction_data, aes(x = age_years, y = fit, color = gender, fill = gender)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.15, color = NA) +
  geom_line(linewidth = 1)

p2 <- p1 +
  labs(
    title = "Predicted Systolic Blood Pressure By Age And Gender",
    subtitle = "Predictions come from a simple linear model fit to adult NHANES observations",
    x = "Age (years)",
    y = "Predicted mean systolic blood pressure",
    color = "Gender",
    fill = "Gender"
  ) +
  theme_minimal()

p2
```

# Put Observed Data In The Background

```{r}
p1 <- ggplot(adult_bp, aes(x = age_years, y = systolic_mean)) +
  geom_point(alpha = 0.05, color = "gray40") +
  geom_line(
    data = prediction_data,
    aes(x = age_years, y = fit, color = gender),
    linewidth = 1
  )

p2 <- p1 +
  labs(
    title = "Model Summaries Should Stay Connected To The Data",
    x = "Age (years)",
    y = "Mean systolic blood pressure",
    color = "Gender"
  ) +
  theme_minimal()

p2
```

# Takeaways

- Plot the data before plotting the model.
- Model plots are summaries of assumptions and fitted patterns.
- Prediction plots can be easier to interpret than coefficient tables.
- Keep model summaries connected to the data they summarize.
