---
title: "Data Preparation Before Plotting"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
if (!requireNamespace("knitr", quietly = TRUE)) {
  install.packages("knitr")
}

library(dplyr)
library(ggplot2)
```

# Goal

This notebook shows a basic workflow for inspecting and preparing data before plotting.

We will save a few intermediate data frames so it is clear what changed at each step.

# Analytic Questions

We will use the NHANES file from the data repository to practice the following `dplyr` actions:

- inspect
- filter
- recode
- count
- summarize
- arrange

# Load Data

Keep the first object close to the file as read from disk.

```{r}
raw_data <- readr::read_csv("data/real/nhanes_2021_2023_analysis.csv")
```

# Inspect The Data

Start by looking at the structure and a few rows.

```{r}
glimpse(raw_data)
```

Or you can use the pipe format (`|>` *pipes* the value and feeds it to the first argument of the next function in the pipe)

```{r}
raw_data |>
  glimpse()
```

Head shows the first handful of rows, which is nice for giving readers of your notebook a sense of the data.

```{r}
head(raw_data)
```

**Warning:** Be careful about using `glimpse()` and `head()` in notebooks with PHI that you will share.

See the names of columns in the data frame:

```{r}
names(raw_data)
```

Each row is one participant in the NHANES teaching file.

**Note:** NHANES is a complex survey. Do not treat these teaching examples as real national estimates.

# Select Variables

Use `select()` to keep the variables needed for the current question.

```{r}
bp_data <- raw_data |>
  select(
    respondent_id,
    age_years,
    age_group,
    gender,
    race_ethnicity,
    systolic_mean,
    diastolic_mean,
    pulse_mean
  )

bp_data |>
  head()
```

# Filter Rows

Use `filter()` to keep the observations that match the question.

Here we keep adults with a non-missing systolic blood pressure value.

<!-- data-prep-anatomy-start -->
::: {.data-prep-anatomy}
| Data preparation step | Functions in this chunk |
|---|---|
| Inspect data | `glimpse()` |
| Keep or remove rows | `filter()` |
:::
<!-- data-prep-anatomy-end -->

```{r}
adult_bp <- bp_data |>
  filter(
    age_years >= 18,
    !is.na(systolic_mean)
  )

adult_bp |>
  glimpse()
```

# Count Categories

Use `count()` to check categorical variables before plotting or grouping.

<!-- data-prep-anatomy-start -->
::: {.data-prep-anatomy}
| Data preparation step | Functions in this chunk |
|---|---|
| Count or summarize | `count()` |
:::
<!-- data-prep-anatomy-end -->

```{r}
adult_bp |>
  count(age_group, sort = TRUE)
```

This is basically the same as `table()`, except you get back a data frame with nicely named columns, which are easier to work with:

```{r}
table(adult_bp$age_group)
```

Gender:

<!-- data-prep-anatomy-start -->
::: {.data-prep-anatomy}
| Data preparation step | Functions in this chunk |
|---|---|
| Count or summarize | `count()` |
:::
<!-- data-prep-anatomy-end -->

```{r}
adult_bp |>
  count(gender, sort = TRUE)
```

# Create Or Recode Variables

Use `mutate()` for new variables or recoded categories.

<!-- data-prep-anatomy-start -->
::: {.data-prep-anatomy}
| Data preparation step | Functions in this chunk |
|---|---|
| Choose or rename columns | `select()` |
| Create or change columns | `mutate()`; `case_when()` |
:::
<!-- data-prep-anatomy-end -->

```{r}
clean_data <- adult_bp |>
  mutate(
    systolic_group = case_when(
      systolic_mean < 120 ~ "Less than 120",
      systolic_mean < 130 ~ "120 to 129",
      systolic_mean < 140 ~ "130 to 139",
      systolic_mean >= 140 ~ "140 or higher",
      TRUE ~ NA_character_
    ),
    pulse_group = case_when(
      is.na(pulse_mean) ~ "Missing pulse",
      pulse_mean < 60 ~ "Less than 60",
      pulse_mean < 80 ~ "60 to 79",
      pulse_mean < 100 ~ "80 to 99",
      pulse_mean >= 100 ~ "100 or higher"
    )
  )

clean_data |>
  select(respondent_id, age_years, gender, systolic_mean, systolic_group, pulse_mean, pulse_group) |>
  head()
```

# Arrange Rows For Inspection

Use `arrange()` to inspect records at the high or low end of a variable.

<!-- data-prep-anatomy-start -->
::: {.data-prep-anatomy}
| Data preparation step | Functions in this chunk |
|---|---|
| Choose or rename columns | `select()` |
| Sort rows | `arrange()`; `desc()` |
:::
<!-- data-prep-anatomy-end -->

```{r}
clean_data |>
  arrange(desc(systolic_mean)) |>
  select(respondent_id, age_years, gender, systolic_mean, diastolic_mean, pulse_group) |>
  head()
```

Sorting does not change the data. It helps us inspect the rows behind summaries and plots.

# Summarize The Whole Table

Use `summarize()` when you want one row of summary values.

<!-- data-prep-anatomy-start -->
::: {.data-prep-anatomy}
| Data preparation step | Functions in this chunk |
|---|---|
| Count or summarize | `summarize()`; `n()` |
:::
<!-- data-prep-anatomy-end -->

```{r}
clean_data |>
  summarize(
    participants = n(),
    high_systolic_count = sum(systolic_mean >= 130),
    mean_systolic = mean(systolic_mean),
    median_systolic = median(systolic_mean),
    min_systolic = min(systolic_mean),
    max_systolic = max(systolic_mean),
    iqr_systolic = IQR(systolic_mean),
    sd_systolic = sd(systolic_mean)
  )
```

# Summarize By Group

Use `group_by()` before `summarize()` when you want one summary row per group.

<!-- data-prep-anatomy-start -->
::: {.data-prep-anatomy}
| Data preparation step | Functions in this chunk |
|---|---|
| Count or summarize | `group_by()`; `summarize()`; `n()` |
| Sort rows | `arrange()`; `desc()` |
:::
<!-- data-prep-anatomy-end -->

```{r}
gender_summary <- clean_data |>
  group_by(gender) |>
  summarize(
    participants = n(),
    high_systolic_count = sum(systolic_mean >= 130),
    mean_systolic = mean(systolic_mean),
    median_systolic = median(systolic_mean),
    min_systolic = min(systolic_mean),
    max_systolic = max(systolic_mean),
    iqr_systolic = IQR(systolic_mean),
    sd_systolic = sd(systolic_mean),
    .groups = "drop"
  ) |>
  arrange(desc(mean_systolic))

gender_summary
```

# Format A Table With `kable()`

Plain data frames print fine while we work, but rendered notebooks often look better with a formatted table.

`knitr::kable()` takes a data frame and prints it as a simple table in the rendered document.

```{r}
gender_summary |>
  knitr::kable(
    digits = 1,
    caption = "Systolic blood pressure summaries by gender"
  )
```

This does not change `gender_summary`. It only changes how the table is displayed in the notebook output.

# Factors And Category Order

Categorical variables are often stored as character vectors or factors.

For plotting, the important issue is order: what comes first, second, third, and so on.

If we do nothing, `ggplot2` often uses alphabetical order for character categories. This is not always the order we want.

Blood pressure groups have a natural order from lower to higher. We should preserve that order instead of sorting alphabetically or by count.

<!-- data-prep-anatomy-start -->
::: {.data-prep-anatomy}
| Data preparation step | Functions in this chunk |
|---|---|
| Create or change columns | `mutate()` |
:::
<!-- data-prep-anatomy-end -->

```{r}
clean_data <- clean_data |>
  mutate(
    systolic_group = factor(
      systolic_group,
      levels = c(
        "Less than 120",
        "120 to 129",
        "130 to 139",
        "140 or higher"
      )
    )
  )
```

The `levels` argument gives the factor its order.

# Let `ggplot2` Count The Rows

For a simple count plot, `ggplot2` can count rows for us. This is the default behavior of `geom_bar()`.

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Aesthetic mappings | `x` |
| Geoms / stats | `geom_bar()` |
| Scales / coordinates / facets / labels | `coord_flip()`; `labs()`; `theme_minimal()` |
:::
<!-- plot-anatomy-end -->

```{r}
ggplot(clean_data, aes(x = systolic_group)) +
  geom_bar() +
  coord_flip() +
  labs(
    title = "Adult Participants By Systolic Blood Pressure Group",
    x = NULL,
    y = "Number of participants"
  ) +
  theme_minimal()
```

This is convenient when the transformation is simple, but the count table is hidden inside the plot step.

# Transform First, Then Plot With Identity

Instead of asking `ggplot2` to count rows inside the plot, we can do the transformation ourselves and check the table before plotting.

<!-- data-prep-anatomy-start -->
::: {.data-prep-anatomy}
| Data preparation step | Functions in this chunk |
|---|---|
| Count or summarize | `count()` |
:::
<!-- data-prep-anatomy-end -->

```{r}
plot_data <- clean_data |>
  count(systolic_group)

plot_data
```

Now `stat = "identity"` tells `geom_bar()` to use the `n` values already in `plot_data`. As you get more comfortable, you may use this explicit pattern more often.

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Aesthetic mappings | `x`; `y` |
:::
<!-- plot-anatomy-end -->

```{r}
ggplot(plot_data, aes(x = systolic_group, y = n)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(
    title = "Adult Participants By Systolic Blood Pressure Group",
    x = NULL,
    y = "Number of participants"
  ) +
  theme_minimal()
```

# Data Frame Naming

The named data frames in this notebook have specific jobs:

- `raw_data`: the file as read from disk
- `bp_data`: only the variables needed for the blood pressure question
- `adult_bp`: rows filtered to adults with systolic blood pressure data
- `clean_data`: recoded variables added
- `plot_data`: a small table made for one plot

This is enough. Avoid making a new object after every line if the object does not represent a useful stage.

# Summary

- Inspect data before visualizing.
- Keep the raw file import separate from cleaned or filtered data.
- Use `select()`, `filter()`, `mutate()`, `arrange()`, `count()`, `group_by()`, and `summarize()` to prepare data in small steps.
- Use summary functions such as `n()`, `sum()`, `mean()`, `median()`, `min()`, `max()`, `IQR()`, and `sd()` to check the data before plotting.
- Make a small plot-specific table when it clarifies what the plot is showing.
