---
title: "Time Trends And Small Multiples"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# Goal

This notebook shows how to visualize change over time and compare time patterns across groups.

It adapts the clinical-course time-trend examples to the course flu vaccination dataset.

# Analytic Questions

We will ask:

- How do flu vaccination estimates change across seasons?
- How do time patterns differ by age group?
- When are small multiples clearer than many colored lines?

# Load Data

```{r}
flu_path <- "data/real/flu_vaccination_age_time_clean.csv"

flu <- read_csv(flu_path, show_col_types = FALSE)
```

# Prepare Data

Use national estimates from the same month in each season.

```{r}
us_may <- flu |>
  filter(
    geography == "United States",
    month == 5,
    !is.na(estimate)
  )

us_may |>
  count(age_group)
```

# One Group Over Time

```{r}
older_adults <- us_may |>
  filter(age_group == ">=65 Years")

p1 <- ggplot(older_adults, aes(x = season_start_year, y = estimate)) +
  geom_line() +
  geom_point()

p1
```

```{r}
p2 <- p1 +
  labs(
    title = "Flu Vaccination Estimates Over Time",
    subtitle = "United States, month 5, ages 65 and older",
    x = "Season start year",
    y = "Vaccination estimate (%)"
  ) +
  theme_minimal()

p2
```

# Several Groups With Color

```{r}
selected_age_groups <- us_may |>
  filter(age_group %in% c("6 Months - 4 Years", "18-49 Years", "50-64 Years", ">=65 Years"))

p1 <- ggplot(
  selected_age_groups,
  aes(x = season_start_year, y = estimate, color = age_group)
) +
  geom_line(linewidth = 1) +
  geom_point(size = 2)

p2 <- p1 +
  labs(
    title = "Flu Vaccination Time Trends Differ By Age Group",
    subtitle = "United States, month 5",
    x = "Season start year",
    y = "Vaccination estimate (%)",
    color = "Age group"
  ) +
  theme_minimal()

p2
```

# Small Multiples

When many groups compete for attention, facets can make the patterns easier to scan.

```{r}
p1 <- ggplot(selected_age_groups, aes(x = season_start_year, y = estimate)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 1.8) +
  facet_wrap(vars(age_group))

p2 <- p1 +
  labs(
    title = "Small Multiples Separate The Age-Group Patterns",
    subtitle = "United States, month 5",
    x = "Season start year",
    y = "Vaccination estimate (%)"
  ) +
  theme_minimal()

p2
```

# Takeaways

- Line plots work well when the x-axis has a meaningful order.
- Use the same month or comparable time point when comparing seasons.
- Color works for a few groups; facets often work better for many groups.
- Interpret time patterns with attention to changing groups, definitions, and uncertainty.
