---
title: "Before/After And Trajectories"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# Goal

This notebook shows how to visualize change around a reference event or time point.

# Analytic Questions

We will ask:

- How does headache reporting change across diary days?
- How can we mark a reference day?
- How do group-level trajectories differ by medication group?

# Load Data

```{r}
diary_path <- "data/real/migraine_headache_diary.csv"

diary <- read_csv(diary_path, show_col_types = FALSE)
```

# Prepare Data

The source data use day 0 as a study reference day. Negative days are before that reference day; positive days are after it.

```{r}
daily_summary <- diary |>
  filter(!is.na(day), !is.na(medication), !is.na(headache_reported)) |>
  group_by(day, medication) |>
  summarize(
    n = n(),
    headache_percent = mean(headache_reported) * 100,
    .groups = "drop"
  )

daily_summary |>
  head(10)
```

Check `n` before interpreting bumps or dips; small daily counts can make a percent look noisy.

# Overall Trajectory

```{r}
overall_daily <- diary |>
  filter(!is.na(day), !is.na(headache_reported)) |>
  group_by(day) |>
  summarize(
    n = n(),
    headache_percent = mean(headache_reported) * 100,
    .groups = "drop"
  )

p1 <- ggplot(overall_daily, aes(x = day, y = headache_percent)) +
  geom_line() +
  geom_point()

p1
```

```{r}
p2 <- p1 +
  geom_vline(xintercept = 0, color = "gray50", linetype = "dashed") +
  labs(
    title = "Headache Reporting Across Diary Days",
    subtitle = "Dashed line marks the study reference day, day 0",
    x = "Diary day",
    y = "Headache reported (%)"
  ) +
  theme_minimal()

p2
```

# Compare Medication Groups

```{r}
p1 <- ggplot(daily_summary, aes(x = day, y = headache_percent, color = medication)) +
  geom_vline(xintercept = 0, color = "gray70", linetype = "dashed") +
  geom_line(linewidth = 1) +
  geom_point(size = 1.3)

p2 <- p1 +
  labs(
    title = "Group-Level Headache Reporting By Medication Group",
    x = "Diary day",
    y = "Headache reported (%)",
    color = "Medication group"
  ) +
  theme_minimal()

p2
```

# Small Multiples For Trajectories

```{r}
p1 <- ggplot(daily_summary, aes(x = day, y = headache_percent)) +
  geom_vline(xintercept = 0, color = "gray75", linetype = "dashed") +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 1.3) +
  facet_wrap(vars(medication))

p2 <- p1 +
  labs(
    title = "Facets Make Group-Level Patterns Easier To Compare",
    x = "Diary day",
    y = "Headache reported (%)"
  ) +
  theme_minimal()

p2
```

# Takeaways

- Mark reference points directly on time plots.
- Compare groups on the same time scale when the time variable has shared meaning.
- Facets can make trajectory comparisons easier than many overlapping lines.
- Here, "trajectory" means a group-level time pattern, not one person's path across all days.
