---
title: "Histograms And Density Plots"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
```

# Goal

This notebook shows how to visualize the distribution of one continuous variable.

We focus on center, spread, shape, and unusual values/outliers.

A distribution is the collection of observed positions along a measurement scale. The summaries and plots in this notebook are different ways to visualize the distribution.

# Analytic Questions

We will use a simplified NHANES teaching file to ask:

- What does the distribution of adult systolic blood pressure look like?
- Which summaries describe center and spread?
- How does the bin width change a histogram?
- What does a density plot emphasize differently from a histogram?

# Load Data

```{r}
nhanes <- readr::read_csv("data/real/nhanes_2021_2023_analysis.csv")

nhanes |>
  glimpse()
```

Each row is one participant in the joined NHANES teaching file.

NHANES is a complex survey. In this course, these are exploratory sample visualizations, not national estimates.

# Prepare Adult Blood Pressure Data

Use `filter()` to keep the observations that match the question: adults with a non-missing blood pressure value.

<!-- data-prep-anatomy-start -->
::: {.data-prep-anatomy}
| Data preparation step | Functions in this chunk |
|---|---|
| Keep or remove rows | `filter()` |
| Choose or rename columns | `select()` |
:::
<!-- data-prep-anatomy-end -->

```{r}
adult_bp <- nhanes |>
  filter(
    age_years >= 18,
    !is.na(systolic_mean)
  )

adult_bp |>
  select(respondent_id, age_years, gender, age_group, systolic_mean) |>
  head()
```

# Summarize Before Plotting

Summary statistics are aggregations. They reduce many participant values to a few numbers.

<!-- data-prep-anatomy-start -->
::: {.data-prep-anatomy}
| Data preparation step | Functions in this chunk |
|---|---|
| Count or summarize | `summarize()`; `n()` |
:::
<!-- data-prep-anatomy-end -->

```{r}
adult_bp_summary <- adult_bp |>
  summarize(
    n = n(),
    mean_systolic = mean(systolic_mean),
    median_systolic = median(systolic_mean),
    min_systolic = min(systolic_mean),
    max_systolic = max(systolic_mean),
    iqr_systolic = IQR(systolic_mean),
    sd_systolic = sd(systolic_mean)
  )

adult_bp_summary
```

The mean and median describe center. The range, IQR, and standard deviation describe spread. The plot will show shape and unusual values that these summaries compress.

This is a good EDA habit: summarize to get oriented, then plot to see what the summaries don't show you.

# First Histogram

Start with the data and x-axis mapping.

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Aesthetic mappings | `x` |
:::
<!-- plot-anatomy-end -->

```{r}
p1 <- ggplot(
  adult_bp,
  aes(x = systolic_mean)
)

p1
```

Add a histogram.

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Geoms / stats | `geom_histogram()` |
:::
<!-- plot-anatomy-end -->

```{r}
p2 <- p1 +
  geom_histogram(binwidth = 5)

p2
```

Add labels and a theme.

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Scales / coordinates / facets / labels | `labs()`; `theme_minimal()` |
:::
<!-- plot-anatomy-end -->

```{r}
p3 <- p2 +
  labs(
    title = "Distribution Of Adult Mean Systolic Blood Pressure",
    subtitle = "NHANES 2021-2023 teaching file, adults with BP readings",
    x = "Mean systolic blood pressure",
    y = "Number of participants"
  ) +
  theme_minimal()

p3
```

The distribution is centered a little above 120 mmHg and has a right tail with some higher readings.

# Use Fill To Highlight A Threshold

We can also use color or fill to highlight a practical threshold.

For example, below, we fill only the part of the distribution for people with mean daily energy intake over 4,000 calories.

<!-- data-prep-anatomy-start -->
::: {.data-prep-anatomy}
| Data preparation step | Functions in this chunk |
|---|---|
| Keep or remove rows | `filter()` |
| Create or change columns | `mutate()` |
:::
<!-- data-prep-anatomy-end -->

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Aesthetic mappings | `x`; `fill` |
| Geoms / stats | `geom_histogram()` |
| Scales / coordinates / facets / labels | `scale_fill_manual()`; `labs()`; `theme_minimal()` |
:::
<!-- plot-anatomy-end -->

```{r}
diet_4000 <- nhanes |>
  filter(!is.na(mean_daily_energy_kcal)) |>
  mutate(
    energy_over_4000 = mean_daily_energy_kcal > 4000
  )

ggplot(
  diet_4000,
  aes(x = mean_daily_energy_kcal, fill = energy_over_4000)
) +
  geom_histogram(binwidth = 250) +
  scale_fill_manual(
    values = c("TRUE" = "firebrick", "FALSE" = "gray75"),
    labels = c("FALSE" = "4,000 kcal or less", "TRUE" = "Over 4,000 kcal"),
    name = NULL
  ) +
  labs(
    title = "Mean Daily Energy Intake",
    x = "Mean daily energy intake, kcal",
    y = "Number of participants"
  ) +
  theme_minimal()
```

This does not make 4,000 calories an official outlier rule. It is just a threshold we chose to inspect.

# Bin Width

Bin width changes the level of detail in the histogram.

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Aesthetic mappings | `x` |
:::
<!-- plot-anatomy-end -->

```{r}
p1 <- ggplot(adult_bp, aes(x = systolic_mean)) +
  geom_histogram(binwidth = 2)

p1
```

```{r}
p2 <- ggplot(adult_bp, aes(x = systolic_mean)) +
  geom_histogram(binwidth = 10)

p2
```

The smaller bin width shows more detail. The larger bin width smooths over more of the small variation.

# Density Plot

A density plot shows a smoothed distribution.

The y-axis is density, not number of participants. The total area under the curve is scaled to 1.

Because density plots are smoothed, they can suggest more certainty or structure than the raw data support. Use them as a way to inspect shape, not as proof that every bump is meaningful.

```{r}
p1 <- ggplot(adult_bp, aes(x = systolic_mean))

p1
```

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Geoms / stats | `geom_density()` |
| `geom_density()` properties | `fill`; `alpha` |
:::
<!-- plot-anatomy-end -->

```{r}
p2 <- p1 +
  geom_density(fill = "steelblue", alpha = 0.3)

p2
```

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Scales / coordinates / facets / labels | `labs()`; `theme_minimal()` |
:::
<!-- plot-anatomy-end -->

```{r}
p3 <- p2 +
  labs(
    title = "Density Of Adult Mean Systolic Blood Pressure",
    x = "Mean systolic blood pressure",
    y = "Density"
  ) +
  theme_minimal()

p3
```

# Histogram And Density Together

You can stack layers in `ggplot2`. The same plot can include a histogram and a density curve.

To put them on the same y-axis, tell the histogram to use density instead of counts.

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Aesthetic mappings | `x`; `y` |
| Geoms / stats | `geom_histogram()`; `geom_density()` |
| `geom_histogram()` properties | `fill`; `color` |
| `geom_density()` properties | `color`; `linewidth` |
:::
<!-- plot-anatomy-end -->

```{r}
ggplot(adult_bp, aes(x = systolic_mean)) +
  geom_histogram(
    aes(y = after_stat(density)),
    binwidth = 5,
    fill = "gray85",
    color = "white"
  ) +
  geom_density(color = "steelblue", linewidth = 1) +
  labs(
    title = "Adult Mean Systolic Blood Pressure",
    x = "Mean systolic blood pressure",
    y = "Density"
  ) +
  theme_minimal()
```

The bars show the binned data. The line shows a smoothed version of the same distribution.

# Bimodal Distributions

Sometimes a distribution has two peaks. This can happen when the data combine two groups with different typical values.

Here is a small simulated example with two groups.

```{r}
set.seed(123)

simulated_people <- data.frame(
  sex = rep(c("Female", "Male"), each = 400),
  measurement = c(
    rnorm(400, mean = 64, sd = 3),
    rnorm(400, mean = 70, sd = 3)
  )
)
```

If we plot everyone together, we can spot the bimodal shape.

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Aesthetic mappings | `x` |
| Geoms / stats | `geom_histogram()` |
:::
<!-- plot-anatomy-end -->

```{r}
ggplot(simulated_people, aes(x = measurement)) +
  geom_histogram(binwidth = 1, fill = "gray75", color = "white") +
  labs(
    title = "A Simulated Bimodal Distribution",
    x = "Measurement",
    y = "Number of people"
  ) +
  theme_minimal()
```

Faceting by sex shows why the combined distribution has two peaks.

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Scales / coordinates / facets / labels | `facet_wrap()`; `labs()`; `theme_minimal()` |
:::
<!-- plot-anatomy-end -->

```{r}
ggplot(simulated_people, aes(x = measurement)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
  facet_wrap(~ sex) +
  labs(
    title = "The Two Peaks Come From Two Groups",
    x = "Measurement",
    y = "Number of people"
  ) +
  theme_minimal()
```

When doing EDA, we will sometimes see patterns like this, which then inform us about the structure of the data and the underlying phenomenon of inquiry.

# Takeaways

- `filter()` and `select()` prepare the data for the question.
- `summarize()` aggregates many values into summary statistics.
- Histograms help us visualize the distribution of a continuous variable.
- Bin width affects what patterns are visible.  Too small or too big each can make things harder to see.
- Density plots emphasize smoothed shape rather than counts; use with care.
- Smoothed density can mislead with small samples, bounded variables, or overly detailed interpretations.
- EDA can help us uncover important features of our data and topic of study.
