---
title: "Practice: Saving Visualizations"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
```

# Goal

Practice creating a named plot object and choosing save settings.

The saving chunks are evaluated when the notebook renders. They write image files into `saved_images/` in this module's directory.

# 1. Load Data

```{r}
states <- readr::read_csv("data/real/state_population_ranks.csv")
```

# 2. Build A Plot Object

Name the plot object `final_plot`. This makes it easier to save exactly the plot you want.

```{r}
top_states <- states |>
  slice_min(rank, n = 12) |>
  mutate(state_name = reorder(state_name, population))

final_plot <- ggplot(top_states, aes(x = state_name, y = population)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  scale_y_continuous(labels = scales::label_number(scale_cut = scales::cut_short_scale())) +
  labs(
    title = "Largest States In The Population Rank Dataset",
    x = NULL,
    y = "Population"
  ) +
  theme_minimal()

final_plot
```

# 3. Save A PNG

PNG is a good default for slides, websites, and most course submissions.

```{r}
ggsave(
  filename = "practice/05-communication-and-grab-bag/saved_images/top-states.png",
  plot = final_plot,
  width = 7,
  height = 5,
  dpi = 300,
  bg = "white"
)
```

# 4. Save A Wider Version

Sometimes the same plot needs a different shape. Change the saved dimensions intentionally instead of dragging the plot pane.

```{r}
ggsave(
  filename = "practice/05-communication-and-grab-bag/saved_images/top-states-wide.png",
  plot = final_plot,
  width = 9,
  height = 4.5,
  dpi = 300,
  bg = "white"
)
```

# 5. Save A Vector Version

PDF is useful when the figure needs to stay sharp in print or in a document workflow.

```{r}
ggsave(
  filename = "practice/05-communication-and-grab-bag/saved_images/top-states.pdf",
  plot = final_plot,
  width = 7,
  height = 5
)
```
