---
title: "2. Setup Check"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
```

Use this notebook to confirm that R, the course packages, and Quarto rendering work on your computer.

If this notebook fails because a package is missing, run this in the R Console from the course project root:

```r
source("install-packages.R")
```

After opening this file:

- In RStudio, click `Render` in the editor toolbar above the notebook.
- In Positron, click `Preview` in the editor toolbar above the notebook.

If rendering works, the checks below should run automatically and produce an HTML file with a table and plot.

# Check 1: R Works

This calculation should run during rendering.

```{r}
2 + 2
```

You should see `4` in the output.

# Check 2: Data Works

R includes a small built-in dataset called `mtcars`. We can use it to test data work without downloading anything else.

```{r}
mtcars |>
  head()
```

# Check 3: Summaries Work

This summary should run during rendering.

<!-- data-prep-anatomy-start -->
::: {.data-prep-anatomy}
| Data preparation step | Functions in this chunk |
|---|---|
| Count or summarize | `summarize()`; `n()` |
:::
<!-- data-prep-anatomy-end -->

```{r}
mtcars |>
  summarize(
    rows = n(),
    average_mpg = mean(mpg),
    minimum_mpg = min(mpg),
    maximum_mpg = max(mpg)
  )
```

# Check 4: ggplot2 Works

This plot should run during rendering.

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Aesthetic mappings | `x` |
:::
<!-- plot-anatomy-end -->

```{r}
p1 <- ggplot(mtcars, aes(x = mpg))

p1
```

The blank plot is expected. We have provided data and an x-axis mapping, but no geometric object yet.

Add a histogram.

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Geoms / stats | `geom_histogram()` |
:::
<!-- plot-anatomy-end -->

```{r}
p2 <- p1 +
  geom_histogram(binwidth = 5)

p2
```

Add labels.

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Scales / coordinates / facets / labels | `labs()` |
:::
<!-- plot-anatomy-end -->

```{r}
p3 <- p2 +
  labs(
    title = "Distribution Of Fuel Efficiency",
    subtitle = "Built-in mtcars dataset",
    x = "Miles per gallon",
    y = "Number of cars"
  )

p3
```

Add a theme.

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Scales / coordinates / facets / labels | `theme_minimal()` |
:::
<!-- plot-anatomy-end -->

```{r}
p4 <- p3 +
  theme_minimal()

p4
```

# Check 5: Render Works

After rendering or previewing, confirm that:

- an HTML file was created or previewed
- the table appears
- the plot appears
- there are no error messages
