---
title: "Practice: Histograms And Summaries"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
```

# Learning Tactic

- For in-class practice, we are going to try an approach where we study working code and tweak it.
- This will help provide hands-on experience working with data visualization code.

# 1. Load Data

```{r}
davis <- readr::read_csv("data/real/davis_height_weight.csv")

davis |>
  glimpse()
```

# 2. Prepare One Variable

This keeps participants with a non-missing measured height value.

```{r}
height_data <- davis |>
  filter(!is.na(height_cm))

height_data |>
  select(participant_id, sex, height_cm, weight_kg, measured_bmi) |>
  head()
```

Try adding `filter(sex == "female")` to focus on one group.

# 3. Summarize Center And Spread

These summaries describe center, spread, range, and sample size for measured height.

```{r}
height_data |>
  summarize(
    n = n(),
    mean_height = mean(height_cm),
    median_height = median(height_cm),
    iqr_height = IQR(height_cm),
    sd_height = sd(height_cm),
    min_height = min(height_cm),
    max_height = max(height_cm)
  )
```

Try adding a quantile to this summary, such as `p90_height = quantile(height_cm, 0.90)`.

# 4. First Histogram

This histogram uses 5-centimeter bins.

```{r}
ggplot(height_data, aes(x = height_cm)) +
  geom_histogram(binwidth = 5, fill = "gray75", color = "white") +
  labs(
    title = "Measured Height",
    x = "Measured height, cm",
    y = "Number of participants"
  ) +
  theme_minimal()
```

Try changing `binwidth = 5` to `binwidth = 2`.

# 5. Larger Bins

This version uses the same data with larger bins.

```{r}
ggplot(height_data, aes(x = height_cm)) +
  geom_histogram(binwidth = 10, fill = "gray75", color = "white") +
  labs(
    title = "Measured Height",
    x = "Measured height, cm",
    y = "Number of participants"
  ) +
  theme_minimal()
```

Write briefly: what did the larger bin width make easier or harder to see?

# 6. Density Plot

This version uses a smoothed density curve instead of count bins.

```{r}
ggplot(height_data, aes(x = height_cm)) +
  geom_density(fill = "steelblue", alpha = 0.3) +
  labs(
    title = "Density Of Measured Height",
    x = "Measured height, cm",
    y = "Density"
  ) +
  theme_minimal()
```

Try changing the fill color or the transparency value `alpha`.

# On Your Own

Choose one numeric variable from the Davis data:

- `weight_kg`
- `reported_weight_kg`
- `measured_bmi`

Create one summary table and one histogram. Write 1-2 sentences describing center, spread, shape, and unusual values.
