---
title: "Practice: Boxplots And Unusual Values"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
```

# Goal

Practice using boxplots and simple IQR fences to inspect spread and unusual values.

# 1. Load Data

```{r}
davis <- readr::read_csv("data/real/davis_height_weight.csv")
```

# 2. Prepare Weight Data

This keeps participants with non-missing measured weight.

```{r}
weight_data <- davis |>
  filter(!is.na(weight_kg))

weight_data |>
  select(participant_id, sex, weight_kg, height_cm, measured_bmi) |>
  head()
```

Try adding `filter(sex == "male")` to focus on one group.

# 3. Calculate IQR Fences

These cutoffs use the same basic rule as a default boxplot.

```{r}
weight_cutoffs <- weight_data |>
  summarize(
    q1 = quantile(weight_kg, 0.25),
    q3 = quantile(weight_kg, 0.75),
    iqr = IQR(weight_kg),
    lower_cutoff = q1 - 1.5 * iqr,
    upper_cutoff = q3 + 1.5 * iqr
  )

weight_cutoffs
```

Try changing both `1.5` values to `3` and compare the cutoffs.

# 4. Flag High Values

This flags values above the upper cutoff.

```{r}
weight_flagged <- weight_data |>
  mutate(
    high_weight_flag = weight_kg > weight_cutoffs$upper_cutoff
  )

weight_flagged |>
  count(high_weight_flag)
```

Try changing `>` to `>=`. Does the count change?

# 5. Inspect Flagged Rows

This shows the highest flagged values.

```{r}
weight_flagged |>
  filter(high_weight_flag) |>
  select(participant_id, sex, height_cm, weight_kg, measured_bmi) |>
  arrange(desc(weight_kg)) |>
  head(10)
```

The table shows the rows. This histogram shows where the flagged values sit in the full distribution.

```{r}
ggplot(
  weight_flagged,
  aes(x = weight_kg, fill = high_weight_flag)
) +
  geom_histogram(binwidth = 5, color = "white") +
  scale_fill_manual(
    values = c("FALSE" = "gray75", "TRUE" = "firebrick"),
    labels = c("FALSE" = "Not flagged", "TRUE" = "Flagged"),
    name = NULL
  ) +
  labs(
    title = "Measured Weight",
    x = "Measured weight, kg",
    y = "Number of participants"
  ) +
  theme_minimal()
```

Write briefly: do the flagged values look isolated, or do they look like part of a long right tail?

# 6. One-Variable Boxplot

This shows the distribution as one boxplot.

```{r}
ggplot(weight_data, aes(y = weight_kg)) +
  geom_boxplot() +
  geom_hline(
    yintercept = weight_cutoffs$upper_cutoff,
    color = "firebrick",
    linewidth = 1
  ) +
  labs(
    title = "Measured Weight With IQR Upper Fence",
    x = NULL,
    y = "Measured weight, kg"
  ) +
  theme_minimal()
```

Try removing `geom_hline()`. What did the line help show?

# 7. Grouped Boxplot

This compares measured weight by sex.

```{r}
ggplot(
  weight_data,
  aes(x = sex, y = weight_kg)
) +
  geom_boxplot() +
  labs(
    title = "Measured Weight By Sex",
    x = "Sex",
    y = "Measured weight, kg"
  ) +
  theme_minimal()
```

Try changing `weight_kg` to `measured_bmi`.

# 8. Add Jittered Values

This adds the individual values behind the boxplots.

```{r}
ggplot(
  weight_flagged,
  aes(x = sex, y = weight_kg)
) +
  geom_boxplot(outlier.shape = NA, fill = "white") +
  geom_jitter(width = 0.15, alpha = 0.25, size = 0.7) +
  labs(
    title = "Measured Weight By Sex",
    x = "Sex",
    y = "Measured weight, kg"
  ) +
  theme_minimal()
```

Try changing `alpha = 0.25` to `alpha = 0.6`.

# On Your Own

Choose one small task:

- Make a grouped boxplot for `height_cm` by `sex`.
- Make a grouped boxplot for `reported_weight_kg` by `sex`.
- Calculate IQR fences for a different numeric variable.

Write 1-2 sentences describing center, spread, and unusual values.
