---
title: "Practice: Quantiles, QQ Plots, And Transformations"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
```

# Goal

Practice creating quantile summaries, QQ plots, and transformed displays.

# 1. Load Data

```{r}
recovery <- readr::read_csv("data/real/post_coma_recovery.csv")
```

# 2. Prepare A Right-Skewed Variable

This keeps recovery assessments with positive, non-missing days post-coma.

```{r}
recovery_days <- recovery |>
  filter(
    !is.na(days_post_coma),
    days_post_coma > 0
  )

recovery_days |>
  select(patient_id, sex, age_at_injury_years, coma_duration_days, days_post_coma) |>
  head()
```

Try adding `filter(coma_duration_days > 0)` to focus on patients with recorded coma duration.

# 3. Selected Quantiles

These summaries calculate several percentiles.

```{r}
recovery_days |>
  summarize(
    p10 = quantile(days_post_coma, 0.10),
    p25 = quantile(days_post_coma, 0.25),
    p50 = quantile(days_post_coma, 0.50),
    p75 = quantile(days_post_coma, 0.75),
    p90 = quantile(days_post_coma, 0.90)
  )
```

Try adding another percentile, such as `p95 = quantile(days_post_coma, 0.95)`.

# 4. Raw Histogram

This histogram shows the raw days scale.

```{r}
ggplot(recovery_days, aes(x = days_post_coma)) +
  geom_histogram(binwidth = 250, fill = "gray75", color = "white") +
  labs(
    title = "Days Post-Coma At Assessment",
    x = "Days post-coma",
    y = "Number of assessments"
  ) +
  theme_minimal()
```

Try changing binwidth. Try smaller and larger values. What works best?

# 5. Raw QQ Plot

This QQ plot compares the raw variable to a normal reference distribution.

```{r}
ggplot(recovery_days, aes(sample = days_post_coma)) +
  stat_qq() +
  stat_qq_line(color = "firebrick", linewidth = 1) +
  labs(
    title = "QQ Plot Of Days Post-Coma",
    x = "Theoretical normal quantiles",
    y = "Observed days post-coma quantiles"
  ) +
  theme_minimal()
```

Write briefly: do the points stay near the line, curve away, or depart strongly in the tails?

What does this tell us about the distribution?

# 6. Log Transformation

This creates a log-transformed version of the same variable.

```{r}
recovery_log <- recovery_days |>
  mutate(
    log_days_post_coma = log(days_post_coma)
  )

recovery_log |>
  select(days_post_coma, log_days_post_coma) |>
  head()
```

# 7. Log Histogram

This histogram shows the transformed variable.

```{r}
ggplot(recovery_log, aes(x = log_days_post_coma)) +
  geom_histogram(binwidth = 0.1, fill = "gray75", color = "white") +
  labs(
    title = "Log Days Post-Coma At Assessment",
    x = "Log days post-coma",
    y = "Number of assessments"
  ) +
  theme_minimal()
```

Try changing `binwidth = 0.1` to `binwidth = 0.2`.

# 8. Log QQ Plot

This QQ plot compares the transformed variable to a normal reference distribution.

```{r}
ggplot(recovery_log, aes(sample = log_days_post_coma)) +
  stat_qq() +
  stat_qq_line(color = "firebrick", linewidth = 1) +
  labs(
    title = "QQ Plot Of Log Days Post-Coma",
    x = "Theoretical normal quantiles",
    y = "Observed log days post-coma quantiles"
  ) +
  theme_minimal()

```

Write briefly: did the log transformation make the QQ plot closer to the reference line?

# On Your Own

Choose one small task:

- Create a QQ plot for `coma_duration_days`.
- Repeat the quantile summary for `performance_iq`.
- Try a log transformation for another positive variable and decide whether it helps.

Write 2-3 sentences comparing the raw and transformed displays. Use cautious language: "seems," "suggests," or "appears" is better than a pass/fail claim.
