---
title: "Extra: Reusable Plot Styling"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
```

# Goal

This optional notebook shows how to avoid repeating the same plot styling code over and over.

The point is not to build a full design system. The point is to notice when repeated code can become a small reusable object or function.

# Load Data

```{r}
states <- readr::read_csv("data/real/state_population_ranks.csv")
```

# A Plain Plot

<!-- data-prep-anatomy-start -->
::: {.data-prep-anatomy}
| Data preparation step | Functions in this chunk |
|---|---|
| Keep or remove rows | `filter()` |
| Create or change columns | `mutate()` |
:::
<!-- data-prep-anatomy-end -->

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Aesthetic mappings | `x`; `y` |
| Geoms / stats | `geom_bar()` |
| Scales / coordinates / facets / labels | `coord_flip()`; `labs()` |
:::
<!-- plot-anatomy-end -->

```{r}
top_states <- states |>
  filter(rank <= 15) |>
  mutate(state_name = reorder(state_name, population))

p1 <- ggplot(top_states, aes(x = state_name, y = population)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(
    title = "Largest States By Population",
    x = NULL,
    y = "Population"
  )

p1
```

# Add Styling Once

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Scales / coordinates / facets / labels | `scale_y_continuous()`; `theme_minimal()`; `theme()` |
:::
<!-- plot-anatomy-end -->

```{r}
p2 <- p1 +
  scale_y_continuous(labels = scales::label_number(scale_cut = scales::cut_short_scale())) +
  theme_minimal() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold"),
    axis.title.x = element_text(margin = margin(t = 8))
  )

p2
```

This is fine for one plot. It becomes annoying when the same styling appears in many plots.

# Save Reusable Pieces

Start small. A reusable color value is often enough.

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Aesthetic mappings | `x`; `y` |
| Geoms / stats | `geom_bar()` |
| `geom_bar()` properties | `fill` |
| Scales / coordinates / facets / labels | `coord_flip()`; `scale_y_continuous()`; `labs()`; `theme_minimal()` |
:::
<!-- plot-anatomy-end -->

```{r}
course_blue <- "steelblue"
course_gray <- "gray75"

ggplot(top_states, aes(x = state_name, y = population)) +
  geom_bar(stat = "identity", fill = course_blue) +
  coord_flip() +
  scale_y_continuous(labels = scales::label_number(scale_cut = scales::cut_short_scale())) +
  labs(
    title = "Largest States By Population",
    x = NULL,
    y = "Population"
  ) +
  theme_minimal()
```

# Write A Small Theme Function

A theme function is useful when the same formatting appears in several figures.

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Scales / coordinates / facets / labels | `theme_minimal()`; `theme()` |
:::
<!-- plot-anatomy-end -->

```{r}
theme_course_plot <- function() {
  theme_minimal() +
    theme(
      panel.grid.major.y = element_blank(),
      panel.grid.minor = element_blank(),
      plot.title = element_text(face = "bold"),
      axis.title.x = element_text(margin = margin(t = 8))
    )
}
```

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Aesthetic mappings | `x`; `y` |
| Geoms / stats | `geom_bar()` |
| `geom_bar()` properties | `fill` |
| Scales / coordinates / facets / labels | `coord_flip()`; `scale_y_continuous()`; `labs()`; `theme_course_plot()` |
:::
<!-- plot-anatomy-end -->

```{r}
ggplot(top_states, aes(x = state_name, y = population)) +
  geom_bar(stat = "identity", fill = course_blue) +
  coord_flip() +
  scale_y_continuous(labels = scales::label_number(scale_cut = scales::cut_short_scale())) +
  labs(
    title = "Largest States By Population",
    x = NULL,
    y = "Population"
  ) +
  theme_course_plot()
```

# Highlight With Reusable Choices

<!-- data-prep-anatomy-start -->
::: {.data-prep-anatomy}
| Data preparation step | Functions in this chunk |
|---|---|
| Create or change columns | `mutate()` |
:::
<!-- data-prep-anatomy-end -->

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Aesthetic mappings | `x`; `y`; `fill` |
| Scales / coordinates / facets / labels | `coord_flip()`; `scale_fill_manual()`; `scale_y_continuous()`; `labs()`; `theme_course_plot()` |
:::
<!-- plot-anatomy-end -->

```{r}
top_states |>
  mutate(focus = state_name == "Texas") |>
  ggplot(aes(x = state_name, y = population, fill = focus)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_manual(
    values = c("TRUE" = "firebrick", "FALSE" = course_gray),
    guide = "none"
  ) +
  scale_y_continuous(labels = scales::label_number(scale_cut = scales::cut_short_scale())) +
  labs(
    title = "Highlight One State, Keep Context",
    x = NULL,
    y = "Population"
  ) +
  theme_course_plot()
```

# When Not To Reuse

Reusable styling can help, but it can also hide choices that should be made figure by figure.

Be careful with reusable code when:

- the figure needs a different scale
- the audience changes
- the color should encode data differently
- the plot type has different readability needs

# Takeaways

- Reusable colors and theme functions can reduce repeated code.
- Keep reusable styling small enough to understand.
- Do not let a theme function make design decisions that should depend on the specific figure.
