---
title: "Extra: Iterative Refinement, Aggregation, And Faceting"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
```

# Goal

This optional notebook shows how a figure can improve through several small revisions.

We focus on three ideas:

- start with one clear comparison
- use facets when repeating the same comparison across groups
- calculate a difference when the gap is the real question

# Load Data

```{r}
prams <- readr::read_csv("data/real/prams_2011_selected.csv")
```

Each row is a pre-aggregated percentage for one location, subgroup category, and subgroup.

# Prepare One Comparison

<!-- data-prep-anatomy-start -->
::: {.data-prep-anatomy}
| Data preparation step | Functions in this chunk |
|---|---|
| Keep or remove rows | `filter()` |
| Choose or rename columns | `select()` |
:::
<!-- data-prep-anatomy-end -->

```{r}
birth_history <- prams |>
  filter(subgroup_cat == "Number of Previous Live Births") |>
  filter(!is.na(depression_within_3_months_birth))

birth_history |>
  select(location_abbr, location, subgroup, depression_within_3_months_birth) |>
  head()
```

# Start With One State

<!-- data-prep-anatomy-start -->
::: {.data-prep-anatomy}
| Data preparation step | Functions in this chunk |
|---|---|
| Keep or remove rows | `filter()` |
:::
<!-- data-prep-anatomy-end -->

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Aesthetic mappings | `x`; `y` |
| Geoms / stats | `geom_bar()` |
| `geom_bar()` properties | `fill` |
| Scales / coordinates / facets / labels | `labs()`; `theme_minimal()` |
:::
<!-- plot-anatomy-end -->

```{r}
wisconsin_birth_history <- birth_history |>
  filter(location_abbr == "WI")

ggplot(
  wisconsin_birth_history,
  aes(x = subgroup, y = depression_within_3_months_birth)
) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(
    title = "Depression Before Birth By Previous Live Births",
    x = "Previous live births",
    y = "Percent reporting depression"
  ) +
  theme_minimal()
```

This answers a narrow question for one state.

# Repeat The Comparison Across States

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Scales / coordinates / facets / labels | `facet_wrap()`; `labs()`; `theme_minimal()` |
:::
<!-- plot-anatomy-end -->

```{r}
ggplot(
  birth_history,
  aes(x = subgroup, y = depression_within_3_months_birth)
) +
  geom_bar(stat = "identity", fill = "steelblue") +
  facet_wrap(~ location_abbr) +
  labs(
    title = "The Same Comparison Repeated Across States",
    x = "Previous live births",
    y = "Percent reporting depression"
  ) +
  theme_minimal()
```

Facets help repeat the same comparison. They also create a lot to scan.

# Calculate The Gap

If the question is about how the two subgroups differ, calculate that difference directly.

<!-- data-prep-anatomy-start -->
::: {.data-prep-anatomy}
| Data preparation step | Functions in this chunk |
|---|---|
| Keep or remove rows | `filter()` |
| Choose or rename columns | `select()` |
| Create or change columns | `mutate()` |
| Sort rows | `arrange()` |
| Reshape data | `pivot_wider()` |
:::
<!-- data-prep-anatomy-end -->

```{r}
state_gaps <- birth_history |>
  select(location_abbr, location, subgroup, depression_within_3_months_birth) |>
  tidyr::pivot_wider(
    names_from = subgroup,
    values_from = depression_within_3_months_birth
  ) |>
  mutate(
    difference = `0` - `1 or more`
  ) |>
  filter(!is.na(difference)) |>
  arrange(difference) |>
  mutate(location = factor(location, levels = location))

state_gaps |>
  select(location, `0`, `1 or more`, difference) |>
  head()
```

# Plot The Difference

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Aesthetic mappings | `x`; `y`; `xend`; `yend` |
| Geoms / stats | `geom_hline()`; `geom_segment()`; `geom_point()` |
| `geom_hline()` properties | `color` |
| `geom_segment()` properties | `color` |
| `geom_point()` properties | `size`; `color` |
| Scales / coordinates / facets / labels | `coord_flip()`; `labs()`; `theme_minimal()` |
:::
<!-- plot-anatomy-end -->

```{r}
ggplot(state_gaps, aes(x = location, y = difference)) +
  geom_hline(yintercept = 0, color = "gray60") +
  geom_segment(aes(xend = location, y = 0, yend = difference), color = "gray70") +
  geom_point(size = 2.5, color = "steelblue") +
  coord_flip() +
  labs(
    title = "Difference In Depression Estimates By Birth History",
    subtitle = "Positive values mean higher estimates for first births",
    x = NULL,
    y = "Percentage-point difference: 0 previous births minus 1 or more"
  ) +
  theme_minimal()
```

The difference plot is simpler because it plots the quantity we are comparing.

# Why This Is Iteration

The final figure is not just a prettier version of the first plot. The analytic question changed:

- first plot: what are the two values in Wisconsin?
- faceted plot: how does that comparison repeat across states?
- difference plot: where is the gap larger or smaller?

# Takeaways

- Iteration should clarify the question, not just decorate the plot.
- Facets are useful when the same comparison repeats across groups.
- If the difference is the question, calculate and plot the difference.
