---
title: "Extra: Correlation Patterns And Coefficient Plots"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
```

# Goal

This optional notebook shows two model-adjacent visualization ideas:

- the same correlation can hide different visual patterns
- model coefficients can be plotted directly when the model result is the focus

# Same Correlation, Different Patterns

A correlation is one number. It does not show form, spread, outliers, or groups.

<!-- data-prep-anatomy-start -->
::: {.data-prep-anatomy}
| Data preparation step | Functions in this chunk |
|---|---|
| Count or summarize | `group_by()`; `summarize()` |
:::
<!-- data-prep-anatomy-end -->

```{r}
set.seed(123)

correlation_examples <- tibble::tibble(
  pattern = rep(
    c("Linear", "Curved", "One outlier", "Two groups"),
    each = 80
  ),
  x = c(
    rnorm(80),
    runif(80, -2, 2),
    c(rnorm(79), 4),
    c(rnorm(40, -1.5), rnorm(40, 1.5))
  ),
  y = c(
    0.6 * x[1:80] + rnorm(80, sd = 0.8),
    0.8 * x[81:160]^2 + rnorm(80, sd = 0.4),
    c(rnorm(79), 4),
    c(rnorm(40, 1.5), rnorm(40, -1.5))
  )
)

correlation_examples |>
  group_by(pattern) |>
  summarize(correlation = cor(x, y), .groups = "drop")
```

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Aesthetic mappings | `x`; `y` |
| Geoms / stats | `geom_point()`; `geom_smooth()` |
| `geom_point()` properties | `alpha` |
| `geom_smooth()` properties | `se`; `color` |
| Scales / coordinates / facets / labels | `facet_wrap()`; `labs()`; `theme_minimal()` |
:::
<!-- plot-anatomy-end -->

```{r}
ggplot(correlation_examples, aes(x = x, y = y)) +
  geom_point(alpha = 0.65) +
  geom_smooth(se = FALSE, color = "steelblue") +
  facet_wrap(~ pattern, scales = "free") +
  labs(
    title = "Correlation Does Not Replace The Scatterplot",
    x = "x",
    y = "y"
  ) +
  theme_minimal()
```

The scatterplots show information that the correlations compress.

# Plot A Model Coefficient Table

Sometimes the model result is a set of coefficients rather than a fitted line.

This example uses a simple linear model for systolic blood pressure.

<!-- data-prep-anatomy-start -->
::: {.data-prep-anatomy}
| Data preparation step | Functions in this chunk |
|---|---|
| Keep or remove rows | `filter()` |
:::
<!-- data-prep-anatomy-end -->

```{r}
nhanes <- readr::read_csv("data/real/nhanes_2021_2023_analysis.csv")

bp_model_data <- nhanes |>
  filter(
    age_years >= 18,
    !is.na(systolic_mean),
    !is.na(age_years),
    !is.na(gender),
    !is.na(family_income_to_poverty)
  )

bp_model <- lm(
  systolic_mean ~ age_years + gender + family_income_to_poverty,
  data = bp_model_data
)

summary(bp_model)
```

Extract the coefficient estimates and confidence intervals.

<!-- data-prep-anatomy-start -->
::: {.data-prep-anatomy}
| Data preparation step | Functions in this chunk |
|---|---|
| Keep or remove rows | `filter()` |
| Choose or rename columns | `rename()` |
| Create or change columns | `mutate()`; `case_when()` |
| Combine data frames | `left_join()` |
:::
<!-- data-prep-anatomy-end -->

```{r}
coefficient_table <- as.data.frame(coef(summary(bp_model))) |>
  tibble::rownames_to_column("term") |>
  rename(
    estimate = Estimate,
    standard_error = `Std. Error`,
    statistic = `t value`,
    p_value = `Pr(>|t|)`
  )

interval_table <- as.data.frame(confint(bp_model)) |>
  tibble::rownames_to_column("term") |>
  rename(conf_low = `2.5 %`, conf_high = `97.5 %`)

coefficient_plot_data <- coefficient_table |>
  left_join(interval_table, by = "term") |>
  filter(term != "(Intercept)") |>
  mutate(
    term_label = case_when(
      term == "age_years" ~ "Age, years",
      term == "gendermale" ~ "Gender: male",
      term == "family_income_to_poverty" ~ "Family income-to-poverty ratio",
      TRUE ~ term
    ),
    term_label = reorder(term_label, estimate)
  )

coefficient_plot_data
```

# Coefficient Plot

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Aesthetic mappings | `x`; `y`; `xmin`; `xmax` |
| Geoms / stats | `geom_vline()`; `geom_errorbarh()`; `geom_point()` |
| `geom_vline()` properties | `color`; `linetype` |
| `geom_errorbarh()` properties | `height`; `color` |
| `geom_point()` properties | `size`; `color` |
| Scales / coordinates / facets / labels | `labs()`; `theme_minimal()` |
:::
<!-- plot-anatomy-end -->

```{r}
ggplot(coefficient_plot_data, aes(x = estimate, y = term_label)) +
  geom_vline(xintercept = 0, color = "gray60", linetype = "dashed") +
  geom_errorbarh(aes(xmin = conf_low, xmax = conf_high), height = 0.15, color = "gray50") +
  geom_point(size = 2.5, color = "steelblue") +
  labs(
    title = "Coefficient Plot For A Simple Blood Pressure Model",
    x = "Estimated difference in mean systolic blood pressure",
    y = NULL
  ) +
  theme_minimal()
```

Read this as a model summary, not raw data. Each point is an estimated coefficient from the fitted model. The interval gives a range of plausible values under the model.

# Takeaways

- Correlations should be paired with scatterplots.
- Curvature, outliers, and groups can change the story.
- Coefficient plots are useful when the model output is the object being communicated.
- A coefficient plot needs clear labels because the terms come from a model formula.
