---
title: "Extra: Cleveland-Style Diagnostic Plots"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
```

# Goal

This optional notebook shows a few Cleveland-inspired diagnostic graphics that go beyond the main Day 3 material.

These plots are useful when we want to compare location, spread, and residual structure across groups.

# Simulate Example Data

<!-- data-prep-anatomy-start -->
::: {.data-prep-anatomy}
| Data preparation step | Functions in this chunk |
|---|---|
| Keep or remove rows | `filter()` |
| Count or summarize | `group_by()`; `summarize()`; `n()` |
| Combine data frames | `bind_rows()` |
:::
<!-- data-prep-anatomy-end -->

```{r}
set.seed(123)

hospital_data <- bind_rows(
  tibble::tibble(hospital = "Hospital A", recovery_days = rnorm(120, mean = 8, sd = 2)),
  tibble::tibble(hospital = "Hospital B", recovery_days = rnorm(120, mean = 10, sd = 2)),
  tibble::tibble(hospital = "Hospital C", recovery_days = exp(rnorm(120, mean = 2.1, sd = 0.45))),
  tibble::tibble(
    hospital = "Hospital D",
    recovery_days = c(rnorm(60, mean = 6, sd = 1.2), rnorm(60, mean = 12, sd = 1.5))
  )
) |>
  filter(recovery_days > 0)

hospital_data |>
  group_by(hospital) |>
  summarize(
    n = n(),
    mean_recovery = mean(recovery_days),
    median_recovery = median(recovery_days),
    iqr_recovery = IQR(recovery_days),
    .groups = "drop"
  )
```

# Start With The Distributions

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Aesthetic mappings | `x`; `y` |
| Geoms / stats | `geom_boxplot()`; `geom_jitter()` |
| `geom_boxplot()` properties | `fill`; `outlier.alpha` |
| `geom_jitter()` properties | `width`; `alpha`; `size` |
| Scales / coordinates / facets / labels | `labs()`; `theme_minimal()` |
:::
<!-- plot-anatomy-end -->

```{r}
ggplot(hospital_data, aes(x = hospital, y = recovery_days)) +
  geom_boxplot(fill = "gray90", outlier.alpha = 0.35) +
  geom_jitter(width = 0.12, alpha = 0.2, size = 1) +
  labs(
    title = "Recovery Time By Hospital",
    x = NULL,
    y = "Recovery days"
  ) +
  theme_minimal()
```

The boxplots show center, spread, and unusual values. They do not directly show how spread changes with location.

# Location And Spread Together

Calculate one location and spread summary per hospital.

<!-- data-prep-anatomy-start -->
::: {.data-prep-anatomy}
| Data preparation step | Functions in this chunk |
|---|---|
| Count or summarize | `group_by()`; `summarize()` |
:::
<!-- data-prep-anatomy-end -->

```{r}
hospital_summary <- hospital_data |>
  group_by(hospital) |>
  summarize(
    median_recovery = median(recovery_days),
    iqr_recovery = IQR(recovery_days),
    .groups = "drop"
  )

hospital_summary
```

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Aesthetic mappings | `x`; `y`; `label` |
| Geoms / stats | `geom_point()`; `geom_text()` |
| `geom_point()` properties | `size`; `color` |
:::
<!-- plot-anatomy-end -->

```{r}
ggplot(hospital_summary, aes(x = median_recovery, y = iqr_recovery, label = hospital)) +
  geom_point(size = 3, color = "steelblue") +
  geom_text(nudge_y = 0.25) +
  labs(
    title = "Location And Spread By Hospital",
    x = "Median recovery days",
    y = "Interquartile range"
  ) +
  theme_minimal()
```

This plot asks whether hospitals with larger typical values also have larger spread.

# Residuals From Group Means

Residuals can be useful even in a simple group comparison. Here, a residual is each observation minus its hospital mean.

<!-- data-prep-anatomy-start -->
::: {.data-prep-anatomy}
| Data preparation step | Functions in this chunk |
|---|---|
| Choose or rename columns | `select()` |
| Create or change columns | `mutate()` |
| Count or summarize | `group_by()`; `ungroup()` |
:::
<!-- data-prep-anatomy-end -->

```{r}
hospital_residuals <- hospital_data |>
  group_by(hospital) |>
  mutate(
    fitted_group_mean = mean(recovery_days),
    residual = recovery_days - fitted_group_mean
  ) |>
  ungroup()

hospital_residuals |>
  select(hospital, recovery_days, fitted_group_mean, residual) |>
  head()
```

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Aesthetic mappings | `x`; `y` |
| Geoms / stats | `geom_hline()`; `geom_boxplot()`; `geom_jitter()` |
| `geom_hline()` properties | `color` |
| `geom_boxplot()` properties | `fill`; `outlier.alpha` |
| `geom_jitter()` properties | `width`; `alpha`; `size` |
:::
<!-- plot-anatomy-end -->

```{r}
ggplot(hospital_residuals, aes(x = hospital, y = residual)) +
  geom_hline(yintercept = 0, color = "gray60") +
  geom_boxplot(fill = "gray90", outlier.alpha = 0.35) +
  geom_jitter(width = 0.12, alpha = 0.2, size = 1) +
  labs(
    title = "Residuals From Hospital Means",
    x = NULL,
    y = "Observed minus hospital mean"
  ) +
  theme_minimal()
```

This shows which hospitals have residuals that are symmetric, skewed, or unusually spread out after removing the group mean.

# Residuals Against Fitted Values

Plot residual spread against the fitted group means.

<!-- plot-anatomy-start -->
::: {.plot-anatomy}
| Plot component | Code in this chunk |
|---|---|
| Aesthetic mappings | `x`; `y`; `color` |
| Geoms / stats | `geom_hline()`; `geom_jitter()` |
| `geom_jitter()` properties | `width`; `alpha` |
:::
<!-- plot-anatomy-end -->

```{r}
ggplot(hospital_residuals, aes(x = fitted_group_mean, y = residual, color = hospital)) +
  geom_hline(yintercept = 0, color = "gray60") +
  geom_jitter(width = 0.12, alpha = 0.45) +
  labs(
    title = "Residual Spread Around Group Means",
    x = "Fitted hospital mean",
    y = "Residual"
  ) +
  theme_minimal()
```

This is a compact way to ask whether spread changes as fitted values increase.

# Takeaways

- Boxplots and violins are usually enough for early group comparisons.
- Location-spread plots help compare center and variability together.
- Residual plots can reveal skew, unequal spread, and unusual observations after a simple summary has been removed.
- These are diagnostic tools, not final communication figures for every audience.
