---
title: "Practice: Time Trends And Small Multiples"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# Goal

Practice plotting change over time and separating groups into small multiples when needed.

# Suggested Data

Choose one:

- `data/real/flu_vaccination_age_time_clean.csv`
- `data/real/migraine_headache_diary.csv`
- `data/simulated/respiratory_booster_coverage.csv`

Helpful module reference: `modules/06_change/01_time-trends-and-small-multiples.qmd`.

# Load And Inspect

```{r}
# Load your dataset here.
```

```{r}
# Identify the time variable, measure, and grouping variables.
```

# Prepare A Time Series

Choose one outcome over time. Filter to a focused comparison.

```{r}
# Prepare the time trend data here.
```

# Plot

Make a line plot. If there are several groups, try color or facets.

```{r}
# Build your time trend plot here.
```

# Interpret

> TODO: What changes over time? Are there group differences? Are the time points equally spaced and comparable?
