---
title: "Practice: Time Trends And Small Multiples"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# What We Practiced

The module showed how to:

- put ordered time on the x-axis
- compare change across groups
- use color for a few groups
- use small multiples when color becomes crowded

# Task

Pick one dataset and create:

1. a focused time-series table
2. one line plot
3. one grouped or faceted version if useful
4. a short interpretation of change over time

Suggested datasets:

- `data/real/flu_vaccination_age_time_clean.csv`
- `data/real/migraine_headache_diary.csv`
- `data/simulated/respiratory_booster_coverage.csv`

Helpful module reference: `modules/06_change/01_time-trends-and-small-multiples.qmd`.

# Work

```{r}
# Load data, prepare a focused time series, and build one or two time plots.
```

# Interpretation

Write briefly: what changes over time, whether groups differ, and whether the time points are comparable.
