---
title: "Practice: Before-After And Trajectories"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# Goal

Practice visualizing repeated measurements, before-after comparisons, or trajectories.

# Suggested Data

Choose one:

- `data/real/migraine_headache_diary.csv`
- `data/real/post_coma_recovery.csv`
- `data/simulated/toothache_diary.csv`

Helpful module reference: `modules/06_change/02_before-after-and-trajectories.qmd`.

# Load And Inspect

```{r}
# Load your dataset here.
```

```{r}
# Identify the time/stage variable and the outcome.
```

# Summarize Or Select Trajectories

Decide whether to show individual trajectories, group summaries over time, or before-after differences.

```{r}
# Prepare trajectory or before-after data here.
```

# Plot

```{r}
# Build your trajectory or before-after plot here.
```

# Interpret

> TODO: What changes? Are you showing individual change, group change, or both?
