---
title: "Practice: Grouped Distributions"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
```

# Goal

Practice comparing a continuous variable across groups before reducing each group to one number.

# 1. Load Data

```{r}
nhanes <- readr::read_csv("data/real/nhanes_2021_2023_analysis.csv")

nhanes |>
  glimpse()
```

# 2. Prepare Adult Blood Pressure Data

This keeps adults with a non-missing mean systolic blood pressure value.

```{r}
adult_bp <- nhanes |>
  filter(
    age_years >= 18,
    !is.na(age_group),
    !is.na(systolic_mean)
  )

adult_bp |>
  select(respondent_id, age_years, age_group, gender, systolic_mean) |>
  head()
```

Try adding `filter(!is.na(gender))` if you want to compare gender groups later.

# 3. Check Group Sizes

Group sizes affect how much we should trust a grouped distribution comparison.

```{r}
adult_bp |>
  count(age_group)
```

# 4. Summarize By Group

These summaries give center and spread by age group.

```{r}
bp_summary <- adult_bp |>
  group_by(age_group) |>
  summarize(
    n = n(),
    median_systolic = median(systolic_mean),
    iqr_systolic = IQR(systolic_mean),
    mean_systolic = mean(systolic_mean),
    .groups = "drop"
  )

bp_summary
```

Try adding `max_systolic = max(systolic_mean)`.

# 5. Plot Grouped Distributions

This boxplot compares the full distributions, not just one summary.

```{r}
ggplot(adult_bp, aes(x = age_group, y = systolic_mean)) +
  geom_boxplot(fill = "gray90", outlier.alpha = 0.4) +
  labs(
    title = "Adult Systolic Blood Pressure By Age Group",
    x = "Age group",
    y = "Mean systolic blood pressure"
  ) +
  theme_minimal()
```

Try adding `geom_jitter(width = 0.12, alpha = 0.15)` after `geom_boxplot()`.

# Interpretation

Write 1-2 sentences: which group differs in center, spread, or unusual values? What does the plot show that the summary table compresses?
