---
title: "Practice: Grouped Summaries"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
```

# Goal

Practice calculating one statistic per group and plotting the grouped summary.

# 1. Load Data

```{r}
places <- readr::read_csv("data/real/places_county_core_measures_long.csv")

places |>
  glimpse()
```

# 2. Prepare One Measure

This keeps county obesity estimates for Maryland, Virginia, and the District of Columbia.

```{r}
obesity_counties <- places |>
  filter(measure_key == "obesity") |>
  filter(state %in% c("MD", "VA", "DC")) |>
  filter(!is.na(estimate))

obesity_counties |>
  select(state_name, county_name, measure, estimate, ci_lower, ci_upper) |>
  head()
```

Try changing `measure_key == "obesity"` to `measure_key == "high_blood_pressure"`.

# 3. Check Group Sizes

Here, each row is a county estimate. Count counties before comparing state summaries.

```{r}
obesity_counties |>
  count(state_name)
```

# 4. Summarize By State

These summaries reduce many county estimates to one row per state.

```{r}
state_obesity <- obesity_counties |>
  group_by(state_name) |>
  summarize(
    n_counties = n(),
    mean_estimate = mean(estimate),
    median_estimate = median(estimate),
    min_estimate = min(estimate),
    max_estimate = max(estimate),
    .groups = "drop"
  ) |>
  mutate(
    state_name = reorder(state_name, mean_estimate)
  )

state_obesity
```

Try adding `sd_estimate = sd(estimate)`.

# 5. Plot The Summary

This dot plot shows one summary value per state.

```{r}
ggplot(
  state_obesity,
  aes(x = state_name, y = mean_estimate)
) +
  geom_point(size = 3, color = "steelblue") +
  coord_flip() +
  labs(
    title = "Mean County Obesity Estimate By State",
    x = NULL,
    y = "Mean county estimate (%)"
  ) +
  theme_minimal()
```

Try changing `mean_estimate` to `median_estimate` in the plot.

# Interpretation

Write 1-2 sentences: what does one point represent, and what information was lost when counties were summarized to states? What should we do in a situation where we have very few data points, like DC? o
