---
title: "Practice: Difference Plots"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
```

# Goal

Practice calculating a difference and plotting the difference directly.

Most of the code is complete. Study the output, then make one small change at a time.

# 1. Load Data

```{r}
davis <- readr::read_csv("data/real/davis_height_weight.csv")
```

# 2. Calculate Reporting Error

This keeps participants with measured and reported height, then calculates reported minus measured height.

```{r}
height_error <- davis |>
  filter(!missing_reported_height) |>
  filter(!likely_swapped_measured_values) |>
  transmute(
    participant_id,
    sex,
    measured_height = height_cm,
    reported_height = reported_height_cm,
    difference = reported_height - measured_height
  )

height_error |>
  head()
```

Positive values mean the reported value is higher than the measured value. Negative values mean the reported value is lower.

# 3. Inspect The Difference

```{r}
height_error |>
  summarize(
    n = n(),
    mean_difference = mean(difference),
    median_difference = median(difference),
    min_difference = min(difference),
    max_difference = max(difference)
  )
```

Try adding `sd_difference = sd(difference)`.

# 4. Plot Larger Differences

This plot keeps participants with reporting differences of at least 5 cm in either direction.

```{r}
large_height_errors <- height_error |>
  filter(abs(difference) >= 5) |>
  arrange(difference) |>
  mutate(
    participant_id = factor(participant_id, levels = participant_id)
  )

ggplot(
  large_height_errors,
  aes(x = participant_id, y = difference)
) +
  geom_hline(yintercept = 0, color = "gray60") +
  geom_point(size = 2, color = "steelblue") +
  coord_flip() +
  labs(
    title = "Larger Height Reporting Differences",
    x = "Participant",
    y = "Reported minus measured height, cm"
  ) +
  theme_minimal()
```

Try changing `abs(difference) >= 5` to `abs(difference) >= 3` to `abs(difference) >= 0`.

This is an individual-level check, because `participant_id` is on the axis. This is useful for exploration, but not for reporting.

# 5. Group Differences By Sex

For a grouped comparison, summarize the reporting difference by a meaningful category.

```{r}
sex_height_error <- height_error |>
  group_by(sex) |>
  summarize(
    n = n(),
    mean_difference = mean(difference),
    se_difference = sd(difference) / sqrt(n),
    lower = mean_difference - 1.96 * se_difference,
    upper = mean_difference + 1.96 * se_difference,
    .groups = "drop"
  )

sex_height_error
```

```{r}
ggplot(sex_height_error, aes(x = sex, y = mean_difference)) +
  geom_hline(yintercept = 0, color = "gray60") +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.15, color = "gray40") +
  geom_point(size = 3, color = "steelblue") +
  labs(
    title = "Average Height Reporting Difference By Sex",
    x = "Sex",
    y = "Mean reported minus measured height, cm"
  ) +
  theme_minimal()
```

Try adding `median_difference = median(difference)` to the grouped summary. Does the median tell the same story as the mean?

# Interpretation

Write 1-2 sentences: what does zero mean, what does a positive value mean, and what does a negative value mean?
