---
title: "Practice: Scatterplots, Smoothing, And Groups"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
```

# Goal

Practice visualizing association between two numeric variables.

# 1. Load Data

```{r}
davis <- readr::read_csv("data/real/davis_height_weight.csv")
```

# 2. Prepare Data

This keeps participants with measured and reported weight values.

```{r}
weight_data <- davis |>
  filter(
    !missing_reported_weight,
    !is.na(weight_kg),
    !is.na(reported_weight_kg),
    !likely_swapped_measured_values
  )

weight_data |>
  select(participant_id, sex, weight_kg, reported_weight_kg, weight_reporting_error_kg) |>
  head()
```

# 3. Basic Scatterplot

This plot compares measured weight with reported weight.

```{r}
ggplot(weight_data, aes(x = weight_kg, y = reported_weight_kg)) +
  geom_point(alpha = 0.7) +
  labs(
    title = "Measured And Reported Weight",
    x = "Measured weight (kg)",
    y = "Reported weight (kg)"
  ) +
  theme_minimal()
```

Try adding `color = sex` inside `aes()`.

# 4. Add A Reference Line

The diagonal line shows perfect agreement between measured and reported weight.

```{r}
ggplot(weight_data, aes(x = weight_kg, y = reported_weight_kg)) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, color = "gray50") +
  labs(
    title = "Measured And Reported Weight",
    x = "Measured weight (kg)",
    y = "Reported weight (kg)"
  ) +
  theme_minimal()
```

Try changing the point transparency to `alpha = 0.4`.

# 5. Plot The Error Directly

Sometimes a derived variable makes the question clearer.

Here, reported minus measured weight puts the reporting error on the y-axis. Values below zero mean under-reporting.

```{r}
ggplot(weight_data, aes(x = weight_kg, y = weight_reporting_error_kg)) +
  geom_hline(yintercept = 0, color = "gray60") +
  geom_point(alpha = 0.7) +
  labs(
    title = "Weight Reporting Error By Measured Weight",
    x = "Measured weight (kg)",
    y = "Reported minus measured weight (kg)"
  ) +
  theme_minimal()
```

Try adding `geom_smooth(se = FALSE)` after `geom_point()`.

# 6. Add Groups

This version compares the reporting-error pattern by sex.

```{r}
ggplot(
  weight_data,
  aes(x = weight_kg, y = weight_reporting_error_kg, color = sex)
) +
  geom_hline(yintercept = 0, color = "gray70") +
  geom_point(alpha = 0.7) +
  geom_smooth(se = FALSE) +
  labs(
    title = "Weight Reporting Error By Measured Weight And Sex",
    x = "Measured weight (kg)",
    y = "Reported minus measured weight (kg)",
    color = "Sex"
  ) +
  theme_minimal()
```

Try a different smoother type, such as `gam`.

Then try replacing color with `facet_wrap(~ sex)`.

# Interpretation

Write 1-2 sentences describing direction, form, spread, and what zero means in the error plot.
