---
title: "Practice: Model Output Plots"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
```

# Goal

Practice plotting simple model output while keeping the raw data visible.

This is optional extension practice if there is time.

# 1. Load Data

```{r}
davis <- readr::read_csv("data/real/davis_height_weight.csv")
```

# 2. Prepare Data

This keeps participants with measured height, measured weight, and sex.

```{r}
body_size <- davis |>
  filter(!is.na(height_cm)) |>
  filter(!is.na(weight_kg)) |>
  filter(!is.na(sex)) |>
  filter(!likely_swapped_measured_values)

body_size |>
  select(participant_id, sex, height_cm, weight_kg) |>
  head()
```

# 3. Plot The Data First

```{r}
ggplot(body_size, aes(x = height_cm, y = weight_kg, color = sex)) +
  geom_point(alpha = 0.6) +
  labs(
    title = "Measured Height And Weight",
    x = "Measured height, cm",
    y = "Measured weight, kg",
    color = "Sex"
  ) +
  theme_minimal()
```

# 4. Fit A Simple Model

We are using the model to create a visual summary, not to turn this into a modeling exercise.

```{r}
weight_model <- lm(weight_kg ~ height_cm + sex, data = body_size)

summary(weight_model)
```

# 5. Create Prediction Data

```{r}
prediction_grid <- expand.grid(
  height_cm = seq(145, 195, by = 5),
  sex = sort(unique(body_size$sex))
)

predictions <- predict(
  weight_model,
  newdata = prediction_grid,
  interval = "confidence"
)

prediction_data <- bind_cols(
  prediction_grid,
  as.data.frame(predictions)
)

head(prediction_data)
```

# 6. Plot Predictions With Raw Data

```{r}
ggplot(body_size, aes(x = height_cm, y = weight_kg)) +
  geom_point(aes(color = sex), alpha = 0.25) +
  geom_line(
    data = prediction_data,
    aes(x = height_cm, y = fit, color = sex),
    linewidth = 1
  ) +
  labs(
    title = "Predicted Weight By Height And Sex",
    x = "Measured height, cm",
    y = "Measured weight, kg",
    color = "Sex"
  ) +
  theme_minimal()
```

Try removing the raw points. Do you think this is better or worse?

