---
title: "Histograms And Density Plots"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# Goal

This notebook shows how to visualize the distribution of one continuous variable.

We focus on center, spread, shape, and unusual values.

# Analytic Questions

We will use a simplified NHANES teaching file to ask:

- What does the distribution of adult systolic blood pressure look like?
- How does the bin width change a histogram?
- What does a density plot emphasize differently from a histogram?

# Load Data

```{r}
nhanes_path <- "data/real/nhanes_2021_2023_analysis.csv"

nhanes <- read_csv(nhanes_path, show_col_types = FALSE)

nhanes |>
  glimpse()
```

Each row is one participant in the joined NHANES teaching file.

NHANES is a complex survey. In this course, these are exploratory sample visualizations, not national estimates.

# Prepare Adult Blood Pressure Data

```{r}
adult_bp <- nhanes |>
  filter(
    age_years >= 18,
    !is.na(systolic_mean)
  )

adult_bp |>
  select(respondent_id, age_years, gender, age_group, systolic_mean) |>
  head()
```

# First Histogram

Start with the data and x-axis mapping.

```{r}
p1 <- ggplot(
  adult_bp,
  aes(x = systolic_mean)
)

p1
```

Add a histogram.

```{r}
p2 <- p1 +
  geom_histogram(binwidth = 5)

p2
```

Add labels and a theme.

```{r}
p3 <- p2 +
  labs(
    title = "Distribution Of Adult Mean Systolic Blood Pressure",
    subtitle = "NHANES 2021-2023 teaching file, adults with BP readings",
    x = "Mean systolic blood pressure",
    y = "Number of participants"
  ) +
  theme_minimal()

p3
```

Interpretation: the distribution is centered a little above 120 mmHg and has a right tail with some higher readings.

# Bin Width

Bin width changes the level of detail in the histogram.

```{r}
p1 <- ggplot(adult_bp, aes(x = systolic_mean)) +
  geom_histogram(binwidth = 2)

p1
```

```{r}
p2 <- ggplot(adult_bp, aes(x = systolic_mean)) +
  geom_histogram(binwidth = 10)

p2
```

The smaller bin width shows more detail. The larger bin width smooths over more of the small variation.

# Density Plot

A density plot shows a smoothed distribution.

The y-axis is density, not number of participants. The total area under the curve is scaled to 1.

```{r}
p1 <- ggplot(adult_bp, aes(x = systolic_mean))

p1
```

```{r}
p2 <- p1 +
  geom_density(fill = "steelblue", alpha = 0.3)

p2
```

```{r}
p3 <- p2 +
  labs(
    title = "Density Of Adult Mean Systolic Blood Pressure",
    x = "Mean systolic blood pressure",
    y = "Density"
  ) +
  theme_minimal()

p3
```

# Summary Statistics

Use summaries to support what the plot suggests.

```{r}
adult_bp |>
  summarise(
    n = n(),
    mean_systolic = mean(systolic_mean),
    median_systolic = median(systolic_mean),
    min_systolic = min(systolic_mean),
    max_systolic = max(systolic_mean),
    iqr_systolic = IQR(systolic_mean)
  )
```

# Takeaways

- Histograms are good for count-based distribution questions.
- Bin width is a choice that affects what patterns are visible.
- Density plots emphasize smoothed shape rather than counts.
- Always connect distribution shape to the unit of observation and the analytic question.
