---
title: "Grouped Distributions"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# Goal

This notebook shows how to compare continuous distributions across groups.

The goal is not only to compare averages. We also care about spread, shape, and overlap.

# Analytic Questions

We will ask:

- How does adult systolic blood pressure vary by age group?
- What do box plots show well?
- What do violin plots show differently?

# Load Data

```{r}
nhanes_path <- "data/real/nhanes_2021_2023_analysis.csv"

nhanes <- read_csv(nhanes_path, show_col_types = FALSE)

nhanes |>
  glimpse()
```

# Prepare Data

```{r}
adult_bp <- nhanes |>
  filter(
    age_years >= 18,
    !is.na(systolic_mean),
    !is.na(age_group)
  )

adult_bp |>
  count(age_group)
```

# Box Plot

Start with a grouped mapping.

```{r}
p1 <- ggplot(
  adult_bp,
  aes(x = age_group, y = systolic_mean)
)

p1
```

Add a box plot.

```{r}
p2 <- p1 +
  geom_boxplot()

p2
```

Add labels and a theme.

```{r}
p3 <- p2 +
  labs(
    title = "Adult Systolic Blood Pressure By Age Group",
    subtitle = "Box plots show median, middle 50%, and potential outliers",
    x = "Age group",
    y = "Mean systolic blood pressure"
  ) +
  theme_minimal()

p3
```

Interpretation: older age groups tend to have higher median systolic blood pressure, but there is still substantial overlap across groups.

# Add Points Carefully

Points can show the observations, but they need transparency when there are many rows.

```{r}
p1 <- ggplot(
  adult_bp,
  aes(x = age_group, y = systolic_mean)
) +
  geom_boxplot(outlier.shape = NA)

p1
```

```{r}
p2 <- p1 +
  geom_jitter(width = 0.15, alpha = 0.15, size = 0.8)

p2
```

```{r}
p3 <- p2 +
  labs(
    title = "Adult Systolic Blood Pressure By Age Group",
    subtitle = "Jittered points show the individual readings behind the boxes",
    x = "Age group",
    y = "Mean systolic blood pressure"
  ) +
  theme_minimal()

p3
```

# Violin Plot

Violin plots emphasize distribution shape within each group.

```{r}
p1 <- ggplot(
  adult_bp,
  aes(x = age_group, y = systolic_mean)
)

p1
```

```{r}
p2 <- p1 +
  geom_violin(fill = "gray85")

p2
```

```{r}
p3 <- p2 +
  geom_boxplot(width = 0.12, outlier.shape = NA)

p3
```

```{r}
p4 <- p3 +
  labs(
    title = "Distribution Shape By Age Group",
    subtitle = "Violins show shape; box plots show median and middle spread",
    x = "Age group",
    y = "Mean systolic blood pressure"
  ) +
  theme_minimal()

p4
```

# Group Summaries

Summaries help check what the plot suggests.

```{r}
age_group_summary <- adult_bp |>
  group_by(age_group) |>
  summarise(
    n = n(),
    mean_systolic = mean(systolic_mean),
    median_systolic = median(systolic_mean),
    iqr_systolic = IQR(systolic_mean),
    .groups = "drop"
  )

age_group_summary
```

# Takeaways

- Grouped distribution plots compare more than group averages.
- Box plots are compact and good for median/spread comparisons.
- Violin plots show more shape, but can be harder for new readers.
- Add points only when they improve the plot rather than clutter it.
