---
title: "Outliers And Summaries"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
```

# Goal

This notebook shows how to use visualization and summaries to find unusual values.

Outliers are not automatically errors. They are values worth investigating.

# Analytic Questions

We will ask:

- What does the distribution of daily energy intake look like in the teaching file?
- Which values are unusually high by a simple IQR rule?
- How do we visualize outliers without letting them dominate the whole analysis?

# Load Data

```{r}
nhanes_path <- "data/real/nhanes_2021_2023_analysis.csv"

nhanes <- read_csv(nhanes_path, show_col_types = FALSE)

nhanes |>
  glimpse()
```

# Prepare Dietary Data

```{r}
diet <- nhanes |>
  filter(!is.na(mean_daily_energy_kcal))

diet |>
  select(respondent_id, age_years, gender, age_group, mean_daily_energy_kcal) |>
  head()
```

The dietary summaries are simple averages over available recall days. They are for exploratory teaching examples, not formal usual-intake estimation.

# Plot The Distribution

```{r}
p1 <- ggplot(
  diet,
  aes(x = mean_daily_energy_kcal)
)

p1
```

```{r}
p2 <- p1 +
  geom_histogram(binwidth = 250)

p2
```

```{r}
p3 <- p2 +
  labs(
    title = "Distribution Of Mean Daily Energy Intake",
    subtitle = "Simple mean across available dietary recall days",
    x = "Mean daily energy intake, kcal",
    y = "Number of participants"
  ) +
  theme_minimal()

p3
```

# Calculate A Simple Outlier Rule

Use the common 1.5 IQR rule as a screening tool.

```{r}
energy_cutoffs <- diet |>
  summarise(
    q1 = quantile(mean_daily_energy_kcal, 0.25),
    q3 = quantile(mean_daily_energy_kcal, 0.75),
    iqr = IQR(mean_daily_energy_kcal),
    lower_cutoff = q1 - 1.5 * iqr,
    upper_cutoff = q3 + 1.5 * iqr
  )

energy_cutoffs
```

Flag unusual values.

```{r}
diet_flagged <- diet |>
  mutate(
    high_energy_outlier = mean_daily_energy_kcal > energy_cutoffs$upper_cutoff
  )

diet_flagged |>
  count(high_energy_outlier)
```

# Visualize The Flag

```{r}
p1 <- ggplot(
  diet_flagged,
  aes(
    x = mean_daily_energy_kcal,
    fill = high_energy_outlier
  )
) +
  geom_histogram(binwidth = 250)

p1
```

```{r}
p2 <- p1 +
  scale_fill_manual(
    values = c("TRUE" = "firebrick", "FALSE" = "gray70"),
    labels = c("FALSE" = "Not flagged", "TRUE" = "Flagged high"),
    name = NULL
  )

p2
```

```{r}
p3 <- p2 +
  labs(
    title = "High Energy Intake Values Flagged By IQR Rule",
    x = "Mean daily energy intake, kcal",
    y = "Number of participants"
  ) +
  theme_minimal()

p3
```

# Inspect Flagged Rows

```{r}
diet_flagged |>
  filter(high_energy_outlier) |>
  select(respondent_id, age_years, gender, age_group, dietary_days, mean_daily_energy_kcal) |>
  arrange(desc(mean_daily_energy_kcal)) |>
  head(10)
```

# Compare With A Box Plot

```{r}
p1 <- ggplot(
  diet,
  aes(x = age_group, y = mean_daily_energy_kcal)
) +
  geom_boxplot()

p1
```

```{r}
p2 <- p1 +
  labs(
    title = "Mean Daily Energy Intake By Age Group",
    x = "Age group",
    y = "Mean daily energy intake, kcal"
  ) +
  theme_minimal()

p2
```

# Takeaways

- Outliers can be errors, true rare values, or signs that the variable needs context.
- A rule like 1.5 IQR is a screening tool, not a final judgment.
- Visualization helps show whether unusual values are isolated or part of a tail.
- Inspect the rows behind surprising values before deciding what to do.
